### 2024-12-16 landscapemodels.Rmd
analysis

---
title: "Landscape models"
output: html_notebook
---

```{r setup, echo = FALSE}
library(tidyverse)
library(performance)
library(psych)

wfs <- readRDS("data/maxwfsresults_20241209.rds")
lsm <- read.csv("data/lsm_reduced.csv") %>% 
  mutate(hu = gsub("hu", "", hu))

wfs_lsm <- wfs %>% left_join(lsm, by = "hu") %>%
  select(c(hu, huc2, name, margin.ha, lu_ta,
           pollinators_base:sed_export_base, 
           delta_pollinators:delta_sed, 
           fields_max, field.ha, margin.ha, 
           lu_shdi:ntr_frac_mn)) %>%
  mutate(prn_ann_ratio = log(prn_agr_pland/ann_agr_pland),
         prn_pagr = 100*(prn_agr_pland/(ann_agr_pland+prn_agr_pland)),
         field_pland = field.ha/lu_ta,
         margin_dens = margin.ha/lu_ta)

scaled.data <- wfs_lsm %>% mutate(across(lu_shdi:field_pland, ~scale(.x)[,1]))

```

## Correlation between independent variables.

I selected a subset of variables that aren't strongly correlated and that represent potential drivers of ecosystem service patterns.

```{r correlation, echo = FALSE}

pairs.panels(wfs_lsm %>% 
               select(ntr_pland, prn_agr_pland, aglu_iji,
                      lu_shdi, aglu_frac_mn, margin_dens), 
             cex.cor = 1, cex.labels=1, cex=0.5)

```

## Pollinator models

### Pollinator delta

```{r pollinator_delta}

pol.dmod0 <- lm(delta_pollinators ~ ntr_pland + margin_dens + prn_agr_pland, data = scaled.data)
pol.dmod1 <- lm(delta_pollinators ~ ntr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
pol.dmod2 <- lm(delta_pollinators ~ ntr_pland + I(ntr_pland^2) + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)

performance::compare_performance(pol.dmod0, pol.dmod1, pol.dmod2)

summary(pol.dmod2)

```

## Seasonal water yield

### Seasonal water yield delta

```{r swy_delta}

swy.dmod0 <- lm(delta_qb ~ ntr_pland + margin_dens + prn_agr_pland, data = scaled.data)
swy.dmod1 <- lm(delta_qb ~ ntr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
swy.dmod2 <- lm(delta_qb ~ ntr_pland + I(ntr_pland^2) + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)

performance::compare_performance(swy.dmod0, swy.dmod1, swy.dmod2)

summary(swy.dmod2)

```

## Nutrient delivery ratio

### Nutrient delivery ratio delta

```{r ndr_delta}

ndr.dmod0 <- lm(delta_n_export ~ ntr_pland + margin_dens + prn_agr_pland, data = scaled.data)
ndr.dmod1 <- lm(delta_n_export ~ ntr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
ndr.dmod2 <- lm(delta_n_export ~ ntr_pland + I(ntr_pland^2) + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)

performance::compare_performance(ndr.dmod0, ndr.dmod1, ndr.dmod2)

summary(ndr.dmod2)

```

## Sediment delivery ratio

### Sediment delivery ratio delta

```{r sdr_delta}

sdr.dmod0 <- lm(delta_sed ~ ntr_pland + margin_dens + prn_agr_pland, data = scaled.data)
sdr.dmod1 <- lm(delta_sed ~ ntr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
sdr.dmod2 <- lm(delta_sed ~ ntr_pland + I(ntr_pland^2) + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)

performance::compare_performance(sdr.dmod0, sdr.dmod1, sdr.dmod2)

summary(sdr.dmod2)

```

### 2024-12-16 R/utils-cor.R

@@ -1,0 +1,41 @@
+# Define a distance function https://davetang.org/muse/2010/11/26/hierarchical-clustering-with-p-values-using-spearman/
+abs.spearman.dist <- function(x, ...) {
+  x <- as.matrix(x)
+  res <- as.dist(1 - abs(cor(x, method = "spearman", use = "everything")))
+  res <- as.dist(res)
+  attr(res, "method") <- "abs. spearman"
+  return(res)
+}
+
+# Make non-significance symbols for heatmaps
+make.sig.mat <- function(pmat, x.90 = TRUE){
+  pval <- pmat$P
+  sig <- ifelse(pval<=0.05, "", "O")
+  sig2 <- ifelse(pval<=0.01, "", "o")
+  sig3 <- ifelse(pval<=0.001, "", ".")
+  sig123 <- ifelse(sig=="", ifelse(sig2=="",sig3, sig2), sig)
+  
+  if(x.90){
+    sig123 <- ifelse(pmat$r>=.9,"X",sig123)
+  }
+  
+  sig123
+}
+
+# heatmap function
+speardist.heatmap <- function(varmat){
+  spearcor <- Hmisc::rcorr(as.matrix(varmat), type = "spearman")
+  
+  spearcor.pval <- make.sig.mat(spearcor) # add significance
+  
+  gplots::heatmap.2(as.matrix(abs.spearman.dist(varmat)), trace = "none",
+                    distfun = as.dist,
+                    hclustfun = function(d) hclust(d, method = "complete"),
+                    col=cm.colors(64), margin=c(5, 10),
+                    offsetCol = 0, offsetRow = 0,
+                    srtCol=45,  adjCol = c(0.9,0),
+                    cellnote = spearcor.pval,
+                    notecol = "black"
+  )
+}
+

### 2024-12-16 get_results.R

@@ -1,0 +1,53 @@
+library(tidyverse)
+library(sf)
+library(terra)
+
+projdir <- "/Users/kevinli/Box Sync/InVEST/Optimization/Data/maxWFS/"
+hu <- st_read(paste0(projdir, "data/hu12watersheds.gpkg"))
+regions <- st_read("/Users/kevinli/Box Sync/InVEST/GIS_data/NEMWregion/focalhuc2.gpkg")
+
+hu.names <- hu$huc12
+
+# previously calculted data
+baseline <- readRDS("data-raw/baseline.rds")
+maxwfs <- readRDS("data-raw/maxwfs.rds")
+
+# combine results 
+baselinedf <- bind_rows(baseline) %>% mutate(hu = hu.names)
+maxwfsdf <- bind_rows(maxwfs) %>% mutate(hu=hu.names)
+
+allresults <- baselinedf %>% left_join(maxwfsdf, by = "hu", suffix = c("_base", "_max")) %>%
+  relocate(hu)
+
+# get the total field area and margin area
+total.fields <- c()
+total.margin <- c()
+for(i in 1:length(hu.names)){
+  hu.name <- hu.names[i]
+  
+  hudir <- paste0(projdir, "hu/hu",hu.name,"/")
+  hu.fields <- rast(paste0(hudir, "data/lu/fields", ".tif"))
+  hu.margins <- rast(paste0(hudir, "data/lu/margins", ".tif"))
+  
+  total.fields[i] <- length(values(hu.fields$id, mat=FALSE, na.rm=TRUE))
+  total.margin[i] <- length(values(hu.margins$id, mat=FALSE, na.rm=TRUE))
+}
+
+# combine all the results
+
+huresults <- allresults %>% 
+  bind_cols(field.ha = total.fields*(9/100), margin.ha=total.margin*(9/100)) %>%
+  left_join(hu %>% st_drop_geometry(), by = c("hu" = "huc12")) %>% 
+  mutate(delta_pollinators = pollinators_max-pollinators_base,  # change per ha
+         delta_n_export = n_export_base - n_export_max,  # change per ha
+         delta_qb = qb_max - qb_base,   # is an average per ha
+         delta_sed = sed_export_base - sed_export_max   # change per ha
+         ) %>%
+  mutate(pc_chng_pollinators = 100*delta_pollinators/pollinators_base,
+         pc_chng_n_export = 100*delta_n_export/n_export_base,
[…]

### 2024-12-16 landmetrics.R

@@ -1,0 +1,106 @@
+library(landscapemetrics)
+library(terra)
+library(tidyverse)
+
+datadir <- "/Users/kevinli/Box Sync/InVEST/Optimization/Data/maxWFS/"
+lcgroup <- read.csv("lc_grouping.csv")
+
+# view landscape metrics
+viewlsm <- list_lsm()
+
+hulist <- list.files(paste0(datadir,"hu"))
+
+# aggregated classes reclass table
+lc.agg <- lcgroup %>% 
+  mutate(landcover_ag = abbreviate(landcover_ag,3),
+         landuse_group = abbreviate(landuse_group,3)) %>%
+  unite(col = "landuse_group", landcover_ag,landuse_group, na.rm = TRUE,sep = "_") %>%
+  select(Value, CDL_name, landuse_group) %>%
+  mutate(groupcode = as.numeric(as.factor(landuse_group)))
+
+reclass.agg <- lc.agg %>% select(Value, groupcode)
+
+# landscape metrics
+lsm.list <- c("lsm_l_ta", "lsm_l_prd", "lsm_l_pr", "lsm_l_shdi")
+lsm.agg.list <- c("lsm_l_iji", "lsm_l_frac_mn", "lsm_l_frac_cv", "lsm_l_para_mn", "lsm_l_para_cv", "lsm_l_contag", "lsm_l_ed", "lsm_l_split", "lsm_l_lsi")
+
+# class metrics
+list.agg.cm <- c("lsm_c_frac_mn", "lsm_c_frac_cv", "lsm_c_para_mn", "lsm_c_para_cv", "lsm_c_area_mn", "lsm_c_area_cv", "lsm_c_pland", "lsm_c_pd", "lsm_c_cohesion", "lsm_c_ed", "lsm_c_split", "lsm_c_lsi")
+
+ls.result <- list()
+ls.result.agg <- list()
+ls.result.class <- list()
+ls.check <- list()
+
+for(i in seq_along(hulist)){
+  
+  hu.i <- terra::rast(paste0(datadir,"hu/",hulist[i],"/data/lu/fields.tif"))
+  mask.i <- terra::vect(paste0(datadir,"hu/",hulist[i],"/data/watershed.gpkg"))
+  hucrop <- terra::mask(hu.i$cdl.clean, mask.i)
+  
+  hucrop.agg <- hucrop %>% classify(rcl = reclass.agg)
+  
[…]

### 2024-12-16 landmetrics_corplot.R

@@ -1,0 +1,133 @@
+library(tidyverse)
+library(corrplot)
+library(gplots)
+library(Hmisc)
+library(pvclust)
+
+lsm <- read.csv("lsm.csv")
+names(lsm)
+
+lsm2 <- lsm %>% select(-dvl_area_mn, -non_area_mn, -wtr_area_mn,
+                       -dvl_area_cv, -non_area_cv, -wtr_area_cv,
+                       -dvl_frac_mn, -non_frac_mn, -wtr_frac_mn,
+                       -dvl_frac_cv, -non_frac_cv, -wtr_frac_cv,
+                       -dvl_para_mn, -non_para_mn, -wtr_para_mn,
+                       -dvl_para_cv, -non_para_cv, -wtr_para_cv,
+                       -dvl_cohesion, -non_cohesion, -wtr_cohesion,
+                       -dvl_ed, -non_ed, -wtr_ed,
+                       -dvl_pd, -non_pd, -wtr_pd,
+                       -dvl_lsi, -non_lsi, -wtr_lsi,
+                       -dvl_pland, -non_pland, -wtr_pland,
+                       -dvl_split, -non_split, -wtr_split)
+
+# Shapiro Wilks test for normality
+lsm.shap <- lapply(lsm2[,2:ncol(lsm2)], \(x) data.frame(statistic=shapiro.test(x)$statistic, p.value = shapiro.test(x)$p.value)) %>%
+  bind_rows(.id = "metric")
+
+lsmcor <- cor(lsm2[,2:ncol(lsm2)], method = "pearson")
+
+# spearman correlation with p values
+lsmcor.spear <- rcorr(as.matrix(lsm2[,2:ncol(lsm2)]), type = "spearman")
+
+# plot(hclust(dist(lsmcor)))
+
+# corrplot(cor(lsm2[,2:ncol(lsm2)]), type = "upper", order = "hclust", method = "ellipse")
+
+# combine cluster and corrplot
+# heatmap.2(lsmcor, trace = "none",
+#           distfun = dist,
+#           hclustfun = hclust,
+#           col=cm.colors(64), margin=c(5, 10),
+#           offsetCol = 0, offsetRow = 0,
+#           srtCol=45,  adjCol = c(0.9,0)
+#           )
+# 
+# heatmap.2(lsmcor.spear$r, trace = "none",
[…]

### 2024-12-16 landscapecor.R

@@ -1,0 +1,33 @@
+library(tidyverse)
+library(psych)
+
+wfs <- readRDS("data/maxwfsresults_20241209.rds")
+# test <- readRDS("maxwfs_fullresults_20240816_v2.rds")
+lsm <- read.csv("data/lsm_reduced.csv") %>% 
+  mutate(hu = gsub("hu", "", hu))
+lsm_all <- read.csv("data/lsm.csv") %>% 
+  mutate(hu = gsub("hu", "", hu))
+
+wfs_lsm <- wfs %>% left_join(lsm, by = "hu") %>%
+  select(c(hu, delta_pollinators:delta_sed, lu_shdi:field_para_mn))
+
+pairs.panels(wfs %>% left_join(lsm, by = "hu") %>%
+               select(c(delta_pollinators:delta_sed, lu_shdi:field_para_mn)),
+             histogram = TRUE, stars = TRUE, 
+             method = "spearman", pch=".", cex.labels = 1.1, cex.cor = 1)
+
+# description
+library(egg)
+lsm_about <- read.csv("data/lsm_description.csv")
+
+wfs_lsm2 <- wfs_lsm %>% pivot_longer(lu_shdi:field_para_mn, names_to = "landscape.metric", values_to = "value")
+
+ggplot(wfs_lsm2, aes(value, delta_pollinators)) + geom_point() + 
+  facet_wrap(~landscape.metric, scales = "free_x") + theme_article()
+ggplot(wfs_lsm2, aes(value, delta_n_export)) + geom_point() + 
+  facet_wrap(~landscape.metric, scales = "free_x") + theme_article()
+ggplot(wfs_lsm2, aes(value, delta_qb)) + geom_point() + 
+  facet_wrap(~landscape.metric, scales = "free_x") + theme_article()
+ggplot(wfs_lsm2, aes(value, delta_sed)) + geom_point() + 
+  facet_wrap(~landscape.metric, scales = "free_x") + theme_article()
+

### 2024-12-16 landscapemodels.R

@@ -1,0 +1,125 @@
+library(tidyverse)
+library(lme4)
+library(glmmTMB)
+library(DHARMa)
+library(performance)
+library(piecewiseSEM)
+library(psych)
+source("R/utils-cor.R")
+
+wfs <- readRDS("data/maxwfsresults_20241209.rds")
+lsm <- read.csv("data/lsm_reduced.csv") %>% 
+  mutate(hu = gsub("hu", "", hu))
+
+wfs_lsm <- wfs %>% left_join(lsm, by = "hu") %>%
+  select(c(hu, huc2, name, margin.ha, lu_ta,
+           pollinators_base:sed_export_base, 
+           delta_pollinators:delta_sed, 
+           fields_max, field.ha, margin.ha, 
+           lu_shdi:ntr_frac_mn)) %>%
+  mutate(prn_ann_ratio = log(prn_agr_pland/ann_agr_pland),
+         prn_pagr = 100*(prn_agr_pland/(ann_agr_pland+prn_agr_pland)),
+         field_pland = field.ha/lu_ta,
+         margin_dens = margin.ha/lu_ta)
+
+scaled.data <- wfs_lsm %>% mutate(across(lu_shdi:field_pland, ~scale(.x)[,1]))
+
+# % land area relationships
+lsm.area <- wfs_lsm %>% 
+  select(ntr_ed,
+         ann_agr_ed,
+         prn_agr_ed,
+         margin_dens,
+         ntr_pland,
+         ann_agr_pland,
+         prn_agr_pland,
+         field_pland)
+
+pairs.panels(lsm.area, cex.cor = 1.2, cex.labels=0.7, cex=0.5)
+
+pairs.panels( wfs_lsm %>% 
+                select(pollinators_base, delta_pollinators, field_frac_mn, field.ha, ntr_pland, field_ed), cex.cor = 1.2, cex.labels=0.7, cex=0.5)
+
+# shape relationships
+lsm.shape <- wfs_lsm %>% 
+  select(ntr_pland, ann_agr_pland, field.ha, prn_agr_pland, 
+         aglu_frac_mn,
+         ntr_frac_mn,
+         ann_agr_frac_mn,
+         prn_agr_frac_mn, 
+         field_frac_mn)
+pairs.panels(lsm.shape, cex.cor = 1, cex.labels=1)
+
+speardist.heatmap(lsm.shape)
+
+# aggregation relationships
+lsm.agg <- wfs_lsm %>% 
[…]

### 2024-12-16 selectmetrics.R

@@ -1,0 +1,167 @@
+library(tidyverse)
+library(landscapemetrics)
+library(fuzzyjoin)
+library(PerformanceAnalytics)
+library(psych)
+library(Hmisc)
+
+# view lHmisc# view landscape metrics
+viewlsm <- list_lsm()
+
+lsm <- read.csv("data/lsm.csv")
+
+lsm2 <- lsm %>% select(-dvl_area_mn, -non_area_mn, -wtr_area_mn,
+                       -dvl_area_cv, -non_area_cv, -wtr_area_cv,
+                       -dvl_frac_mn, -non_frac_mn, -wtr_frac_mn,
+                       -dvl_frac_cv, -non_frac_cv, -wtr_frac_cv,
+                       -dvl_para_mn, -non_para_mn, -wtr_para_mn,
+                       -dvl_para_cv, -non_para_cv, -wtr_para_cv,
+                       -dvl_cohesion, -non_cohesion, -wtr_cohesion,
+                       -dvl_ed, -non_ed, -wtr_ed,
+                       -dvl_pd, -non_pd, -wtr_pd,
+                       -dvl_lsi, -non_lsi, -wtr_lsi,
+                       -dvl_pland, -non_pland, -wtr_pland,
+                       -dvl_split, -non_split, -wtr_split)
+
+names(lsm2)
+
+# backwards engineer metric descriptors
+# not pretty but I did this
+lsm2.class <- data.frame(lsm = names(lsm2)[2:ncol(lsm2)]) %>% 
+  separate(lsm, into = c("left", "right"), extra = "merge", sep = "_", remove = FALSE) %>%
+  left_join(viewlsm, by = c("right" = "metric")) %>%
+  select(lsm, left, type, right) %>% rename("lsm.metric" = "right", "landvar"="left") %>% distinct() %>% 
+  separate(lsm, into = c("left", "right", "right2"), extra = "merge", sep = "_", remove = FALSE) %>%
+  left_join(viewlsm, by = c("right2" = "metric"), suffix=c("",".2")) %>% 
+  unite("landvar.2", c(left,right), remove = FALSE) %>%
+  select(lsm, landvar, landvar.2, type, lsm.metric, left, right2, type.2) %>%
+  rename("lsm.metric2" = "right2") %>% distinct() %>%
+  mutate(metric = ifelse(is.na(type),lsm.metric2,lsm.metric),
+         landvar = ifelse(is.na(type),landvar.2,landvar),
[…]

### 2024-12-18 landscapemodels.Rmd
update analysis

@@
 
-wfs <- readRDS("data/maxwfsresults_20241209.rds")
+wfs <- readRDS("data/maxwfsresults_20241217_fix.rds")
 lsm <- read.csv("data/lsm_reduced.csv") %>% 
@@
            pollinators_base:sed_export_base, 
+           pollinators_max:sed_export_max, 
            delta_pollinators:delta_sed, 
@@
 
-pol.dmod0 <- lm(delta_pollinators ~ ntr_pland + margin_dens + prn_agr_pland, data = scaled.data)
+pol.dmod0 <- lm(delta_pollinators ~ margin_dens, data = scaled.data)
 pol.dmod1 <- lm(delta_pollinators ~ ntr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
@@
 
-swy.dmod0 <- lm(delta_qb ~ ntr_pland + margin_dens + prn_agr_pland, data = scaled.data)
+swy.dmod0 <- lm(delta_qb ~ margin_dens, data = scaled.data)
 swy.dmod1 <- lm(delta_qb ~ ntr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
@@
 
-ndr.dmod0 <- lm(delta_n_export ~ ntr_pland + margin_dens + prn_agr_pland, data = scaled.data)
+ndr.dmod0 <- lm(delta_n_export ~ margin_dens, data = scaled.data)
 ndr.dmod1 <- lm(delta_n_export ~ ntr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
@@
 
-sdr.dmod0 <- lm(delta_sed ~ ntr_pland + margin_dens + prn_agr_pland, data = scaled.data)
+sdr.dmod0 <- lm(delta_sed ~ margin_dens, data = scaled.data)
 sdr.dmod1 <- lm(delta_sed ~ ntr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
@@
 ```
+
+## Plotting
+
+```{r plot_function}
+
+plot_fun <- function(mod, pred.vars, mod.data, color_by = "margin_dens"){
+  predict.list <- lapply(pred.vars,\(x){ggeffects::ggpredict(pol.dmod2, x)})
+  names(predict.list) <- pred.vars
+
+  pred.trends <- dplyr::bind_rows(predict.list, .id = "pred.var") %>%
+    dplyr::rename_with(~as.character(mod$terms[[2]]), predicted)
+
+  pred.data <- mod.data %>%
+    select(all_of(c(as.character(mod$terms[[2]]), pred.vars))) %>%
+    mutate(color_by = get(color_by)) %>%
+    pivot_longer(all_of(pred.vars), names_to = "pred.var", values_to = "x") %>%
+    rename(!!color_by := color_by)
+
+  list(pred.trends, pred.data)
+}
+
+```
+
+### Pollinators
+
+```{r pollinator_plots}
+library(egg)
+
+pol.vars <- c("margin_dens", "ntr_pland", "aglu_frac_mn")
+
+pol.preds <- plot_fun(pol.dmod2, pol.vars, scaled.data)
+
+ggplot(pol.preds[[1]], aes(x, delta_pollinators)) +  
+  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
+  geom_point(data = pol.preds[[2]], 
+             mapping = aes(color = margin_dens), 
+             size = 0.6) + scale_color_continuous(type = "viridis") +
+  facet_wrap(~pred.var, scales = "free_x") +
+  xlab("z-score") +
+  theme_article()
+
+```
+
+### Seasonal water yield
+
+```{r swy_plots}
+
+swy.vars <- c("margin_dens", "ntr_pland", "prn_agr_pland")
+
+swy.preds <- plot_fun(swy.dmod2, swy.vars, scaled.data)
+
+ggplot(swy.preds[[1]], aes(x, delta_qb)) +  
+  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
+  geom_point(data = swy.preds[[2]], 
+             mapping = aes(color = margin_dens), 
+             size = 0.6) + scale_color_continuous(type = "viridis") +
+  facet_wrap(~pred.var, scales = "free_x") +
+  xlab("z-score") +
+  theme_article()
+
+```
+
+### Nutrient delivery ratio
+
+```{r ndr_plots}
+
+ndr.vars <- c("margin_dens", "ntr_pland", "prn_agr_pland", 
+              "aglu_iji", "aglu_frac_mn", "lu_shdi")
+
+ndr.preds <- plot_fun(ndr.dmod2, ndr.vars, scaled.data)
+
+ggplot(ndr.preds[[1]], aes(x, delta_n_export)) +  
+  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
+  geom_point(data = ndr.preds[[2]], 
+             mapping = aes(color = margin_dens), 
+             size = 0.6) + scale_color_continuous(type = "viridis") +
+  facet_wrap(~pred.var, scales = "free_x") +
+  xlab("z-score") +
+  theme_article()
+
+```
+
+### Sediment delivery ratio
+
+```{r sdr_plots}
+
+sdr.vars <- c("margin_dens", "ntr_pland", "prn_agr_pland", "lu_shdi")
+
+sdr.preds <- plot_fun(sdr.dmod2, sdr.vars, scaled.data)
+
+ggplot(sdr.preds[[1]], aes(x, delta_sed)) +  
+  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
+  geom_point(data = sdr.preds[[2]], 
+             mapping = aes(color = margin_dens), 
+             size = 0.6) + scale_color_continuous(type = "viridis") +
+  facet_wrap(~pred.var, scales = "free_x") +
+  xlab("z-score") +
+  theme_article()
+
+```
+
+## Residuals-based models
+
+Since margin density is a major effect, I regress the other predictors against the residuals of univariate models with only margin density
+
+```{r pollinators_resid}
+
+pol.resid <- data.frame(pol_resid = pol.dmod0$residuals, 
+                        scaled.data %>% 
+                          select(ntr_pland, prn_agr_pland, aglu_iji, 
+                                 aglu_frac_mn, lu_shdi))
+
+pol.resid.mod <- lm(pol_resid ~ ntr_pland + I(ntr_pland^2) + prn_agr_pland + aglu_iji + aglu_frac_mn + lu_shdi, data = pol.resid)
+
+summary(pol.resid.mod)
+
+pol.resid.vars <- c("ntr_pland", "prn_agr_pland", "aglu_frac_mn")
+
+pol.resid.preds <- plot_fun(pol.resid.mod, pol.resid.vars, pol.resid, "ntr_pland")
+
+ggplot(pol.resid.preds[[1]], aes(x, pol_resid)) +  
+  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
+  geom_point(data = pol.resid.preds[[2]], 
+             mapping = aes(color = ntr_pland), 
+             size = 0.6) + scale_color_continuous(type = "viridis") +
+  facet_wrap(~pred.var, scales = "free_x") +
+  xlab("z-score") +
+  theme_article()
+
+```

### 2024-12-18 get_results.R

@@ -9,8 +9,8 @@ regions <- st_read("/Users/kevinli/Box Sync/InVEST/GIS_data/NEMWregion/focalhuc2
 hu.names <- hu$huc12
 
 # previously calculted data
-baseline <- readRDS("data-raw/baseline.rds")
-maxwfs <- readRDS("data-raw/maxwfs.rds")
+baseline <- readRDS("data-raw/baseline_20241217_fix.rds")
+maxwfs <- readRDS("data-raw/maxwfs_20241217_fix.rds")
 
 # combine results 
 baselinedf <- bind_rows(baseline) %>% mutate(hu = hu.names)
@@ -50,4 +50,4 @@ huresults <- allresults %>%
   left_join(regions %>% select(huc2,name) %>% st_drop_geometry(), by="huc2", 
             suffix = c("",".huc2"))
 
-saveRDS(huresults, "data/maxwfsresults_20241209.rds")
+# saveRDS(huresults, "data/maxwfsresults_20241217_fix.rds")

### 2024-12-18 landscapemodels.R

@@ -7,7 +7,7 @@ library(piecewiseSEM)
 library(psych)
 source("R/utils-cor.R")
 
-wfs <- readRDS("data/maxwfsresults_20241209.rds")
+wfs <- readRDS("data/maxwfsresults_20241217_fix.rds")
 lsm <- read.csv("data/lsm_reduced.csv") %>% 
   mutate(hu = gsub("hu", "", hu))
 
@@ -26,19 +26,37 @@ scaled.data <- wfs_lsm %>% mutate(across(lu_shdi:field_pland, ~scale(.x)[,1]))
 
 # % land area relationships
 lsm.area <- wfs_lsm %>% 
-  select(ntr_ed,
-         ann_agr_ed,
-         prn_agr_ed,
-         margin_dens,
+  select(
          ntr_pland,
          ann_agr_pland,
          prn_agr_pland,
-         field_pland)
+         prn_pagr,
+         field_pland,
+         margin_dens)
 
-pairs.panels(lsm.area, cex.cor = 1.2, cex.labels=0.7, cex=0.5)
+pairs.panels(lsm.area, cex.cor = 1, cex.labels=1, cex=0.5)
 
 pairs.panels( wfs_lsm %>% 
-                select(pollinators_base, delta_pollinators, field_frac_mn, field.ha, ntr_pland, field_ed), cex.cor = 1.2, cex.labels=0.7, cex=0.5)
+                select(ntr_pland,
+                       ann_agr_pland,
+                       prn_agr_pland,
+                       field_pland,
+                       ntr_ed,
+                       ann_agr_ed,
+                       prn_agr_ed,
+                       margin_dens), 
+              cex.cor = 1.2, cex.labels=1, cex=0.5)
+
+# field relationships
+pairs.panels(wfs_lsm %>% 
+               select(ntr_pland, ann_agr_pland,
+                      field_pland,
+                      field_para_mn,
+                      field_frac_mn,
+                      field_area_mn,
+                      field_ed, field_pd,
+                      margin_dens), 
+             cex.cor = 1, cex.labels=1, cex=0.5)
 
 # shape relationships
 lsm.shape <- wfs_lsm %>% 
@@ -59,14 +77,6 @@ lsm.agg <- wfs_lsm %>%
 pairs.panels(lsm.agg, cex.cor = 1, cex.labels=1)
 
 speardist.heatmap(lsm.agg)
-
-# field relationships
-pairs.panels(wfs_lsm %>% 
[…]

### 2024-12-27 landscapemodels_gam.Rmd
more analysis attempts

---
title: "Landscape models with Generalized Additive Models"
---

```{r setup, echo = FALSE}
library(tidyverse)
library(performance)
library(psych)
library(mgcv)
library(ggeffects)
library(piecewiseSEM)
source("R/utils-plot.r")

wfs <- readRDS("data/maxwfsresults_20241217_fix.rds")
lsm <- read.csv("data/lsm_reduced.csv") %>% 
  mutate(hu = gsub("hu", "", hu))

wfs_lsm <- wfs %>% left_join(lsm, by = "hu") %>%
  select(c(hu, huc2, name, margin.ha, lu_ta,
           pollinators_base:sed_export_base, 
           pollinators_max:sed_export_max, 
           delta_pollinators:delta_sed, 
           fields_max, field.ha, margin.ha, 
           lu_shdi:ntr_frac_mn)) %>%
  mutate(prn_ann_ratio = log(prn_agr_pland/ann_agr_pland),
         prn_pagr = 100*(prn_agr_pland/(ann_agr_pland+prn_agr_pland)),
         field_pland = field.ha/lu_ta,
         margin_dens = margin.ha/lu_ta)

scaled.data <- wfs_lsm %>% mutate(across(lu_shdi:field_pland, ~scale(.x)[,1]))

```

## Correlation between independent variables.

I selected a subset of variables that aren't strongly correlated and that represent potential drivers of ecosystem service patterns.

```{r correlation, echo = FALSE}

pairs.panels(wfs_lsm %>% 
               select(ann_agr_pland, prn_agr_pland, aglu_iji,
                      lu_shdi, aglu_frac_mn, margin_dens), 
             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")

```

## Pollinator models

### Pollinator delta

```{r pollinator_delta}

pol.lm0 <- gam(delta_pollinators ~ margin_dens, data = wfs_lsm)
pol.gam0 <- gam(delta_pollinators ~ s(margin_dens), data = wfs_lsm)

anova(pol.lm0, pol.gam0)

pol.gam1 <- gam(delta_pollinators ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
                data = wfs_lsm)

performance::compare_performance(pol.lm0, pol.gam0, pol.gam1)

summary(pol.gam1)

concurvity(pol.gam1)

```

```{r plot_pol}

polplot.df <- plot_fun(pol.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "margin_dens"),
                       wfs_lsm)

ggplot(polplot.df[[1]], aes(x, delta_pollinators)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = polplot.df[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

```{r plot_pol0}

polplot0.df <- plot_fun(pol.gam0, 
                       "margin_dens",
                       wfs_lsm)

ggplot(polplot0.df[[1]], aes(x, delta_pollinators)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = polplot0.df[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### Residual model from gam0

Trying to account for the effect of margins first

```{r pol_margin_resid}

wfs_lsm_resids <- wfs_lsm %>% cbind(pol.resid = residuals(pol.gam0))

polresid.gam1 <- gam(pol.resid ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
                data = wfs_lsm_resids)

summary(polresid.gam1)

concurvity(polresid.gam1)

```

```{r plot_pol_resid}

residpolplot.df <- plot_fun(polresid.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn"),
                       wfs_lsm_resids,
                       "ann_agr_pland")

ggplot(residpolplot.df[[1]], aes(x, pol.resid)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = residpolplot.df[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

## Seasonal water yield

### Seasonal water yield delta

```{r swy_delta}

swy.lm0 <- gam(delta_qb ~ margin_dens, data = wfs_lsm)
swy.gam0 <- gam(delta_qb ~ s(margin_dens), data = wfs_lsm)

anova(swy.lm0, swy.gam0)

swy.gam1 <- gam(delta_qb ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
                data = wfs_lsm)

performance::compare_performance(swy.lm0, swy.gam0, swy.gam1)

summary(swy.gam1)

concurvity(swy.gam1)

```

```{r plot_swy}

swyplot.df <- plot_fun(swy.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "margin_dens"),
                       wfs_lsm)

ggplot(swyplot.df[[1]], aes(x, delta_qb)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = swyplot.df[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

```{r plot_swy0}

swyplot0.df <- plot_fun(swy.gam0, 
                       "margin_dens",
                       wfs_lsm)

ggplot(swyplot0.df[[1]], aes(x, delta_qb)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = swyplot0.df[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### Residual model from gam0

Trying to account for the effect of margins first

```{r swy_margin_resid}

wfs_lsm_resids <- wfs_lsm_resids %>% cbind(swy.resid = residuals(swy.gam0))

swyresid.gam1 <- gam(swy.resid ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
                data = wfs_lsm_resids)

summary(swyresid.gam1)

concurvity(swyresid.gam1)

```

```{r plot_swy_resid}

residswyplot.df <- plot_fun(swyresid.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn"),
                       wfs_lsm_resids,
                       "ann_agr_pland")

ggplot(residswyplot.df[[1]], aes(x, swy.resid)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = residswyplot.df[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

## Nutrient delivery ratio

### Nutrient delivery ratio delta

```{r ndr_delta}

ndr.lm0 <- gam(delta_n_export ~ margin_dens, data = wfs_lsm)
ndr.gam0 <- gam(delta_n_export ~ s(margin_dens), data = wfs_lsm)

anova(ndr.lm0, ndr.gam0)

ndr.gam1 <- gam(delta_n_export ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
                data = wfs_lsm)

performance::compare_performance(ndr.lm0, ndr.gam0, ndr.gam1)

summary(ndr.gam1)

concurvity(ndr.gam1)

```

```{r plot_ndr}

ndrplot.df <- plot_fun(ndr.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "margin_dens", "lu_shdi"),
                       wfs_lsm)

ggplot(ndrplot.df[[1]], aes(x, delta_n_export)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = ndrplot.df[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

```{r plot_ndr0}

ndrplot0.df <- plot_fun(ndr.gam0, 
                       "margin_dens",
                       wfs_lsm)

ggplot(ndrplot0.df[[1]], aes(x, delta_n_export)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = ndrplot0.df[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### Residual model from gam0

Trying to account for the effect of margins first

```{r ndr_margin_resid}

wfs_lsm_resids <- wfs_lsm_resids %>% cbind(ndr.resid = residuals(ndr.gam0))

ndrresid.gam1 <- gam(ndr.resid ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
                data = wfs_lsm_resids)

summary(ndrresid.gam1)

concurvity(ndrresid.gam1)

```

```{r plot_ndr_resid}

residndrplot.df <- plot_fun(ndrresid.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
                       wfs_lsm_resids,
                       "ann_agr_pland")

ggplot(residndrplot.df[[1]], aes(x, ndr.resid)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = residndrplot.df[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

## Sediment delivery ratio

### Sediment delivery ratio delta

```{r sdr_delta}

sdr.lm0 <- gam(delta_sed ~ margin_dens, data = wfs_lsm)
sdr.gam0 <- gam(delta_sed ~ s(margin_dens), data = wfs_lsm)

anova(sdr.lm0, sdr.gam0)

sdr.gam1 <- gam(delta_sed ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
                data = wfs_lsm)

performance::compare_performance(sdr.lm0, sdr.gam0, sdr.gam1)

summary(sdr.gam1)

concurvity(sdr.gam1)

```

```{r plot_sdr}

sdrplot.df <- plot_fun(sdr.gam1, 
                       c("ann_agr_pland", "prn_agr_pland", "margin_dens"),
                       wfs_lsm)

ggplot(sdrplot.df[[1]], aes(x, delta_sed)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = sdrplot.df[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

```{r plot_sdr0}

sdrplot0.df <- plot_fun(sdr.gam0, 
                       "margin_dens",
                       wfs_lsm)

ggplot(sdrplot0.df[[1]], aes(x, delta_sed)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = sdrplot0.df[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### Residual model from gam0

Trying to account for the effect of margins first

```{r sdr_margin_resid}

wfs_lsm_resids <- wfs_lsm_resids %>% cbind(sdr.resid = residuals(sdr.gam0))

sdrresid.gam1 <- gam(sdr.resid ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
                data = wfs_lsm_resids)

summary(sdrresid.gam1)

concurvity(sdrresid.gam1)

```

```{r plot_sdr_resid}

residsdrplot.df <- plot_fun(sdrresid.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn"),
                       wfs_lsm_resids,
                       "ann_agr_pland")

ggplot(residsdrplot.df[[1]], aes(x, sdr.resid)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = residsdrplot.df[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### 2024-12-27 landscapemodels_logtrans.Rmd
more analysis attempts

--- 2024-12-27 landscapemodels_gam.Rmd
@@
 ---
-title: "Landscape models with Generalized Additive Models"
+title: "Landscape models using log transform"
 ---
@@
          field_pland = field.ha/lu_ta,
-         margin_dens = margin.ha/lu_ta)
-
-scaled.data <- wfs_lsm %>% mutate(across(lu_shdi:field_pland, ~scale(.x)[,1]))
-
-```
-
-## Correlation between independent variables.
-
-I selected a subset of variables that aren't strongly correlated and that represent potential drivers of ecosystem service patterns.
-
-```{r correlation, echo = FALSE}
-
-pairs.panels(wfs_lsm %>% 
-               select(ann_agr_pland, prn_agr_pland, aglu_iji,
-                      lu_shdi, aglu_frac_mn, margin_dens), 
-             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")
+         margin_dens = margin.ha/lu_ta,
+         log.delta_pollinators = log(delta_pollinators),
+         log.delta_qb = log(ifelse(delta_qb>0,delta_qb,0.0001)))
 
@@
 
-pol.lm0 <- gam(delta_pollinators ~ margin_dens, data = wfs_lsm)
-pol.gam0 <- gam(delta_pollinators ~ s(margin_dens), data = wfs_lsm)
+pol.lm0 <- gam(log.delta_pollinators ~ margin_dens, data = wfs_lsm)
+pol.gam0 <- gam(log.delta_pollinators ~ s(margin_dens), data = wfs_lsm)
 
@@
 
-pol.gam1 <- gam(delta_pollinators ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
+pol.gam1 <- gam(log.delta_pollinators ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
                 data = wfs_lsm)
@@
 
-ggplot(polplot.df[[1]], aes(x, delta_pollinators)) +  
+ggplot(polplot.df[[1]], aes(x, log.delta_pollinators)) +  
   geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
@@
 
-ggplot(polplot0.df[[1]], aes(x, delta_pollinators)) +  
+ggplot(polplot0.df[[1]], aes(x, log.delta_pollinators)) +  
   geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
@@
 
-swy.lm0 <- gam(delta_qb ~ margin_dens, data = wfs_lsm)
-swy.gam0 <- gam(delta_qb ~ s(margin_dens), data = wfs_lsm)
+swy.lm0 <- gam(log.delta_qb ~ margin_dens, data = wfs_lsm)
+swy.gam0 <- gam(log.delta_qb ~ s(margin_dens), data = wfs_lsm)
 
@@
 
-swy.gam1 <- gam(delta_qb ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
+swy.gam1 <- gam(log.delta_qb ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
                 data = wfs_lsm)
@@
 
-ggplot(swyplot.df[[1]], aes(x, delta_qb)) +  
+ggplot(swyplot.df[[1]], aes(x, log.delta_qb)) +  
   geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
@@
 ```
-
-```{r plot_swy0}
-
-swyplot0.df <- plot_fun(swy.gam0, 
-                       "margin_dens",
-                       wfs_lsm)
-
-ggplot(swyplot0.df[[1]], aes(x, delta_qb)) +  
-  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
-  geom_point(data = swyplot0.df[[2]], 
-             mapping = aes(color = margin_dens), 
-             size = 0.6) + scale_color_continuous(type = "viridis") +
-  facet_wrap(~pred.var, scales = "free_x") +
-  egg::theme_article()
-
-```
-
-### Residual model from gam0
-
-Trying to account for the effect of margins first
-
-```{r swy_margin_resid}
-
-wfs_lsm_resids <- wfs_lsm_resids %>% cbind(swy.resid = residuals(swy.gam0))
-
-swyresid.gam1 <- gam(swy.resid ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
-                data = wfs_lsm_resids)
-
-summary(swyresid.gam1)
-
-concurvity(swyresid.gam1)
-
-```
-
-```{r plot_swy_resid}
-
-residswyplot.df <- plot_fun(swyresid.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn"),
-                       wfs_lsm_resids,
-                       "ann_agr_pland")
-
-ggplot(residswyplot.df[[1]], aes(x, swy.resid)) +  
-  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
-  geom_point(data = residswyplot.df[[2]], 
-             mapping = aes(color = ann_agr_pland), 
-             size = 0.6) + scale_color_continuous(type = "viridis") +
-  facet_wrap(~pred.var, scales = "free_x") +
-  egg::theme_article()
-
-```
-
-## Nutrient delivery ratio
-
-### Nutrient delivery ratio delta
-
-```{r ndr_delta}
-
-ndr.lm0 <- gam(delta_n_export ~ margin_dens, data = wfs_lsm)
-ndr.gam0 <- gam(delta_n_export ~ s(margin_dens), data = wfs_lsm)
-
-anova(ndr.lm0, ndr.gam0)
-
-ndr.gam1 <- gam(delta_n_export ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
-                data = wfs_lsm)
-
-performance::compare_performance(ndr.lm0, ndr.gam0, ndr.gam1)
-
-summary(ndr.gam1)
-
-concurvity(ndr.gam1)
-
-```
-
-```{r plot_ndr}
-
-ndrplot.df <- plot_fun(ndr.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "margin_dens", "lu_shdi"),
-                       wfs_lsm)
-
-ggplot(ndrplot.df[[1]], aes(x, delta_n_export)) +  
-  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
-  geom_point(data = ndrplot.df[[2]], 
-             mapping = aes(color = margin_dens), 
-             size = 0.6) + scale_color_continuous(type = "viridis") +
-  facet_wrap(~pred.var, scales = "free_x") +
-  egg::theme_article()
-
-```
-
-```{r plot_ndr0}
-
-ndrplot0.df <- plot_fun(ndr.gam0, 
-                       "margin_dens",
-                       wfs_lsm)
-
-ggplot(ndrplot0.df[[1]], aes(x, delta_n_export)) +  
-  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
-  geom_point(data = ndrplot0.df[[2]], 
-             mapping = aes(color = margin_dens), 
-             size = 0.6) + scale_color_continuous(type = "viridis") +
-  facet_wrap(~pred.var, scales = "free_x") +
-  egg::theme_article()
-
-```
-
-### Residual model from gam0
-
-Trying to account for the effect of margins first
-
-```{r ndr_margin_resid}
-
-wfs_lsm_resids <- wfs_lsm_resids %>% cbind(ndr.resid = residuals(ndr.gam0))
-
-ndrresid.gam1 <- gam(ndr.resid ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
-                data = wfs_lsm_resids)
-
-summary(ndrresid.gam1)
-
-concurvity(ndrresid.gam1)
-
-```
-
-```{r plot_ndr_resid}
-
-residndrplot.df <- plot_fun(ndrresid.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
-                       wfs_lsm_resids,
-                       "ann_agr_pland")
-
-ggplot(residndrplot.df[[1]], aes(x, ndr.resid)) +  
-  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
-  geom_point(data = residndrplot.df[[2]], 
-             mapping = aes(color = ann_agr_pland), 
-             size = 0.6) + scale_color_continuous(type = "viridis") +
-  facet_wrap(~pred.var, scales = "free_x") +
-  egg::theme_article()
-
-```
-
-## Sediment delivery ratio
-
-### Sediment delivery ratio delta
-
-```{r sdr_delta}
-
-sdr.lm0 <- gam(delta_sed ~ margin_dens, data = wfs_lsm)
-sdr.gam0 <- gam(delta_sed ~ s(margin_dens), data = wfs_lsm)
-
-anova(sdr.lm0, sdr.gam0)
-
-sdr.gam1 <- gam(delta_sed ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(margin_dens) + s(lu_shdi), 
-                data = wfs_lsm)
-
-performance::compare_performance(sdr.lm0, sdr.gam0, sdr.gam1)
-
-summary(sdr.gam1)
-
-concurvity(sdr.gam1)
-
-```
-
-```{r plot_sdr}
-
-sdrplot.df <- plot_fun(sdr.gam1, 
-                       c("ann_agr_pland", "prn_agr_pland", "margin_dens"),
-                       wfs_lsm)
-
-ggplot(sdrplot.df[[1]], aes(x, delta_sed)) +  
-  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
-  geom_point(data = sdrplot.df[[2]], 
-             mapping = aes(color = margin_dens), 
-             size = 0.6) + scale_color_continuous(type = "viridis") +
-  facet_wrap(~pred.var, scales = "free_x") +
-  egg::theme_article()
-
-```
-
-```{r plot_sdr0}
-
-sdrplot0.df <- plot_fun(sdr.gam0, 
-                       "margin_dens",
-                       wfs_lsm)
-
-ggplot(sdrplot0.df[[1]], aes(x, delta_sed)) +  
-  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
-  geom_point(data = sdrplot0.df[[2]], 
-             mapping = aes(color = margin_dens), 
-             size = 0.6) + scale_color_continuous(type = "viridis") +
-  facet_wrap(~pred.var, scales = "free_x") +
-  egg::theme_article()
-
-```
-
-### Residual model from gam0
-
-Trying to account for the effect of margins first
-
-```{r sdr_margin_resid}
-
-wfs_lsm_resids <- wfs_lsm_resids %>% cbind(sdr.resid = residuals(sdr.gam0))
-
-sdrresid.gam1 <- gam(sdr.resid ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
-                data = wfs_lsm_resids)
-
-summary(sdrresid.gam1)
-
-concurvity(sdrresid.gam1)
-
-```
-
-```{r plot_sdr_resid}
-
-residsdrplot.df <- plot_fun(sdrresid.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn"),
-                       wfs_lsm_resids,
-                       "ann_agr_pland")
-
-ggplot(residsdrplot.df[[1]], aes(x, sdr.resid)) +  
-  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
-  geom_point(data = residsdrplot.df[[2]], 
-             mapping = aes(color = ann_agr_pland), 
-             size = 0.6) + scale_color_continuous(type = "viridis") +
-  facet_wrap(~pred.var, scales = "free_x") +
-  egg::theme_article()
-
-```

### 2024-12-27 landscapemodels_margin.Rmd
more analysis attempts

---
title: "R Notebook"
---

```{r setup, echo = FALSE}
library(tidyverse)
library(performance)
library(psych)
library(mgcv)
library(ggeffects)
library(piecewiseSEM)
source("R/utils-plot.r")

wfs <- readRDS("data/maxwfsresults_20241217_fix.rds")
lsm <- read.csv("data/lsm_reduced.csv") %>% 
  mutate(hu = gsub("hu", "", hu))

wfs_lsm <- wfs %>% left_join(lsm, by = "hu") %>%
  select(c(hu, huc2, name, margin.ha, lu_ta,
           pollinators_base:sed_export_base, 
           pollinators_max:sed_export_max, 
           delta_pollinators:delta_sed, 
           fields_max, field.ha, margin.ha, 
           lu_shdi:ntr_frac_mn)) %>%
  mutate(prn_ann_ratio = log(prn_agr_pland/ann_agr_pland),
         prn_pagr = 100*(prn_agr_pland/(ann_agr_pland+prn_agr_pland)),
         field_pland = field.ha/lu_ta,
         margin_dens = margin.ha/lu_ta,
         delta_pollinators.margin = delta_pollinators/margin.ha,
         delta_qb.margin = delta_qb/margin.ha,
         delta_n_export.margin = delta_n_export/margin.ha,
         delta_sed.margin = delta_sed/margin.ha)

scaled.data <- wfs_lsm %>% mutate(across(lu_shdi:field_pland, ~scale(.x)[,1]))

```

## Correlation between independent variables.

I selected a subset of variables that aren't strongly correlated and that represent potential drivers of ecosystem service patterns.

```{r correlation, echo = FALSE}

pairs.panels(wfs_lsm %>% 
               select(ann_agr_pland, prn_agr_pland, aglu_iji,
                      lu_shdi, aglu_frac_mn, margin_dens), 
             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")

```

## Pollinator models

### Pollinator delta

```{r pollinator_delta}

pol.lm0 <- gam(delta_pollinators.margin ~ margin_dens, data = wfs_lsm)

summary(pol.lm0)

pol.gam1 <- gam(delta_pollinators.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
                data = wfs_lsm)

performance::compare_performance(pol.lm0,pol.gam1)

summary(pol.gam1)

concurvity(pol.gam1)

```

```{r plot_pol}

polplot.df <- plot_fun(pol.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
                       wfs_lsm, "ann_agr_pland")

ggplot(polplot.df[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = polplot.df[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### Seasonal water yield delta

```{r swy_delta}

swy.lm0 <- gam(delta_qb.margin ~ margin_dens, data = wfs_lsm)

summary(swy.lm0)

swy.gam1 <- gam(delta_qb.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
                data = wfs_lsm)

performance::compare_performance(swy.lm0,swy.gam1)

summary(swy.gam1)

concurvity(swy.gam1)

```

```{r plot_swy}

swyplot.df <- plot_fun(swy.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
                       wfs_lsm, "ann_agr_pland")

ggplot(swyplot.df[[1]], aes(x, delta_qb.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = swyplot.df[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### Nutrient delivery delta

```{r ndr_delta}

ndr.lm0 <- gam(delta_n_export.margin ~ margin_dens, data = wfs_lsm)

summary(ndr.lm0)

ndr.gam1 <- gam(delta_n_export.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
                data = wfs_lsm)

performance::compare_performance(ndr.lm0,ndr.gam1)

summary(ndr.gam1)

concurvity(ndr.gam1)

```

```{r plot_ndr}

ndrplot.df <- plot_fun(ndr.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
                       wfs_lsm, "ann_agr_pland")

ggplot(ndrplot.df[[1]], aes(x, delta_n_export.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = ndrplot.df[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### Sediment delivery ratio delta

```{r sdr_delta}

sdr.lm0 <- gam(delta_sed.margin ~ margin_dens, data = wfs_lsm)

summary(sdr.lm0)

sdr.gam1 <- gam(delta_sed.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
                data = wfs_lsm)

performance::compare_performance(sdr.lm0,sdr.gam1)

summary(sdr.gam1)

concurvity(sdr.gam1)

```

```{r plot_sdr}

sdrplot.df <- plot_fun(sdr.gam1, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
                       wfs_lsm, "ann_agr_pland")

ggplot(sdrplot.df[[1]], aes(x, delta_sed.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = sdrplot.df[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### 2024-12-27 docs/R/utils-plot.R
site

@@ -1,0 +1,16 @@
+plot_fun <- function(mod, pred.vars, mod.data, color_by = "margin_dens"){
+  predict.list <- lapply(pred.vars,\(x){ggeffects::ggpredict(mod, x)})
+  names(predict.list) <- pred.vars
+  
+  pred.trends <- dplyr::bind_rows(predict.list, .id = "pred.var") %>%
+    dplyr::rename_with(~as.character(mod$terms[[2]])[length(as.character(mod$terms[[2]]))],
+                       predicted)
+  
+  pred.data <- mod.data %>%
+    select(all_of(c(as.character(mod$terms[[2]])[length(as.character(mod$terms[[2]]))], pred.vars))) %>%
+    mutate(color_by = get(color_by)) %>%
+    pivot_longer(all_of(pred.vars), names_to = "pred.var", values_to = "x") %>%
+    rename(!!color_by := color_by)
+  
+  list(pred.trends, pred.data)
+}

### 2025-01-08 landscapemodels_gam.Rmd
remake site

@@
 polplot.df <- plot_fun(pol.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "margin_dens"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "margin_dens", "lu_shdi"),
                        wfs_lsm)
@@
 residpolplot.df <- plot_fun(polresid.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
                        wfs_lsm_resids,
@@
 swyplot.df <- plot_fun(swy.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "margin_dens"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "margin_dens", "lu_shdi"),
                        wfs_lsm)
@@
 residswyplot.df <- plot_fun(swyresid.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
                        wfs_lsm_resids,
@@
 residndrplot.df <- plot_fun(ndrresid.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn","lu_shdi"),
                        wfs_lsm_resids,
@@
 sdrplot.df <- plot_fun(sdr.gam1, 
-                       c("ann_agr_pland", "prn_agr_pland", "margin_dens"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "margin_dens", "lu_shdi"),
                        wfs_lsm)
@@
 residsdrplot.df <- plot_fun(sdrresid.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
                        wfs_lsm_resids,

### 2025-01-08 landscapemodels_margin.Rmd
remake site

@@
          margin_dens = margin.ha/lu_ta,
-         delta_pollinators.margin = delta_pollinators/margin.ha,
-         delta_qb.margin = delta_qb/margin.ha,
-         delta_n_export.margin = delta_n_export/margin.ha,
-         delta_sed.margin = delta_sed/margin.ha)
+         delta_pollinators.margin = delta_pollinators/margin_dens,
+         delta_qb.margin = delta_qb/margin_dens,
+         delta_n_export.margin = delta_n_export/margin_dens,
+         delta_sed.margin = delta_sed/margin_dens)
 
@@
 
-## Correlation between independent variables.
+## Normalize ecosystem service results by margin
 
-I selected a subset of variables that aren't strongly correlated and that represent potential drivers of ecosystem service patterns.
-
-```{r correlation, echo = FALSE}
-
-pairs.panels(wfs_lsm %>% 
-               select(ann_agr_pland, prn_agr_pland, aglu_iji,
-                      lu_shdi, aglu_frac_mn, margin_dens), 
-             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")
-
-```
+Change in ecosystem service is normalized by hectares of margin added so that the dependent variable is unit ES per area margin instead of per area watershed
 
@@
 
-pol.gam1 <- gam(delta_pollinators.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
+pol.gam1 <- gam(delta_pollinators.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi) + s(lu_ta), 
                 data = wfs_lsm)
@@
 polplot.df <- plot_fun(pol.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi", "lu_ta"),
                        wfs_lsm, "ann_agr_pland")
@@
 
-swy.gam1 <- gam(delta_qb.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
+swy.gam1 <- gam(delta_qb.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi) + s(lu_ta), 
                 data = wfs_lsm)
@@
 swyplot.df <- plot_fun(swy.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi", "lu_ta"),
                        wfs_lsm, "ann_agr_pland")
@@
 
-ndr.gam1 <- gam(delta_n_export.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
+ndr.gam1 <- gam(delta_n_export.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi) + s(lu_ta), 
                 data = wfs_lsm)
@@
 ndrplot.df <- plot_fun(ndr.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi", "lu_ta"),
                        wfs_lsm, "ann_agr_pland")
@@
 
-sdr.gam1 <- gam(delta_sed.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi), 
+sdr.gam1 <- gam(delta_sed.margin ~ s(ann_agr_pland) + s(aglu_iji) + s(prn_agr_pland) + s(aglu_frac_mn) + s(lu_shdi) + s(lu_ta), 
                 data = wfs_lsm)
@@
 sdrplot.df <- plot_fun(sdr.gam1, 
-                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi"),
+                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi", "lu_ta"),
                        wfs_lsm, "ann_agr_pland")

### 2025-01-08 landscapemodels_margin.Rmd
update site with pairs plot for margin norm page

@@
 
-Change in ecosystem service is normalized by hectares of margin added so that the dependent variable is unit ES per area margin instead of per area watershed
+Change in ecosystem service is normalized by hectares of margin added so that the dependent variable is unit ES per area margin instead of per area watershed.
+I added total watershed area into the full models just to see if there is any association.
+
+```{r correlation, echo = FALSE}
+
+pairs.panels(wfs_lsm %>% 
+               select(ann_agr_pland, prn_agr_pland, aglu_iji,
+                      lu_shdi, aglu_frac_mn, lu_ta), 
+             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")
+
+```
 

### 2025-01-15 landscapemodels_groups.Rmd
update the site with newest analysis

---
title: "Landscape analysis grouped by natural habitat level"
---

```{r setup, echo = FALSE, message=FALSE}
library(tidyverse)
library(performance)
library(psych)
library(mgcv)
library(ggeffects)
library(mgcViz)
source("R/utils-plot.r")

wfs <- readRDS("data/maxwfsresults_20241217_fix.rds")
lsm <- read.csv("data/lsm_reduced.csv") %>% 
  mutate(hu = gsub("hu", "", hu))

wfs_lsm <- wfs %>% left_join(lsm, by = "hu") %>%
  select(c(hu, huc2, name, margin.ha, lu_ta,
           pollinators_base:sed_export_base, 
           pollinators_max:sed_export_max, 
           delta_pollinators:delta_sed, 
           fields_max, field.ha, margin.ha, 
           lu_shdi:ntr_frac_mn)) %>%
  mutate(prn_ann_ratio = log(prn_agr_pland/ann_agr_pland),
         prn_pagr = 100*(prn_agr_pland/(ann_agr_pland+prn_agr_pland)),
         prn_ann_lograt = log(prn_agr_pland/ann_agr_pland),
         field_pland = field.ha/lu_ta,
         margin_dens = margin.ha/lu_ta,
         delta_pollinators.margin = delta_pollinators/margin_dens,
         delta_qb.margin = delta_qb/margin_dens,
         delta_n_export.margin = delta_n_export/margin_dens,
         delta_sed.margin = delta_sed/margin_dens) %>%
  mutate(habcat = cut(ntr_pland, breaks = c(-Inf,20,50,Inf),
                      labels = c("simple","moderate","natural")))

scaled.data <- wfs_lsm %>% mutate(across(lu_shdi:field_pland, ~scale(.x)[,1]))

wfs_sim <- wfs_lsm %>% filter(habcat == "simple")
wfs_mod <- wfs_lsm %>% filter(habcat == "moderate")
wfs_nat <- wfs_lsm %>% filter(habcat == "natural")

```

## Pollinator models

### Simple landscapes (<20% natural habitat)

#### Correlation between independent variables in simple landscapes.

```{r correlation_sim, echo = FALSE}

pairs.panels(wfs_sim %>% 
               select(ann_agr_pland, prn_agr_pland, aglu_iji,
                      lu_shdi, aglu_frac_mn, ntr_pland, prn_ann_lograt), 
             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")

```

#### Pollinator delta for simple landscapes

```{r pollinator_delta_simple}

dpol.sim.gam <- gam(delta_pollinators.margin ~
                  s(ann_agr_pland) +
                  s(aglu_iji) +
                  s(prn_agr_pland) +
                  s(aglu_frac_mn) +
                  s(lu_shdi) + 
                  s(lu_ta), 
                data = wfs_sim)

summary(dpol.sim.gam)

concurvity(dpol.sim.gam, full = FALSE)

# gam.check(dpol.sim.gam)

```

Partial effects plots and partial residuals. If variables are highly correlated, variance in the partial residual plot can be much less than the actual variance: https://en.wikipedia.org/wiki/Partial_residual_plot

```{r part_plot_sim, echo=FALSE}

dpol.sim.viz <- getViz(dpol.sim.gam)

dpol.sim.vizplot <- plot(dpol.sim.viz, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.sim.vizplot, pages = 1)

```

Predicted effects against actual points. Covariates of each plot(not on x axis) held at simple landscape averages.

```{r plot_pol_sim, echo=FALSE}

dpol.sim.pred <- plot_fun(dpol.sim.gam, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland",
                         "aglu_frac_mn", "lu_shdi", "lu_ta"),
                       wfs_sim, "ann_agr_pland")

ggplot(dpol.sim.pred[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = dpol.sim.pred[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

#### Without perennial agriculture or Shannon diversity

These two variables increased concurvity when included with annual agriculture

```{r pollinator_delta_simple2}

dpol.sim.gam2 <- gam(delta_pollinators.margin ~
                  s(ann_agr_pland) +
                  s(aglu_iji) +
                  s(aglu_frac_mn) +
                  # s(lu_shdi) + 
                  s(lu_ta), 
                data = wfs_sim)

summary(dpol.sim.gam2)

concurvity(dpol.sim.gam2)

```

Partial effects plots and partial residuals

```{r part_plot_sim2, echo=FALSE}

dpol.sim.viz2 <- getViz(dpol.sim.gam2)

dpol.sim.vizplot2 <- plot(dpol.sim.viz2, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.sim.vizplot2, pages = 1)

```

Predicted effects against actual points. Covariates of each plot(not on x axis) held at simple landscape averages.

```{r plot_pol_sim2, echo=FALSE}

dpol.sim.pred2 <- plot_fun(dpol.sim.gam2, 
                       c("ann_agr_pland", "aglu_iji",
                         "aglu_frac_mn", "lu_ta"),
                       wfs_sim, "ann_agr_pland")

ggplot(dpol.sim.pred2[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = dpol.sim.pred2[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

#### A composite variable of log ratio of annual to perennial ag

The two types of ag become relative to each other so the connection to the original natural habitat gradient is lost. Therefore I added `ntr_pland` into the model.

```{r pollinator_delta_simple3}

dpol.sim.gam3 <- gam(delta_pollinators.margin ~
                  s(prn_ann_lograt) +
                  s(ntr_pland) +
                  s(aglu_iji) +
                  s(aglu_frac_mn) +
                  # s(lu_shdi) +
                  s(lu_ta), 
                data = wfs_sim)

summary(dpol.sim.gam3)

concurvity(dpol.sim.gam3)

```

Partial effects plots and partial residuals

```{r part_plot_sim3, echo=FALSE}

dpol.sim.viz3 <- getViz(dpol.sim.gam3)

dpol.sim.vizplot3 <- plot(dpol.sim.viz3, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.sim.vizplot3, pages = 1)

```

Predicted effects against actual points. Covariates of each plot(not on x axis) held at simple landscape averages.

```{r plot_pol_sim3, echo=FALSE}

dpol.sim.pred3 <- plot_fun(dpol.sim.gam3, 
                       c("prn_ann_lograt", "ntr_pland", "aglu_iji",
                         "aglu_frac_mn", "lu_ta"),
                       wfs_sim, "prn_ann_lograt")

ggplot(dpol.sim.pred3[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = dpol.sim.pred3[[2]], 
             mapping = aes(color = prn_ann_lograt), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### Pollinator delta for moderate landscapes (20-50% natural habitat)

#### Correlation between predictors

```{r correlation_mod}

pairs.panels(wfs_mod %>% 
               select(ann_agr_pland, prn_agr_pland, aglu_iji,
                      lu_shdi, aglu_frac_mn, ntr_pland, prn_ann_lograt), 
             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")

```

Like in simple landscapes, strong concurvity.

```{r pollinator_delta_mod}

dpol.mod.gam <- gam(delta_pollinators.margin ~
                  s(ann_agr_pland) +
                  s(aglu_iji) +
                  s(prn_agr_pland) +
                  s(aglu_frac_mn) +
                  s(lu_shdi) + 
                  s(lu_ta), 
                data = wfs_mod)

summary(dpol.mod.gam)

concurvity(dpol.mod.gam, full=FALSE)

```

Partial effects plots and partial residuals

```{r part_plot_mod, echo=FALSE}

dpol.mod.viz <- getViz(dpol.mod.gam)

dpol.mod.vizplot <- plot(dpol.mod.viz, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.mod.vizplot, pages = 1)

```

Predicted effects against actual points. Covariates of each plot(not on x axis) held at moderate landscape averages.

```{r plot_pol_mod, echo=FALSE}

dpol.mod.pred <- plot_fun(dpol.mod.gam, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland",
                         "aglu_frac_mn", "lu_shdi"),
                       wfs_mod, "ann_agr_pland")

ggplot(dpol.mod.pred[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = dpol.mod.pred[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

#### Without perennial ag and landscape iji

These variables increased concurvity when included with annual agriculture

```{r pollinator_delta_mod2}

dpol.mod.gam2 <- gam(delta_pollinators.margin ~
                  s(ann_agr_pland) +
                  # s(aglu_iji) +
                  # s(prn_agr_pland) +
                  s(aglu_frac_mn) +
                  s(lu_shdi) +
                  s(lu_ta), 
                data = wfs_mod)

summary(dpol.mod.gam2)

concurvity(dpol.mod.gam2, full = TRUE)

```

Partial effects plots and partial residuals

```{r part_plot_mod2, echo=FALSE}

dpol.mod.viz2 <- getViz(dpol.mod.gam2)

dpol.mod.vizplot2 <- plot(dpol.mod.viz2, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.mod.vizplot2, pages = 1)

```

Predicted effects against actual points. Covariates of each plot(not on x axis) held at moderate landscape averages.

```{r plot_pol_mod2, echo=FALSE}

dpol.mod.pred2 <- plot_fun(dpol.mod.gam2, 
                       c("ann_agr_pland", "lu_shdi",
                         "aglu_frac_mn", "lu_ta"),
                       wfs_mod, "ann_agr_pland")

ggplot(dpol.mod.pred2[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = dpol.mod.pred2[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

#### A composite variable of log ratio of annual to perennial ag

Using the composite variable isn't as helpful now because the log ratio of perennial to annual ag is correlated with natural habitat, iji, and Shannon diversity.

```{r pollinator_delta_mod3}

dpol.mod.gam3 <- gam(delta_pollinators.margin ~
                  s(prn_ann_lograt) +
                  s(ntr_pland) +
                  # s(aglu_iji) +
                  s(aglu_frac_mn) +
                  # s(lu_shdi) +
                  s(lu_ta), 
                data = wfs_mod)

summary(dpol.mod.gam3)

concurvity(dpol.mod.gam3, full = TRUE)

```

Partial effects plots and partial residuals

```{r part_plot_mod3, echo=FALSE}

dpol.mod.viz3 <- getViz(dpol.mod.gam3)

dpol.mod.vizplot3 <- plot(dpol.mod.viz3, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.mod.vizplot3, pages = 1)

```

Predicted effects against actual points. Covariates of each plot(not on x axis) held at moderate landscape averages.

```{r plot_pol_mod3, echo=FALSE}

dpol.mod.pred3 <- plot_fun(dpol.mod.gam3, 
                       c("prn_ann_lograt", "ntr_pland", 
                         "aglu_frac_mn", "lu_ta"),
                       wfs_mod, "prn_ann_lograt")

ggplot(dpol.mod.pred3[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = dpol.mod.pred3[[2]], 
             mapping = aes(color = prn_ann_lograt), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

### Pollinator delta for natural landscapes (>50% natural)

#### Correlation between predictors

```{r correlation_nat}

pairs.panels(wfs_nat %>% 
               select(ann_agr_pland, prn_agr_pland, aglu_iji,
                      lu_shdi, aglu_frac_mn, ntr_pland, prn_ann_lograt), 
             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")

```

```{r pollinator_delta_nat}

dpol.nat.gam <- gam(delta_pollinators.margin ~
                  s(ann_agr_pland) +
                  s(aglu_iji) +
                  s(prn_agr_pland) +
                  s(aglu_frac_mn) +
                  s(lu_shdi) + 
                  s(lu_ta), 
                data = wfs_nat)

summary(dpol.nat.gam)

concurvity(dpol.nat.gam, full = FALSE)

```

Partial effects plots and partial residuals

```{r part_plot_nat, echo=FALSE}

dpol.nat.viz <- getViz(dpol.nat.gam)

dpol.nat.vizplot <- plot(dpol.nat.viz, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.nat.vizplot, pages = 1)

```

Predicted effects against actual points. Covariates of each plot(not on x axis) held at natural landscape averages.

```{r plot_pol_nat, echo=FALSE}

dpol.nat.pred <- plot_fun(dpol.nat.gam, 
                       c("ann_agr_pland", "aglu_iji", "prn_agr_pland",
                         "aglu_frac_mn", "lu_shdi", "lu_ta"),
                       wfs_nat, "ann_agr_pland")

ggplot(dpol.nat.pred[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = dpol.nat.pred[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

#### Without landscape iji

IJI increased concurvity when included with annual agriculture.

```{r pollinator_delta_nat2}

dpol.nat.gam2 <- gam(delta_pollinators.margin ~
                  s(ann_agr_pland) +
                  # s(aglu_iji) +
                  s(prn_agr_pland) +
                  s(aglu_frac_mn) +
                  s(lu_shdi) +
                  s(lu_ta), 
                data = wfs_nat)

summary(dpol.nat.gam2)

concurvity(dpol.nat.gam2, full = TRUE)

```

Partial effects plots and partial residuals

```{r part_plot_nat2, echo=FALSE}

dpol.nat.viz2 <- getViz(dpol.nat.gam2)

dpol.nat.vizplot2 <- plot(dpol.nat.viz2, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.nat.vizplot2, pages = 1)

```

Predicted effects against actual points. Covariates of each plot(not on x axis) held at natural landscape averages.

```{r plot_pol_nat2, echo=FALSE}

dpol.nat.pred2 <- plot_fun(dpol.nat.gam2, 
                       c("ann_agr_pland", "lu_shdi", "prn_agr_pland",
                         "aglu_frac_mn", "lu_ta"),
                       wfs_nat, "ann_agr_pland")

ggplot(dpol.nat.pred2[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = dpol.nat.pred2[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

#### A composite variable of log ratio of annual to perennial ag

Dropping IJI as before, but also Shannon diversity because it correlates with natural habitat

```{r pollinator_delta_nat3}

dpol.nat.gam3 <- gam(delta_pollinators.margin ~
                  s(prn_ann_lograt) +
                  s(ntr_pland) +
                  s(aglu_frac_mn) +
                  s(lu_ta), 
                data = wfs_nat)

summary(dpol.nat.gam3)

concurvity(dpol.nat.gam3, full = TRUE)

```

Partial effects plots and partial residuals

```{r part_plot_nat3, echo=FALSE}

dpol.nat.viz3 <- getViz(dpol.nat.gam3)

dpol.nat.vizplot3 <- plot(dpol.nat.viz3, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.nat.vizplot3, pages = 1)

```

Predicted effects against actual points. Covariates of each plot(not on x axis) held at natural landscape averages.

```{r plot_pol_nat3, echo=FALSE}

dpol.nat.pred3 <- plot_fun(dpol.nat.gam3, 
                       c("prn_ann_lograt", "ntr_pland",
                         "aglu_frac_mn", "lu_ta"),
                       wfs_nat, "prn_ann_lograt")

ggplot(dpol.nat.pred3[[1]], aes(x, delta_pollinators.margin)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = dpol.nat.pred3[[2]], 
             mapping = aes(color = prn_ann_lograt), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  egg::theme_article()

```

## A complete dataset pollinator model attempt

#### Correlation between independent variables in all landscapes.

```{r correlation_all, echo = FALSE}

pairs.panels(wfs_lsm %>% 
               select(ann_agr_pland, prn_agr_pland, aglu_iji,
                      lu_shdi, aglu_frac_mn, ntr_pland, prn_ann_lograt), 
             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")

```

```{r pol_complete}

dpol.all.gam <- gam(delta_pollinators.margin ~
                  te(prn_ann_lograt,ntr_pland) +
                  s(aglu_frac_mn) +
                  # s(aglu_iji) +
                  # s(lu_shdi) +
                  s(lu_ta), 
                data = wfs_lsm)

summary(dpol.all.gam)

concurvity(dpol.all.gam, full = TRUE)

```

Partial effects plots and partial residuals

```{r part_plot_all, echo=FALSE}
library(rgl)

dpol.all.viz <- getViz(dpol.all.gam)

dpol.all.vizplot <- plot(dpol.all.viz, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar()

print(dpol.all.vizplot, pages = 1)

plotRGL(sm(dpol.all.viz, 1), fix = c("z" = 0), residuals = TRUE)
```

### 2025-01-15 docs/R/utils-plot.R

@@ -1,16 +1,23 @@
-plot_fun <- function(mod, pred.vars, mod.data, color_by = "margin_dens"){
-  predict.list <- lapply(pred.vars,\(x){ggeffects::ggpredict(mod, x)})
+plot_fun <- function(mod, pred.vars, mod.data, color_by = "margin_dens", return_data = TRUE, ...){
+  predict.list <- lapply(pred.vars,\(x){ggeffects::ggpredict(mod, x, ...)})
   names(predict.list) <- pred.vars
   
   pred.trends <- dplyr::bind_rows(predict.list, .id = "pred.var") %>%
     dplyr::rename_with(~as.character(mod$terms[[2]])[length(as.character(mod$terms[[2]]))],
                        predicted)
   
-  pred.data <- mod.data %>%
-    select(all_of(c(as.character(mod$terms[[2]])[length(as.character(mod$terms[[2]]))], pred.vars))) %>%
-    mutate(color_by = get(color_by)) %>%
-    pivot_longer(all_of(pred.vars), names_to = "pred.var", values_to = "x") %>%
-    rename(!!color_by := color_by)
+  if(return_data){
+    pred.data <- mod.data %>%
+      select(all_of(c(as.character(mod$terms[[2]])[length(as.character(mod$terms[[2]]))], pred.vars))) %>%
+      mutate(color_by = get(color_by)) %>%
+      pivot_longer(all_of(pred.vars), names_to = "pred.var", values_to = "x") %>%
+      rename(!!color_by := color_by)
+  }
   
-  list(pred.trends, pred.data)
+  if(return_data){
+    list(pred.trends, pred.data)
+  }else{
+    pred.trends
+  }
+  
 }

### 2025-02-01 landscapemetrics_quantile.Rmd
quantile gams and pca

---
title: "Quantile regression"
---

Quantile regression looks at relationships at different quantiles of response.

```{r setup, echo = FALSE, message=FALSE}
library(tidyverse)
library(qgam)
library(mgcViz)
library(ggfortify)

wfs_lsm <- readRDS("data/wfs_lsm.rds")

```

```{r qgam_fit}
fit <- mqgamV(delta_pollinators.margin ~
                  te(prn_ann_lograt,ntr_pland) +
                  s(aglu_frac_mn) +
                  s(aglu_iji) +
                  s(lu_shdi),
                  # s(lu_ta), 
                data = wfs_lsm,
              qu = c(0.2,0.5,0.8))

```

```{r qgam_vis}
print(plot(fit, allTerms = TRUE), pages = 1)
```

Okay that looks similar to what I've already done

```{r margin_relations}

wfs_low <- wfs_lsm %>% filter(ntr_pland < 8.25 & ntr_pland > 7.75)

pairs.panels(wfs_lsm %>% 
               select(delta_pollinators.margin, delta_qb.margin, 
                      delta_n_export.margin, delta_sed.margin), 
             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")

```

```{r es_relations}

pairs.panels(wfs_lsm %>% 
               select(pollinators_base, qb_base, 
                      n_export_base, sed_export_base), 
             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")

```

```{r quant_cut}

wfs_grp <- wfs_lsm %>%
  mutate(habcat = cut(ntr_pland, breaks = c(-Inf,20,50,Inf),
                      labels = c("simple","moderate","natural")))
```

```{r grp_pca_smp}

wfs_pca_all <- prcomp(wfs_grp %>%
                        select(ntr_pland, ann_agr_pland, prn_agr_pland, 
                               aglu_iji, aglu_frac_mn),
                      scale. = TRUE, center = TRUE)

summary(wfs_pca_all)

```

```{r grp_pca_mod}

wfs_mod_pca <- prcomp(wfs_grp %>% select(ntr_pland, ann_agr_pland, prn_agr_pland, 
                                         aglu_iji, aglu_frac_mn),
                      scale. = TRUE, center = TRUE)

autoplot(wfs_pca_all, data = wfs_grp, color = "habcat", 
         loadings = TRUE,
         loadings.label.label,
         choices = c(2,3))

ggbiplot::ggbiplot(wfs_pca_all, groups = wfs_grp$habcat)
ggbiplot::ggbiplot(wfs_pca_all, groups = wfs_grp$habcat, choices = c(2,3))

```

### 2025-02-17 landscapemetrics_quantile.Rmd
update analysis

@@
                         select(ntr_pland, ann_agr_pland, prn_agr_pland, 
-                               aglu_iji, aglu_frac_mn),
+                               aglu_iji, aglu_frac_mn, lu_shdi),
                       scale. = TRUE, center = TRUE)
@@
 
-wfs_mod_pca <- prcomp(wfs_grp %>% select(ntr_pland, ann_agr_pland, prn_agr_pland, 
-                                         aglu_iji, aglu_frac_mn),
-                      scale. = TRUE, center = TRUE)
-
-autoplot(wfs_pca_all, data = wfs_grp, color = "habcat", 
-         loadings = TRUE,
-         loadings.label.label,
-         choices = c(2,3))
-
 ggbiplot::ggbiplot(wfs_pca_all, groups = wfs_grp$habcat)
@@
 ```
+
+```{r grp_pca_smp2}
+
+wfs_pca_all2 <- prcomp(wfs_grp %>%
+                        select(ntr_pland, ann_agr_pland, prn_agr_pland, 
+                               aglu_iji, aglu_frac_mn),
+                      scale. = TRUE, center = TRUE)
+
+summary(wfs_pca_all2)
+
+```
+
+```{r grp_pca_mod2}
+
+ggbiplot::ggbiplot(wfs_pca_all2, groups = wfs_grp$habcat)
+ggbiplot::ggbiplot(wfs_pca_all2, groups = wfs_grp$habcat, choices = c(2,3))
+
+```
+
+```{r create_pc_var}
+
+lsm_pc <- wfs_grp %>% cbind(predict(wfs_pca_all))
+
+ggplot(lsm_pc, aes(PC1, pollinators_base, colour = habcat)) + geom_point()
+
+ggplot(lsm_pc, aes(PC3, sed_export_base, colour = habcat)) + geom_point()
+
+```

### 2025-02-19 landscapemetrics_quantile.Rmd
render site

@@
 ---
-title: "Quantile regression"
+title: "Update for Feb 19, 2025"
 ---
 
-Quantile regression looks at relationships at different quantiles of response.
+## Quantile regression
+
+Quantile regression looks at relationships at different quantiles of response. Below I fit the change in pollinator abundance (per added margin area) to these measures of landscape complexity:
+
+- `ntr_pland` percent natural land
+- `prn_ann_lograt` log-ratio of perennial to annual agriculture. Here a positive value means a greater proportion of perennial agriculture.
+- `aglu_frac_mn` average fractal dimension index
+- `aglu_iji` interspersion and juxtaposition index
+- `lu_shdi` Shannon diversity of land use
 
@@
 
-wfs_lsm <- readRDS("data/wfs_lsm.rds")
+wfs_lsm <- readRDS("data/wfs_lsm.rds") %>%
+  mutate(pollinators_pcchng = (pollinators_max - pollinators_base)/pollinators_base,
+         n_export_pcchng = (n_export_max - n_export_base)/n_export_base,
+         qb_pcchng = (qb_max - qb_base)/qb_base,
+         sed_export_pcchng = (sed_export_max - sed_export_base)/sed_export_base,)
 
@@
 
-```{r qgam_fit}
-fit <- mqgamV(delta_pollinators.margin ~
-                  te(prn_ann_lograt,ntr_pland) +
+I ran quantile GAM regression for 0.1, 0.2, 0.5, 0.8, and 0.9 quantiles.
+I included an interaction between `ntr_pland` and `prn_ann_lograt` because: 1) it is of interest whether the role of perennial vs annual ag changes at different levels of naturalness, and 2) `ntr_pland` and `prn_ann_lograg` still co-vary, even though I made this index to try to create a variable that is more independent from the main agricultural-natural gradient.
+I think the interaction helps mitigate concurvity a little.
+
+### Results
+
+```{r qgam_fit, message=FALSE}
+qgam.fit <- mqgamV(delta_pollinators.margin ~
+                  te(ntr_pland, prn_ann_lograt) +
                   s(aglu_frac_mn) +
@@
                   s(lu_shdi),
-                  # s(lu_ta), 
                 data = wfs_lsm,
-              qu = c(0.2,0.5,0.8))
+              qu = c(0.1, 0.2,0.5,0.8, 0.9))
 
@@
 
-```{r qgam_vis}
-print(plot(fit, allTerms = TRUE), pages = 1)
+Below are the partial effects plots for the range of modeled quantiles.
+
+```{r qgam_vis, echo = FALSE}
+print(plot(qgam.fit, allTerms = TRUE), pages = 1)
 ```
 
-Okay that looks similar to what I've already done
+These results don't suggest anything obviously different happening at different quantiles. Some of the extreme quantiles are different, but I don't know how much to trust those trends.
 
-```{r margin_relations}
+Looking at concurvity of the same model formula fit using `mgcv`, there is still some concurvity.
 
-wfs_low <- wfs_lsm %>% filter(ntr_pland < 8.25 & ntr_pland > 7.75)
+```{r concurvity, echo = FALSE, message = FALSE}
 
-pairs.panels(wfs_lsm %>% 
-               select(delta_pollinators.margin, delta_qb.margin, 
-                      delta_n_export.margin, delta_sed.margin), 
-             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")
+mcgv.fit <- gam(delta_pollinators.margin ~
+                  te(ntr_pland, prn_ann_lograt) +
+                  s(aglu_frac_mn) +
+                  s(aglu_iji) +
+                  s(lu_shdi),
+                data = wfs_lsm)
+
+concurv.tbl <- gratia::model_concurvity(mcgv.fit)
+
+knitr::kable(concurv.tbl)
 
@@
 
-```{r es_relations}
+I wonder if this is because all these metrics are calculated from the same base land use data.
+I expect the metrics of these watershed naturally have some correlations that limit the inferences we can make.
+This led me to try to identify these associations using PCA.
 
-pairs.panels(wfs_lsm %>% 
-               select(pollinators_base, qb_base, 
-                      n_export_base, sed_export_base), 
-             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")
+## PCA of landscape metrics
 
-```
+I ran PCA on the the following 6 landscape metrics:
 
-```{r quant_cut}
+- `ntr_pland` percent natural land
+- `ann_agr_pland` annual agriculture percent landscape
+- `prn_agr_pland` perennial agriculture percent landscape
+- `aglu_frac_mn` average fractal dimension index
+- `aglu_iji` interspersion and juxtaposition index
+- `lu_shdi` Shannon diversity of land use
+
+I included all the major land use groups since the point was to visualize their associations.
+
+```{r quant_cut, echo = FALSE}
 
@@
 
-```{r grp_pca_smp}
+### Results
+
+The first four PCs explain 95% of variation.
+
+```{r grp_pca_smp, echo = FALSE}
 
@@
 
+wfs_pca_all
+
 summary(wfs_pca_all)
@@
 ```
+PC1 positivley weights % natural, along with Shannon diversity and perennial agriculture to lesser degree, and negatively weights annual agriculture, fractional dimension index, and interspersion and juxtaposition index (IJI).
+PC1 strongly represents the natural vs (annual) agricultural gradient across the watersheds.
+This is evident in the biplot below, which has been colored according to simple, moderate, and natural landscape categories.
 
-```{r grp_pca_mod}
+```{r PC12_biplot, echo = FALSE}
 
 ggbiplot::ggbiplot(wfs_pca_all, groups = wfs_grp$habcat)
+
+```
+
+PC2 negatively and strongly weights Shannon diversity, IJI, and perennial agriculture, while positively and moderately weighting natural land cover.There is an arch shape to the point distribution of the PC1 and PC2 axes.
+The arch is oriented so that the negative and positive extremes of PC1 (which I'm interpreting as highly agricultural and highly natural landscapes, respectively), are both positive on the PC2 axis.
+I think positive PC2 values represent landscape homegeneity.
+Highly agricultural and highly natural landscapes are both homogeneous.
+Conversely, negative PC2 values are more heterogeneous, and correspond with higher Shannon diversity, interspersion, and perennial agriculture (maybe incidentally).
+
+I am not sure how to interpret PC3. It weights perennial agriculture and fractal dimension against interspersion/juxtaposition and natural land cover.
+
+```{r PC23_biplot, echo = FALSE}
+
 ggbiplot::ggbiplot(wfs_pca_all, groups = wfs_grp$habcat, choices = c(2,3))
@@
 
-```{r grp_pca_smp2}
-
-wfs_pca_all2 <- prcomp(wfs_grp %>%
-                        select(ntr_pland, ann_agr_pland, prn_agr_pland, 
-                               aglu_iji, aglu_frac_mn),
-                      scale. = TRUE, center = TRUE)
-
-summary(wfs_pca_all2)
-
-```
-
-```{r grp_pca_mod2}
-
-ggbiplot::ggbiplot(wfs_pca_all2, groups = wfs_grp$habcat)
-ggbiplot::ggbiplot(wfs_pca_all2, groups = wfs_grp$habcat, choices = c(2,3))
-
-```
+## Predict change in pollinators using PC1 and PC2
 
@@
 
-ggplot(lsm_pc, aes(PC1, pollinators_base, colour = habcat)) + geom_point()
+pol.delt.lm1 <- lm(delta_pollinators.margin ~ PC1 + PC2, data = lsm_pc)
 
-ggplot(lsm_pc, aes(PC3, sed_export_base, colour = habcat)) + geom_point()
+pol.delt.lm2 <- lm(delta_pollinators.margin ~ PC1 * PC2, data = lsm_pc)
+
+performance::compare_performance(pol.delt.lm1,pol.delt.lm2)
+
+summary(pol.delt.lm2)
 
 ```
+
+```{r prediction_delta}
+
+pred.PC1 <- ggeffects::ggpredict(pol.delt.lm2, terms = c("PC1","PC2")) %>%
+  rename(PC1 = x, delta_pollinators.margin = predicted, PC2 = group)
+
+pred.PC2 <- ggeffects::ggpredict(pol.delt.lm2, terms = c("PC2","PC1")) %>%
+  rename(PC2 = x, delta_pollinators.margin = predicted, PC1 = group)
+
+ggplot(lsm_pc, aes(PC1, delta_pollinators.margin)) + 
+  geom_point(aes(color = PC2)) + scale_color_continuous(type = "viridis") + 
+  geom_ribbon(data = pred.PC1, 
+              mapping = aes(ymin = conf.low, ymax = conf.high, group = PC2, fill = PC2), alpha = 0.2) + egg::theme_article()
+
+ggplot(lsm_pc, aes(PC2, delta_pollinators.margin)) + 
+  geom_point(aes(color = PC1)) + scale_color_continuous(type = "viridis") + 
+  geom_ribbon(data = pred.PC2, 
+              mapping = aes(ymin = conf.low, ymax = conf.high, group = PC1, fill = PC1), alpha = 0.2) + egg::theme_article()
+
+```
+
+## Predict base pollinator abundance with PC1 and PC2
+
+```{r pol.base}
+
+pol.base.lm1 <- lm(pollinators_base ~ PC1 + PC2, data = lsm_pc)
+
+pol.base.lm2 <- lm(pollinators_base ~ PC1 * PC2, data = lsm_pc)
+
+performance::compare_performance(pol.base.lm1,pol.base.lm2)
+
+summary(pol.base.lm2)
+
+```
+
+```{r }
+
+pred.base.PC1 <- ggeffects::ggpredict(pol.base.lm2, terms = c("PC1","PC2")) %>%
+  rename(PC1 = x, pollinators_base = predicted, PC2 = group)
+
+pred.base.PC2 <- ggeffects::ggpredict(pol.base.lm2, terms = c("PC2","PC1")) %>%
+  rename(PC2 = x, pollinators_base = predicted, PC1 = group)
+
+ggplot(lsm_pc, aes(PC1, pollinators_base)) + 
+  geom_point(aes(color = PC2)) + scale_color_continuous(type = "viridis") + 
+  geom_ribbon(data = pred.base.PC1, 
+              mapping = aes(ymin = conf.low, ymax = conf.high, group = PC2, fill = PC2), alpha = 0.2) + egg::theme_article()
+
+ggplot(lsm_pc, aes(PC2, pollinators_base)) + 
+  geom_point(aes(color = PC1)) + scale_color_continuous(type = "viridis") + 
+  geom_ribbon(data = pred.base.PC2, 
+              mapping = aes(ymin = conf.low, ymax = conf.high, group = PC1, fill = PC1), alpha = 0.2) + egg::theme_article()
+
+```
+
+## Relationships between base ecosystem service values and change in ES normalized by margin area
+
+```{r margin_relations, out.width="100%"}
+
+psych::pairs.panels(wfs_lsm %>% 
+               select(margin_dens,
+                      pollinators_base, qb_base, 
+                      n_export_base, sed_export_base,
+                      delta_pollinators.margin, delta_qb.margin,
+                      delta_n_export.margin, delta_sed.margin),
+             cex.cor = 1, cex.labels=1, cex=0.5, method = "pearson")
+
+```

### 2025-02-26 landscapemetrics_onefield.Rmd
add one field data and documentation

---
title: "One Field"
output: html_document
date: "2025-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Add one field

```{r load}
library(tidyverse)
library(psych)
library(egg)

wfs_lsm <- readRDS("data/wfs_lsm.rds") 
of_lsm <- readRDS("data/of_lsm.rds")
of_lsm_margin.med <- readRDS("data/of_lsm_marginmed.rds")

```

```{r recombine_data}
wfs_lsm_pc <- readRDS("data/wfs_lsm_pc.rds")

of_lsm.med <- of_lsm %>% 
  group_by(hu) %>%
  summarize(
    across(delta_pollinators:delta_sed, 
           list(mean = mean, sd = sd, median = median),
           .names = "{.col}_of.{.fn}")
  )

wfs_lsm_of <- wfs_lsm_pc %>% 
  left_join(
    of_lsm_margin.med %>%
      select(hu, 
             delta_pollinators_of.margin.median,
             delta_qb_of.margin.median,
             delta_n_export_of.margin.median,
             delta_sed_of.margin.median
             )
  ) %>% 
  left_join(
    of_lsm.med %>%
      select(hu, 
             delta_pollinators_of.median,
             delta_qb_of.median,
             delta_n_export_of.median,
             delta_sed_of.median
             )
  )

pairs.panels(
  wfs_lsm_of %>% 
    select(
      delta_pollinators.margin,
      delta_qb.margin,
      delta_n_export.margin,
      delta_sed.margin,
      delta_pollinators_of.margin.median,
      delta_qb_of.margin.median,
      delta_n_export_of.margin.median,
      delta_sed_of.margin.median
    )
)

```

```{r pol}

pol_long <- wfs_lsm_of %>% 
  pivot_longer(cols= c(delta_pollinators.margin, delta_pollinators_of.margin.median),
               names_to = "treatment",
               names_prefix = "delta_pollinators.")

pol_long.ws <- wfs_lsm_of %>% 
  pivot_longer(cols= c(delta_pollinators, delta_pollinators_of.median),
               names_to = "treatment", values_to = "pollinators",
               names_prefix = "delta_pollinators.") %>%
  mutate(pollinators_ws = pollinators * lu_ta)

ggplot(pol_long, aes(ntr_pland, value, color = treatment)) + geom_point()
ggplot(pol_long.ws, aes(PC1, pollinators_ws, color = treatment)) + geom_point()

ggplot(wfs_lsm_of, aes(delta_pollinators.margin,delta_pollinators_of.margin.median)) + geom_point()

ggplot(wfs_lsm_of, aes(ntr_pland,delta_pollinators)) + geom_point()
ggplot(wfs_lsm_of, aes(ntr_pland,delta_pollinators_of.median)) + geom_point()

ggplot(wfs_lsm_of, aes(PC1,delta_pollinators)) + geom_point()
ggplot(wfs_lsm_of, aes(ntr_pland,delta_n_export)) + geom_point()

ggplot(wfs_lsm_of, aes(aglu_frac_mn,delta_pollinators.margin, color = habcat)) + geom_point()
ggplot(wfs_lsm_of, aes(PC1,delta_pollinators.margin)) + geom_point()
ggplot(wfs_lsm_of, aes(PC1,delta_pollinators_of.margin.median)) + geom_point()
```

```{r n_exp}

nexp_long <- wfs_lsm_of %>% 
  pivot_longer(cols= c(delta_n_export.margin, delta_n_export_of.margin.median),
               names_to = "treatment",
               names_prefix = "delta_n_export.")

ggplot(nexp_long, aes(PC1, value, color = treatment)) + geom_point()

```

```{r panel_base}

onefield_long <- wfs_lsm_of %>% 
  pivot_longer(cols= c(pollinators_base,
                       n_export_base,
                       qb_base,
                       sed_export_base),
               names_to = "Ecosystem service",
               values_to = "ES value")

ggplot(onefield_long, aes(PC2, `ES value`)) + geom_point() + 
  facet_wrap(~`Ecosystem service`, scales = "free_y")

```

```{r panel_onefield}

onefield_long <- wfs_lsm_of %>% 
  pivot_longer(cols= c(delta_pollinators_of.median,
                       delta_n_export_of.median,
                       delta_qb_of.median,
                       delta_sed_of.median),
               names_to = "Ecosystem service",
               values_to = "ES value",
               names_prefix = "delta_") %>%
  mutate(`ES value` = `ES value`*lu_ta)

ggplot(onefield_long, aes(PC2, `ES value`)) + geom_point() + 
  facet_wrap(~`Ecosystem service`, scales = "free_y")

```

```{r panel_max}
max_long <- wfs_lsm_of %>% 
  pivot_longer(cols= c(delta_pollinators,
                       delta_n_export,
                       delta_qb,
                       delta_sed),
               names_to = "Ecosystem service",
               values_to = "ES value",
               names_prefix = "delta_") %>%
  mutate(`ES value` = `ES value`*lu_ta)%>%
  mutate(
    `Ecosystem service` = fct_recode(
      `Ecosystem service`,
      "Pollinators" = "pollinators",
      "N export reduced" = "n_export",
      "Water recharge" = "qb",
      "Sediment reduced" = "sed"
    )
  )

ggplot(max_long, aes(PC1, `ES value`, color = "Treatment")) + geom_point(alpha = 0.5) + 
  facet_wrap(~`Ecosystem service`, scales = "free_y") + 
  theme_article() +
  theme(strip.text = element_text(size = 12))

```

```{r panel_onefield_margin}

onefield_long.margin <- wfs_lsm_of %>% 
  pivot_longer(cols= c(delta_pollinators_of.margin.median,
                       delta_n_export_of.margin.median,
                       delta_qb_of.margin.median,
                       delta_sed_of.margin.median),
               names_to = "Ecosystem service",
               values_to = "ES value",
               names_prefix = "delta_") %>%
  # filter(!(`Ecosystem service` == "sed_of.margin.median" & `ES value` > 60)) %>%
  mutate(
    `Ecosystem service` = fct_recode(
      `Ecosystem service`,
      "Pollinators" = "pollinators_of.margin.median",
      "N export reduced" = "n_export_of.margin.median",
      "Water recharge" = "qb_of.margin.median",
      "Sediment reduced" = "sed_of.margin.median"
    )
  )

ggplot(onefield_long.margin, aes(PC1, `ES value`, color = habcat)) + 
  geom_point() + 
  facet_wrap(~`Ecosystem service`, scales = "free_y") +
  theme_article()

ggplot(onefield_long.margin, aes(PC2, `ES value`, color = habcat)) + geom_point() + 
  facet_wrap(~`Ecosystem service`, scales = "free_y") +
  theme_article()

```

```{r panel_max_margin}

max_long.margin <- wfs_lsm_of %>% 
  pivot_longer(cols= c(delta_pollinators.margin,
                       delta_n_export.margin,
                       delta_qb.margin,
                       delta_sed.margin),
               names_to = "Ecosystem service",
               values_to = "ES value",
               names_prefix = "delta_") %>%
  mutate(
    `Ecosystem service` = fct_recode(
      `Ecosystem service`,
      "Pollinators" = "pollinators.margin",
      "N export reduced" = "n_export.margin",
      "Water recharge" = "qb.margin",
      "Sediment reduced" = "sed.margin"
    )
  )

ggplot(max_long.margin, aes(PC1, `ES value`, color = habcat)) +
  geom_point() + 
  facet_wrap(~`Ecosystem service`, scales = "free_y") +
  theme_article() +
  theme(strip.text = element_text(size = 12))

ggplot(max_long.margin, aes(PC2, `ES value`, color = habcat)) + 
  geom_point() + 
  facet_wrap(~`Ecosystem service`, scales = "free_y") +
  theme_article() +
  theme(strip.text = element_text(size = 12))

```

```{r both_types_margin}
both_long.margin <- bind_rows(
  onefield_long.margin %>% mutate(Treatment = "One field"),
  max_long.margin %>% mutate(Treatment = "Max fields")
)

ggplot(both_long.margin, aes(PC1, `ES value`, color = Treatment)) +
  geom_point(alpha = 0.5) + 
  facet_wrap(~`Ecosystem service`, scales = "free_y") +
  theme_article() +
  theme(strip.text = element_text(size = 12))

ggplot(both_long.margin, aes(PC2, `ES value`, color = Treatment)) +
  geom_point(alpha = 0.5) + 
  facet_wrap(~`Ecosystem service`, scales = "free_y") +
  theme_article() +
  theme(strip.text = element_text(size = 12))

```

```{r es_pca}

es_base_pca <- prcomp(wfs_lsm_of %>%
                        select(pollinators_base:sed_export_base),
                      scale. = TRUE, center = TRUE)

es_base_pca

summary(es_base_pca)

ggbiplot::ggbiplot(es_base_pca, groups = wfs_lsm_of$habcat) +
  egg::theme_article()

ggbiplot::ggbiplot(es_base_pca, groups = wfs_lsm_of$habcat, choices = c(2,3)) +
  egg::theme_article()
```

```{r es_margin_pca}

es_max.margin_pca <- prcomp(wfs_lsm_of %>%
                        select(delta_pollinators.margin:delta_sed.margin),
                      scale. = TRUE, center = TRUE)

es_max.margin_pca

summary(es_max.margin_pca)

ggbiplot::ggbiplot(es_max.margin_pca, groups = wfs_lsm_of$habcat) +
  egg::theme_article()

ggbiplot::ggbiplot(es_max.margin_pca, groups = wfs_lsm_of$habcat, choices = c(2,3)) +
  egg::theme_article()
```

```{r model}
library(brms)

both_wide.margin <- both_long.margin %>%
  pivot_wider(id_cols = c(hu, Treatment),values_from = `ES value`, names_from = `Ecosystem service`) %>% left_join(wfs_lsm_pc, by = "hu") %>%
  rename(
      pollinators = "Pollinators",
      N_exp_red = "N export reduced",
      water_rech = "Water recharge",
      sed_exp_red = "Sediment reduced"
  )

allmod <- brm(mvbind(pollinators, N_exp_red, water_rech, sed_exp_red) ~ Treatment * PC1 * PC2 + (1|p|hu), data = both_wide.margin)

allmod2 <- brm(mvbind(pollinators, N_exp_red, water_rech, sed_exp_red) ~ Treatment + t2(PC1, PC2, by = Treatment) + (1|p|hu), data = both_wide.margin)

```

### 2025-02-26 data-raw/of_lsm.R

@@ -1,0 +1,33 @@
+## code to prepare `of_lsm` dataset
+#' Joins one field results (10 per watershed) with landscape metrics dataset
+#' Calculates change in ES from baseline normalized by margin area
+#' Result has 4370 rows
+
+onefield <- readRDS("data/hu_onefield.rds")
+
+of_lsm <- onefield %>%
+  left_join(wfs_lsm %>% 
+              select(hu, lu_ta, pollinators_base:sed_export_base)) %>%
+  mutate(margin.ha = (margin_cells*900)/10000,
+         margin_den = margin.ha/lu_ta) %>%
+  mutate(delta_pollinators = pollinators - pollinators_base, # calculate change from baseline (ES values are normalized by total ws area)
+         delta_n_export = -(n_export - n_export_base),
+         delta_qb = qb - qb_base,
+         delta_sed = -(sed_export - sed_export_base)) %>%
+  mutate(across(delta_pollinators:delta_sed, \(x) x/margin_den, .names = "{.col}_of.margin"))  # normalize by margin density results in change in es per margin area (ws area cancels)
+
+saveRDS(of_lsm, "data/of_lsm.rds")
+
+## median dataset
+#' take the median ES change value of each watershed
+#' use the margin-normalized values
+
+of_lsm_margin.med <- of_lsm %>% 
+  group_by(hu) %>%
+  summarize(
+    across(delta_pollinators_of.margin:delta_sed_of.margin, 
+           list(mean = mean, sd = sd, median = median),
+           .names = "{.col}.{.fn}")
+  )
+
+saveRDS(of_lsm_margin.med, "data/of_lsm_marginmed.rds")

### 2025-03-05 data-raw/examine_sim_results.Rmd
landscape characteristics

---
title: "examine simulation results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##

```{r load}
wfs_lsm <- readRDS("data/wfs_lsm.rds")
hu_meta <- readRDS("data/hu_meta.rds")

hulandstats <- readRDS("data/hulandstats.rds")

wfs_meta <- wfs_lsm %>% 
  left_join(hu_meta %>% sf::st_drop_geometry(), 
            by = c("hu" = "huc12", "huc2", "name")) %>%
  left_join(hulandstats)

```

## Plots

```{r wetland}
ggplot(wfs_meta, aes(Wetland, delta_sed.margin)) + 
  geom_point()
```

```{r erosivity}
ggplot(wfs_meta, aes(erosivity_mean, delta_sed.margin)) + 
  geom_point()
```

```{r erodibility}
ggplot(wfs_meta, aes(erodibility_mean, delta_sed.margin)) + 
  geom_point()
```

```{r erodibility_2}
ggplot(wfs_meta, aes(ntr_pland, erodibility_mean)) + 
  geom_point()
```

### 2025-03-05 data-raw/hu_meta.R

@@ -1,0 +1,7 @@
+## code to prepare `hu_meta` dataset goes here
+
+datadir <- "~/Box Sync/InVEST/GIS_data/"
+
+hu_meta <- sf::st_read(paste0(datadir,"NEMWregion/selected_hu12_v3_candidates.gpkg"))
+
+saveRDS(hu_meta, "data/hu_meta.rds")

### 2025-03-05 data-raw/examine_sim_results.Rmd
update site

@@
 ```{r erodibility_2}
-ggplot(wfs_meta, aes(ntr_pland, erodibility_mean)) + 
-  geom_point()
+library(plotly)
+fig <- plot_ly(wfs_meta, x = ~erodibility_mean, y = ~erosivity_mean, z = ~delta_sed.margin, color = ~delta_sed.margin, text = ~hu)
+
+fig
+
 ```
+
+```{r sediment_change}
+ggplot(wfs_meta, aes(sed_export_base, sed_export_max, color = delta_sed.margin)) + 
+  geom_point() + geom_abline(intercept = 0, slope = 1) + scale_color_continuous(type="viridis")
+
+```
+
+```{r nutrient_change}
+ggplot(wfs_meta, aes(n_export_base, n_export_max, color = delta_n_export.margin)) + 
+  geom_point() + geom_abline(intercept = 0, slope = 1) + scale_color_continuous(type="viridis")
+
+```
+
+```{r pollinator_change}
+ggplot(wfs_meta, aes(pollinators_base, pollinators_max, color = delta_pollinators.margin)) + 
+  geom_point() + geom_abline(intercept = 0, slope = 1) + scale_color_continuous(type="viridis")
+
+```
+
+```{r qb_change}
+ggplot(wfs_meta, aes(qb_base, qb_max, color = delta_qb.margin)) + 
+  geom_point() + geom_abline(intercept = 0, slope = 1) + scale_color_continuous(type="viridis")
+
+```

### 2025-03-05 docs/data-raw/hulandstats.R

@@ -1,0 +1,50 @@
+library(exactextractr)
+library(tidyverse)
+library(terra)
+library(sf)
+
+datadir <- "/Users/kevinli/Box Sync/InVEST/Optimization/Data/maxWFS/hu/"
+hu.files <- list.files(datadir)
+
+hustats <- list()
+for(i in seq_along(hu.files)){
+  hu.i <- gsub("hu","",hu.files[i])
+  hu.files.i <- paste0(datadir, hu.files[i],"/data/")
+  ws.i <- sf::st_read(paste0(hu.files.i,"watershed.gpkg"))
+  
+  stats.i <- list()
+  
+  # elevation
+  elev.i <-terra::rast(paste0(hu.files.i,"elev.tif"))
+  stats.i$elev <- exactextractr::exact_extract(elev.i, ws.i, c("mean","stdev","median"))
+  
+  # erodibility
+  erod.i <- terra::rast(paste0(hu.files.i,"erodibility.tif"))
+  stats.i$erodibility <- exactextractr::exact_extract(erod.i, ws.i, c("mean","stdev","median"))
+  
+  # erosivity
+  eros.i <- terra::rast(paste0(hu.files.i,"erosivity.tif"))
+  stats.i$erosivity <- exactextractr::exact_extract(eros.i, ws.i, c("mean","stdev","median"))
+  
+  # et0
+  et0.i <- terra::rast(list.files(paste0(hu.files.i,"et0/"), full.names = TRUE))
+  stats.i$et0 <- data.frame(mean=apply(exactextractr::exact_extract(et0.i, ws.i, c("mean")),1,mean))
+  
+  # precip
+  precip.i <- terra::rast(list.files(paste0(hu.files.i,"precip/"), full.names = TRUE))
+  stats.i$precip <- data.frame(mean=apply(exactextractr::exact_extract(precip.i, ws.i, c("mean")),1,mean))
+  
+  # rainevents
+  rain.i <- read.csv(paste0(hu.files.i,"rainevents.csv"))
+  stats.i$rainevents <- data.frame(mean=mean(rain.i$events))
+  
+  # combine them all
+  hustats[[hu.i]] <- bind_rows(stats.i, .id = "variable") %>%
+    pivot_wider(names_from = "variable", values_from = c("mean", "stdev", "median"),
+                names_vary = "slowest", names_glue = "{variable}_{.value}") %>%
[…]

### 2025-03-08 scripts/Kc_plots.R
batch scripts

@@ -1,0 +1,33 @@
+library(tidyverse)
+library(ggforce)
+devtools::load_all()
+
+curverefs <- read.csv("data-raw/Crop_evap_reference_FAO.csv")
+
+swy_curves <- espareto::biophys_swy %>%
+  left_join(curverefs %>% select(VALUE,Kc.reference, Kc.ref.crop, Winter.Kc.ref.crop), by = c("lucode"="VALUE")) %>%
+  mutate(cropkctitle = paste0(CLASS_NAME, " (", Kc.reference, ")")) %>% rowwise() %>%
+  mutate(cropkctitle2 = ifelse(Kc.reference == "Double",
+                               paste0(CLASS_NAME, " [", Kc.reference, ", ", Kc.ref.crop, "/", Winter.Kc.ref.crop,"]"),
+                               ifelse(Kc.ref.crop == "",
+                                      paste0(CLASS_NAME, " [", Kc.reference, "]"),
+                                      paste0(CLASS_NAME, " [", Kc.reference, ", ", Kc.ref.crop, "]")))) %>%
+  ungroup() %>%
+  pivot_longer(cols=Kc_1:Kc_12, values_to = "Kc", names_to = "Month",
+               names_transform = list(Month=\(x){as.numeric(gsub("Kc_","",x))}))
+
+i <- ceiling(length(unique(swy_curves$CLASS_NAME)) / 8) # set the number of pages
+
+pdf("crop_Kc_curves.pdf", width = 8.5, height = 11)
+lapply(seq(i), function(page) {
+  SdesGG <- swy_curves %>% #launch each time or does not work
+    group_by(cropkctitle2) %>% #mandatory or need to fortify
+    ggplot(aes(x = Month, y = Kc)) +
+    geom_line() + geom_point() +
+    scale_x_continuous(breaks = 1:12) +
+    coord_cartesian(ylim=c(0,1.4)) +
+    facet_wrap_paginate(~cropkctitle2, ncol = 2, nrow = 4, page = page,
+                        labeller = labeller(cropkctitle2 = label_wrap_gen(60))) #ggforce
+})
+dev.off()
+graphics.off()

### 2025-03-08 scripts/atlas_max_one.R

@@ -1,0 +1,94 @@
+library(devtools)
+library(tidyverse)
+library(sf)
+library(terra)
+library(doFuture)
+library(optimEcoServices)
+
+# projdir <- "/Users/kevinli/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/maxWFS/"
+projdir <- "/90daydata/geoecoservices/optimization/maxWFS/"
+
+datadir <- paste0(projdir,"data/")
+
+# biophys directory
+biophysdir <- paste0(datadir, "biophys")
+
+# onefields
+fields.select <- readRDS(paste0(datadir,"selectfields.rds"))
+
+hu <- st_read(paste0(datadir, "hu12watersheds.gpkg"))
+
+hu.names <- hu$huc12
+
+# set up futures
+plan(multisession)
+
+out <- foreach(i = seq_along(hu.names)) %dofuture% {
+  hu.name <- hu.names[i]
+
+  hudir <- paste0(projdir, "hu/hu",hu.name,"/")
+  hu.fields <- rast(paste0(hudir, "data/lu/fields", ".tif"))
+  hu.margins <- rast(paste0(hudir, "data/lu/margins", ".tif"))
+
+  hu.marginids <- unique(terra::values(hu.margins$id, na.rm=TRUE)) # added unique() because ids are section-based, not cell based
+  basemargin <- terra::ifel(test = !is.na(hu.margins$id), 0, 0)
+  maxmargin <- terra::ifel(test = !is.na(hu.margins$id), 1, 0)
+
+  # baseline
+  baseline <- wfs_fitness2(
+    margin.rast = basemargin,
+    fields.rast = hu.fields,
+    margin.lc = 171,   # land cover for 'pollinator habitat'
+    biophys.dir = biophysdir,
+    data.dir =  paste0(hudir, "data"),
+    workspace.dir = paste0(hudir,"baseline"),
+    suffix="base"
+  )
+
+  # max wfs
+  maxwfs <- wfs_fitness2(
+    margin.rast = maxmargin,
+    fields.rast = hu.fields,
+    margin.lc = 171,   # land cover for 'pollinator habitat'
+    biophys.dir = biophysdir,
+    data.dir =  paste0(hudir, "data"),
+    workspace.dir = paste0(hudir,"max"),
+    suffix="max"
+  )
+  
+  # one field loop
+  hu.fields.select.ids <- fields.select %>%
+    filter(hu == hu.name) %>% pull(id)
+  
[…]

### 2025-03-08 scripts/atlas_test.R

@@ -1,0 +1,2 @@
+future::availableCores()
+

### 2025-03-08 scripts/fieldsize.R

@@ -1,0 +1,52 @@
+library(tidyverse)
+library(sf)
+library(terra)
+# remotes::install_github("ClimateEcology/optimEcoServices")
+library(optimEcoServices)
+
+projdir <- "/Users/kevinli/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/maxWFS/"
+datadir <- paste0(projdir,"data/")
+
+hu <- st_read(paste0(datadir, "hu12watersheds.gpkg"))
+
+hu.names <- hu$huc12
+
+# identify medians
+
+hu.fields <- list()
+for(i in seq_along(hu.names)){
+  hu.name <- hu.names[i]
+
+  hudir <- paste0(projdir, "hu/hu",hu.name,"/")
+  hu.field <- rast(paste0(hudir, "data/lu/fields", ".tif"))
+  hu.fields[[hu.name]] <- as.data.frame(table(terra::values(hu.field$id, na.rm=TRUE)))
+}
+
+hu.fields.range <- bind_rows(hu.fields, .id = "hu") %>% group_by(hu) %>%
+  summarize(n = n(),
+            mean = mean(Freq),
+            sd = sd(Freq),
+            min = min(Freq),
+            quant25 = quantile(Freq,0.25),
+            median = median(Freq),
+            quant75 = quantile(Freq,0.75),
+            max = max(Freq))
+
+median(hu.fields.range$median)
+
+# select field IDs within a range
+hu.fields.size <- bind_rows(hu.fields, .id = "hu") %>%
+  filter(Freq >= 30 & Freq <= 50) %>%
+  group_by(hu) %>%
+  summarise(
+    n.40 = n(),
+    mean.40 = mean(Freq),
+    median.40 = median(Freq)
+  ) %>%
+  left_join(hu.fields.range)
+
+hu.fields.select <- bind_rows(hu.fields, .id = "hu") %>%
+  filter(Freq >= 30 & Freq <= 50) %>% rename(id = Var1, size = Freq) %>%
+  mutate(id = as.numeric(as.character(id)))
+
+saveRDS(hu.fields.select, paste0(datadir,"selectfields.rds"))

### 2025-03-08 scripts/getErosivityData.R

@@ -1,0 +1,24 @@
+library(terra)
+library(sf)
+library(tidyverse)
+
+hu <- sf::st_read("/Users/kevinli/Box Sync/InVEST/GIS_data/NEMWregion/selected_hu12_v2_candidates.gpkg")
+
+global.eros <- rast("/Users/kevinli/Box Sync/InVEST/GIS_data/Erosivity/GlobalR/GlobalR_NoPol.tif")
+
+outfolder <- "/Users/kevinli/Box Sync/InVEST/GIS_data/Pareto_proj/hu_437/erosivity/"
+
+if(!dir.exists(outfolder)){dir.create(outfolder)}
+
+hu.buf <- st_buffer(hu, 1000)
+
+for(i in 1:nrow(hu)){
+  hu.i <- hu.buf[i,] %>% sf::st_transform(crs(global.eros))
+  hu.name <- hu.i$huc12
+
+  ext.hu.i <- terra::ext(hu.i)*1.01
+
+  hu.i.filepath <- paste0(outfolder,"eros_hu",hu.name,".tif")
+
+  terra::crop(global.eros, ext.hu.i, snap="out", filename = hu.i.filepath)
+}

### 2025-03-08 scripts/makeCleanInputData.R

@@ -1,0 +1,228 @@
+# clean up file formats of input data
+# reproject and make the same resolution
+# mask all data by the watershed boundaries
+# note about cdl: I am worried that there may be a boundary effect if the
+# landcover layer is cut off right at the watershed boundary
+
+library(terra)
+library(sf)
+library(tidyverse)
+
+hu <- sf::st_read("/Users/kevinli/Box Sync/InVEST/GIS_data/NEMWregion/selected_hu12_v2_candidates.gpkg")
+hu.buf <- st_buffer(hu, 1000)
+
+raw.filepath <- "/Users/kevinli/Box Sync/InVEST/GIS_data/Pareto_proj/hu_437/"
+out.filepath <- "/Users/kevinli/Box Sync/InVEST/GIS_data/Pareto_proj/hu_437_out/"
+
+if(!dir.exists(out.filepath)){dir.create(out.filepath)}
+
+# make cdl folder if it doesn't exist
+cdl.out <- paste0(out.filepath, "cdl/")
+if(!dir.exists(cdl.out)){dir.create(cdl.out)}
+
+for(i in 1:nrow(hu)){
+  hu.i <- hu.buf[i,]
+  hu.name <- hu.i$huc12
+
+  # make cdl mask
+  cdl.i <- rast(paste0(raw.filepath,"cdl/cdl_hu",hu.name,".tif"))
+  if(!same.crs(cdl.i, hu.i)){stop()}  # should be same crs. This is to stop the process if not, for double checking
+
+  terra::mask(cdl.i, vect(hu.i),
+              filename=paste0(cdl.out, "cdl_hu", hu.name, ".tif"),
+              datatype="INT1U",
+              NAflag = 255, overwrite=TRUE)
+}
+
+# test no data values
+for(i in 1:nrow(hu)){
+  hu.i <- hu.buf[i,]
+  hu.name <- hu.i$huc12
+
+  test <- grep(
+    "NoData Value",
+    terra::describe(paste0(cdl.out, "cdl_hu", hu.name, ".tif")),
+    value = TRUE
+  )
+  print(test)
+}
+
+##### rest of data #####
+
+# huc names
+hu.names <- hu.buf$huc12
+
+# erodibility
+erod.out <- paste0(out.filepath, "erodibility2/")
+if(!dir.exists(erod.out)){dir.create(erod.out)}
+
+for(i in 1:length(hu.names)){
+  hu.name <- hu.names[i]
+
[…]

### 2025-03-08 scripts/onedfield_combine.R

@@ -1,0 +1,10 @@
+library(tidyverse)
+
+hu109 <- readRDS("scripts/hu109.rds")
+hu218 <- readRDS("scripts/hu218.rds")
+hu327 <- readRDS("scripts/hu327.rds")
+hu437 <- readRDS("scripts/hu437.rds")
+
+hu.all <- bind_rows(hu109, hu218, hu327, hu437)
+
+saveRDS(hu.all, "scripts/hu_onefield.rds")

### 2025-03-08 scripts/sbatchlandcomplex.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+
+#SBATCH --partition=atlas
+#SBATCH --qos=normal
+#SBATCH --job-name=landopt
+#SBATCH --output=/90daydata/geoecoservices/optimization/landcomplex-%j.out
+#SBATCH --error=/90daydata/geoecoservices/optimization/landcomplex-%j.err
+#SBATCH --account=geoecoservices
+#SBATCH --mail-user=likevin@psu.edu
+#SBATCH --mail-type=NONE
+#SBATCH --time=24:00:00
+#SBATCH --nodes=1
+#SBATCH --ntasks=48
+
+module purge
+
+module load miniconda3
+source activate
+conda activate r-invest2
+
+srun Rscript --vanilla 

### 2025-03-08 scripts/test.R

@@ -1,0 +1,74 @@
+library(devtools)
+library(tidyverse)
+library(sf)
+library(terra)
+library(optimEcoServices)
+
+projdir <- "/Users/kevinli/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/maxWFS/"
+datadir <- paste0(projdir,"data/")
+
+hu <- st_read(paste0(datadir, "hu12watersheds.gpkg"))
+
+hu.names <- hu$huc12
+table(hu$huc2)
+
+
+baseline <- list()
+maxwfs <- list()
+
+# make biophys directory
+if(!dir.exists("biophys")){
+  dir.create("biophys")
+}
+
+# make polliantor biophys tables
+write.csv(biophys_pol, "biophys/biophys_pol.csv", row.names = FALSE)
+write.csv(guild_table, "biophys/pol_guilds.csv", row.names = FALSE)
+
+# make seasonal water yield biophys tables
+write.csv(biophys_swy, "biophys/biophys_swy.csv", row.names = FALSE)
+
+# make sediment delivery ratio biophys tables
+write.csv(biophys_sdr$Borrelli, "biophys/biophys_sdr.csv", row.names = FALSE)
+
+# make nutrient delivery ratio biophys tables
+write.csv(biophys_ndr$BenezSecanho, "biophys/biophys_ndr.csv", row.names = FALSE)
+
+for(i in seq_along(hu.names)){
+  hu.name <- hu.names[i]
+
+  hudir <- paste0(projdir, "hu/hu",hu.name,"/")
+  hu.fields <- rast(paste0(hudir, "data/lu/fields", ".tif"))
+  hu.margins <- rast(paste0(hudir, "data/lu/margins", ".tif"))
+
+  hu.marginids <- unique(terra::values(hu.margins$id, na.rm=TRUE)) # added unique() because ids are section-based, not cell based
+  basevec <- rep(0, length(hu.marginids))
+  maxvec <- rep(1, length(hu.marginids))
+
+  # baseline
+  baseline[[i]] <- wfs_fitness(
+    binary.vector = basevec,
+    margins.rast = hu.margins,
+    fields.rast = hu.fields,
+    margin.lc = 171,   # land cover for 'pollinator habitat'
+    biophys.dir = "biophys",
+    data.dir =  paste0(hudir, "data"),
+    workspace.dir = paste0(hudir,"baseline"),
+    suffix="base"
[…]

### 2025-03-08 scripts/sbatchlandcomplex.sh
script target

@@ -18,4 +18,4 @@ module load miniconda3
 source activate
 conda activate r-invest2
 
-srun Rscript --vanilla 
+srun Rscript --vanilla /project/geoecoservices/optimization/landscape-complexity/scripts/atlas_max_one.R

### 2025-03-08 scripts/atlas_max_one.R
now try array

@@ -1,9 +1,10 @@
-library(devtools)
-library(tidyverse)
-library(sf)
-library(terra)
-library(doFuture)
-library(optimEcoServices)
+arryindx <- as.numeric(commandArgs(trailingOnly = TRUE))
+
+suppressMessages(library(devtools))
+suppressMessages(library(tidyverse))
+suppressMessages(library(sf))
+suppressMessages(library(terra))
+suppressMessages(library(optimEcoServices))
 
 # projdir <- "/Users/kevinli/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/maxWFS/"
 projdir <- "/90daydata/geoecoservices/optimization/maxWFS/"
@@ -20,11 +21,18 @@ hu <- st_read(paste0(datadir, "hu12watersheds.gpkg"))
 
 hu.names <- hu$huc12
 
-# set up futures
-plan(multisession)
+batchnum <- (arryindx-1)*23+seq(1,23)
 
-out <- foreach(i = seq_along(hu.names)) %dofuture% {
+baseline <- list()
+maxwfs <- list()
+onefield <- list()
+
+for(i in seq_along(batchnum)){
+  i.start <- Sys.time()
+  
   hu.name <- hu.names[i]
+  
+  cat(paste(hu.name, "started on", lubridate::as_datetime(i.start), "\n"))
 
   hudir <- paste0(projdir, "hu/hu",hu.name,"/")
   hu.fields <- rast(paste0(hudir, "data/lu/fields", ".tif"))
@@ -35,7 +43,8 @@ out <- foreach(i = seq_along(hu.names)) %dofuture% {
   maxmargin <- terra::ifel(test = !is.na(hu.margins$id), 1, 0)
 
   # baseline
-  baseline <- wfs_fitness2(
+  base.start <- Sys.time()
+  baseline[[hu.name]] <- wfs_fitness2(
     margin.rast = basemargin,
     fields.rast = hu.fields,
     margin.lc = 171,   # land cover for 'pollinator habitat'
@@ -44,9 +53,17 @@ out <- foreach(i = seq_along(hu.names)) %dofuture% {
     workspace.dir = paste0(hudir,"baseline"),
     suffix="base"
   )
+  cat(
+    paste(
+      "baseline took",
+      round(as.numeric(difftime(Sys.time(), base.start, units="secs")), 0),
+      "seconds\n"
+    )
+    )
 
   # max wfs
[…]

### 2025-03-08 scripts/sbatchlandcomplex.sh

@@ -10,7 +10,10 @@
 #SBATCH --mail-type=NONE
 #SBATCH --time=24:00:00
 #SBATCH --nodes=1
-#SBATCH --ntasks=48
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=2
+#SBATCH --get-user-env              ### Import your user environment setup
+#SBATCH --array=1-19                ### Array index
 
 module purge
 
@@ -18,4 +21,4 @@ module load miniconda3
 source activate
 conda activate r-invest2
 
-srun Rscript --vanilla /project/geoecoservices/optimization/landscape-complexity/scripts/atlas_max_one.R
+srun Rscript --vanilla /project/geoecoservices/optimization/landscape-complexity/scripts/atlas_max_one.R ${SLURM_ARRAY_TASK_ID}

### 2025-03-08 scripts/atlas_test.R
atlas test

@@ -1,2 +1,122 @@
-future::availableCores()
+suppressMessages(library(devtools))
+suppressMessages(library(tidyverse))
+suppressMessages(library(sf))
+suppressMessages(library(terra))
+suppressMessages(library(optimEcoServices))
 
+# projdir <- "/Users/kevinli/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/maxWFS/"
+projdir <- "/90daydata/geoecoservices/optimization/maxWFS/"
+
+datadir <- paste0(projdir,"data/")
+
+# biophys directory
+biophysdir <- paste0(datadir, "biophys")
+
+# onefields
+fields.select <- readRDS(paste0(datadir,"selectfields.rds"))
+
+hu <- st_read(paste0(datadir, "hu12watersheds.gpkg"))
+
+hu.names <- hu$huc12
+
+baseline <- list()
+maxwfs <- list()
+onefield <- list()
+
+for(i in 1){
+  i.start <- Sys.time()
+  
+  hu.name <- hu.names[i]
+  
+  cat(paste(hu.name, "started on", lubridate::as_datetime(i.start), "\n"))
+
+  hudir <- paste0(projdir, "hu/hu",hu.name,"/")
+  hu.fields <- rast(paste0(hudir, "data/lu/fields", ".tif"))
+  hu.margins <- rast(paste0(hudir, "data/lu/margins", ".tif"))
+
+  hu.marginids <- unique(terra::values(hu.margins$id, na.rm=TRUE)) # added unique() because ids are section-based, not cell based
+  basemargin <- terra::ifel(test = !is.na(hu.margins$id), 0, 0)
+  maxmargin <- terra::ifel(test = !is.na(hu.margins$id), 1, 0)
+
+  # baseline
+  base.start <- Sys.time()
+  baseline[[hu.name]] <- wfs_fitness2(
+    margin.rast = basemargin,
+    fields.rast = hu.fields,
+    margin.lc = 171,   # land cover for 'pollinator habitat'
+    biophys.dir = biophysdir,
+    data.dir =  paste0(hudir, "data"),
+    workspace.dir = paste0(hudir,"baseline"),
+    suffix="base"
+  )
+  cat(
+    paste(
+      "baseline took",
+      round(as.numeric(difftime(Sys.time(), base.start, units="secs")), 0),
+      "seconds\n"
+    )
+    )
+
+  # max wfs
+  maxwfs.start <- Sys.time()
[…]

### 2025-03-08 scripts/sbatchlandcomplex.sh
batch

@@ -11,7 +11,7 @@
 #SBATCH --time=24:00:00
 #SBATCH --nodes=1
 #SBATCH --ntasks=1
-#SBATCH --cpus-per-task=2
+#SBATCH --cpus-per-task=4
 #SBATCH --get-user-env              ### Import your user environment setup
 #SBATCH --array=1-19                ### Array index
 
@@ -21,4 +21,4 @@ module load miniconda3
 source activate
 conda activate r-invest2
 
-srun Rscript --vanilla /project/geoecoservices/optimization/landscape-complexity/scripts/atlas_max_one.R ${SLURM_ARRAY_TASK_ID}
+srun Rscript --vanilla /project/geoecoservices/optimization/landscape-complexity/scripts/atlas_max_one.R $SLURM_ARRAY_TASK_ID

### 2025-03-08 scripts/atlas_max_one.R
for loop index fix

@@ -27,7 +27,7 @@ baseline <- list()
 maxwfs <- list()
 onefield <- list()
 
-for(i in seq_along(batchnum)){
+for(i in batchnum){
   i.start <- Sys.time()
   
   hu.name <- hu.names[i]

### 2025-03-11 docs/RandomForest/randomforest.Qmd
new results

---
title: "ES Models: Random Forest"
author: "Sarah Goslee"
date: "2025-03-06"
format: 
  html:
    code-fold: true
fig-height: 6
fig-width: 10
---

Per Kevin (2025-02-26), everything is already normalized by area, so dividing by density is correct.

Ecosystem services, per ha:

	Pollinators
	N export
	Sediment export
	Baseflow

Change in ecosystem services (delta) per ha of watershed: delta

Change in ecosystem services per ha per ha of margin: delta.margin

```{r setup}

    library(ranger)
	library(pdp)
    library(gridExtra)

###

    wfs <- readRDS("landscape-complexity-main/data/wfs_lsm.rds")

    dat.predictors <-  wfs[, 20:50]

    # drop log-transform
    dat.predictors <- dat.predictors[, !grepl("log", colnames(dat.predictors))]

### previously-completed analyses

    if(file.exists("ES-RF.RDA")) load("ES-RF.RDA")

```

```{r screening}

#| label: fig-overview
#| fig-cap: "Basic ES responses."
#| warning: false

    par(mfrow=c(1, 3))

    # with(wfs, plot(ann_agr_pland, pollinators_base, col = "black", xlab = "Annual agriculture (%)", ylab = "Ecosystem Service", main = "Pollination"))
    # with(wfs, points(ann_agr_pland, pollinators_max, col = "green"))
    # legend("topright", pch = c(1, 1), col = c("black", "green"), legend = c("Base", "Max"))

    with(wfs, plot(pollinators_base, pollinators_max, col = "darkblue", xlab = "Base value", ylab = "Max", main = "Pollination"))
    abline(c(0, 1), lty = 2)

    with(wfs, plot(pollinators_base, delta_pollinators.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Pollination"))

    with(wfs, plot(ann_agr_pland, delta_pollinators.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Pollination"))

    #

    # with(wfs, plot(ann_agr_pland, n_export_base, col = "black", xlab = "Annual agriculture (%)", ylab = "Ecosystem Service", main = "Nitrogen"))
    # with(wfs, points(ann_agr_pland, n_export_max, col = "purple"))

    with(wfs, plot(n_export_base, n_export_max, col = "darkblue", xlab = "Base value", ylab = "Max", main = "Nitrogen"))
    abline(c(0, 1), lty = 2)

    with(wfs, plot(n_export_base, delta_n_export.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Nitrogen"))

    with(wfs, plot(ann_agr_pland, delta_n_export.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Nitrogen"))

    #

    # with(wfs, plot(ann_agr_pland, sed_export_base, col = "black", xlab = "Annual agriculture (%)", ylab = "Ecosystem Service", main = "Sediment"))
    # with(wfs, points(ann_agr_pland, sed_export_max, col = "purple"))

    with(wfs, plot(sed_export_base, sed_export_max, col = "darkblue", xlab = "Base value", ylab = "Max", main = "Sediment"))
    abline(c(0, 1), lty = 2)

    with(wfs, plot(sed_export_base, delta_sed.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Sediment"))

    with(wfs, plot(ann_agr_pland, delta_sed.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Sediment"))

    #

    # with(wfs, plot(ann_agr_pland, qb_base, col = "black", xlab = "Annual agriculture (%)", ylab = "Ecosystem Service", main = "Baseflow"))
    # with(wfs, points(ann_agr_pland, qb_max, col = "purple"))

    with(wfs, plot(qb_base, qb_max, col = "darkblue", xlab = "Base value", ylab = "Max", main = "Baseflow"))
    abline(c(0, 1), lty = 2)

    with(wfs, plot(qb_base, delta_qb.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Baseflow"))

    with(wfs, plot(ann_agr_pland, delta_qb.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Baseflow"))

```

What factors correlate to the base rate of ecosystem services?

What factors correlate to the per-ha effect of adding pollinator strips?

## Random Forest

These are basic regression RF models looking at the base ES value and at the delta-margin value.

```{r rfstart}

    if(!exists("pollinators_base.rf")) {

        dat <- data.frame(y = wfs$pollinators_base, dat.predictors)
        pollinators_base.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
        pollinators_base.p <- data.frame(importance_pvalues(pollinators_base.rf, method = "altmann", formula = y ~ ., data = dat, num.permutations = 1000))
        rm(dat)

        dat <- data.frame(y = wfs$delta_pollinators.margin, dat.predictors)
        pollinators_margin.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
        pollinators_margin.p <- data.frame(importance_pvalues(pollinators_margin.rf, method = "altmann", formula = y ~ ., data = dat, num.permutations = 1000))
        rm(dat)

    }

	####

    if(!exists("n_export_base.rf")) {

        dat <- data.frame(y = wfs$n_export_base, dat.predictors)
        n_export_base.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
        n_export_base.p <- data.frame(importance_pvalues(n_export_base.rf, method = "altmann", formula = y ~ ., data = dat, num.permutations = 1000))
        rm(dat)

        dat <- data.frame(y = wfs$delta_n_export.margin, dat.predictors)
        n_export_margin.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
        n_export_margin.p <- data.frame(importance_pvalues(n_export_margin.rf, method = "altmann", formula = y ~ ., data = dat, num.permutations = 1000))
        rm(dat)

    }

	####

    if(!exists("sed_export_base.rf")) {

        dat <- data.frame(y = wfs$sed_export_base, dat.predictors)
        sed_export_base.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
        sed_export_base.p <- data.frame(importance_pvalues(sed_export_base.rf, method = "altmann", formula = y ~ ., data = dat, num.permutations = 1000))
        rm(dat)

        dat <- data.frame(y = wfs$delta_sed.margin, dat.predictors)
        sed_export_margin.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
        sed_export_margin.p <- data.frame(importance_pvalues(sed_export_margin.rf, method = "altmann", formula = y ~ ., data = dat, num.permutations = 1000))
        rm(dat)

    }

	####

    if(!exists("qb_base.rf")) {

        dat <- data.frame(y = wfs$qb_base, dat.predictors)
        qb_base.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
        qb_base.p <- data.frame(importance_pvalues(qb_base.rf, method = "altmann", formula = y ~ ., data = dat, num.permutations = 1000))
        rm(dat)

        dat <- data.frame(y = wfs$delta_qb.margin, dat.predictors)
        qb_margin.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
        qb_margin.p <- data.frame(importance_pvalues(qb_margin.rf, method = "altmann", formula = y ~ ., data = dat, num.permutations = 1000))
        rm(dat)

    }

```

### Model performance

These are regression RF, so model performance can be assessed by looking at r2 and at MSE. Note that these are on the training data, not on test data, so not a true validation assessment that would be needed for a predictive model. Since we're doing explanatory modeling currently, that's fine.

*Pollinators:*

   - Base

     r2 = `{r} round(pollinators_base.rf$r.squared, 3)`
     OOB MSE = `{r} round(pollinators_base.rf$prediction.error, 3)`

  - Margin

     r2 = `{r} round(pollinators_margin.rf$r.squared, 3)`
     OOB MSE = `{r} round(pollinators_margin.rf$prediction.error, 3)`

*Nitrogen export:*

   - Base

     r2 = `{r} round(n_export_base.rf$r.squared, 3)`
     OOB MSE = `{r} round(n_export_base.rf$prediction.error, 3)`

  - Margin

     r2 = `{r} round(n_export_margin.rf$r.squared, 3)`
     OOB MSE = `{r} round(n_export_margin.rf$prediction.error, 3)`

*Sediment export:*

   - Base

     r2 = `{r} round(sed_export_base.rf$r.squared, 3)`
     OOB MSE = `{r} round(sed_export_base.rf$prediction.error, 3)`

  - Margin

     r2 = `{r} round(sed_export_margin.rf$r.squared, 3)`
     OOB MSE = `{r} round(sed_export_margin.rf$prediction.error, 3)`

*Baseflow:*

   - Base

     r2 = `{r} round(qb_base.rf$r.squared, 3)`
     OOB MSE = `{r} round(qb_base.rf$prediction.error, 3)`

  - Margin

     r2 = `{r} round(qb_margin.rf$r.squared, 3)`
     OOB MSE = `{r} round(qb_margin.rf$prediction.error, 3)`

Can predict pollination extremely well, N quite well, and sediment and baseflow moderately well.

### Variable importance

I used permutation importance, which is an unbiased measure (there's a huge literature on RF importance metrics), and p-values can be calculated. Blue bars denote p < 0.05.

```{r importance}

#| label: fig-importance
#| fig-cap: "RF importance values."
#| warning: false

    par(mfrow = c(1, 1))
    par(mai = c(2, .8, .8, .4))

    dat <- pollinators_base.p[order(pollinators_base.p$importance), ]
    barplot(dat$importance, col = ifelse(dat$pvalue > .05, "gray", "darkblue"), las = 2, main = "Pollination: Base", names = rownames(dat))
    dat <- pollinators_margin.p[order(pollinators_margin.p$importance), ]
    barplot(dat$importance, col = ifelse(dat$pvalue > .05, "gray", "darkblue"), las = 2, main = "Pollination: Margin", names = rownames(dat))
    rm(dat)

    dat <- n_export_base.p[order(n_export_base.p$importance), ]
    barplot(dat$importance, col = ifelse(dat$pvalue > .05, "gray", "darkblue"), las = 2, main = "Nitrogen: Base", names = rownames(dat))
    dat <- n_export_margin.p[order(n_export_margin.p$importance), ]
    barplot(dat$importance, col = ifelse(dat$pvalue > .05, "gray", "darkblue"), las = 2, main = "Nitrogen: Margin", names = rownames(dat))
    rm(dat)

    dat <- sed_export_base.p[order(sed_export_base.p$importance), ]
    barplot(dat$importance, col = ifelse(dat$pvalue > .05, "gray", "darkblue"), las = 2, main = "Sediment: Base", names = rownames(dat))
    dat <- sed_export_margin.p[order(sed_export_margin.p$importance), ]
    barplot(dat$importance, col = ifelse(dat$pvalue > .05, "gray", "darkblue"), las = 2, main = "Sediment: Margin", names = rownames(dat))
    rm(dat)

    dat <- qb_base.p[order(qb_base.p$importance), ]
    barplot(dat$importance, col = ifelse(dat$pvalue > .05, "gray", "darkblue"), las = 2, main = "Baseflow: Base", names = rownames(dat))
    dat <- qb_margin.p[order(qb_margin.p$importance), ]
    barplot(dat$importance, col = ifelse(dat$pvalue > .05, "gray", "darkblue"), las = 2, main = "Baseflow: Margin", names = rownames(dat))
    rm(dat)

```

I want to look more closely at the response over various metrics.

```{r pdp}

    if(!exists("pollinators_base.ann_agr_pland.pdp")) {
        dat <- data.frame(y = wfs$delta_pollinators.margin, orig = wfs$pollinators_base, dat.predictors)
        pollinators_base.ann_agr_pland.pdp <- partial(pollinators_base.rf, train = dat, pred.var = "ann_agr_pland")
        pollinators_margin.ann_agr_pland.pdp <- partial(pollinators_margin.rf, train = dat, pred.var = "ann_agr_pland")
        pollinators_base.ntr_pland.pdp <- partial(pollinators_base.rf, train = dat, pred.var = "ntr_pland")
        pollinators_margin.ntr_pland.pdp <- partial(pollinators_margin.rf, train = dat, pred.var = "ntr_pland")
        rm(dat)
    }

    if(!exists("n_export_base.ann_agr_pland.pdp")) {
        dat <- data.frame(y = wfs$delta_n_export.margin, orig = wfs$n_export_base, dat.predictors)
        n_export_base.ann_agr_pland.pdp <- partial(n_export_base.rf, train = dat, pred.var = "ann_agr_pland")
        n_export_margin.ann_agr_pland.pdp <- partial(n_export_margin.rf, train = dat, pred.var = "ann_agr_pland")
        n_export_base.ntr_pland.pdp <- partial(n_export_base.rf, train = dat, pred.var = "ntr_pland")
        n_export_margin.ntr_pland.pdp <- partial(n_export_margin.rf, train = dat, pred.var = "ntr_pland")
        rm(dat)
    }

    if(!exists("sed_export_base.ann_agr_pland.pdp")) {
        dat <- data.frame(y = wfs$delta_sed.margin, orig = wfs$sed_export_base, dat.predictors)
        sed_export_base.ann_agr_pland.pdp <- partial(sed_export_base.rf, train = dat, pred.var = "ann_agr_pland")
        sed_export_margin.ann_agr_pland.pdp <- partial(sed_export_margin.rf, train = dat, pred.var = "ann_agr_pland")
        sed_export_base.ntr_pland.pdp <- partial(sed_export_base.rf, train = dat, pred.var = "ntr_pland")
        sed_export_margin.ntr_pland.pdp <- partial(sed_export_margin.rf, train = dat, pred.var = "ntr_pland")
        rm(dat)
    }

    if(!exists("qb_base.ann_agr_pland.pdp")) {
        dat <- data.frame(y = wfs$delta_qb.margin, orig = wfs$qb_base, dat.predictors)
        qb_base.ann_agr_pland.pdp <- partial(qb_base.rf, train = dat, pred.var = "ann_agr_pland")
        qb_margin.ann_agr_pland.pdp <- partial(qb_margin.rf, train = dat, pred.var = "ann_agr_pland")
        qb_base.ntr_pland.pdp <- partial(qb_base.rf, train = dat, pred.var = "ntr_pland")
        qb_margin.ntr_pland.pdp <- partial(qb_margin.rf, train = dat, pred.var = "ntr_pland")
        rm(dat)
    }

    save(list = ls(pattern = "^[npqs]"), file = "ES-RF.RDA")

```

Partial dependence plots show change in the Y variable across the range of a specific predictor (or more than one), when all other variables are held constant.

```{r partialplot1}

    p1 <- plotPartial(pollinators_base.ann_agr_pland.pdp, main = "Pollinators base")
    p2 <- plotPartial(pollinators_margin.ann_agr_pland.pdp, main = "Pollinators margin")
    p3 <- plotPartial(pollinators_base.ntr_pland.pdp, main = "Pollinators base")
    p4 <- plotPartial(pollinators_margin.ntr_pland.pdp, main = "Pollinators margin")
    grid.arrange(p1, p3, p2, p4, ncol = 2)

```

```{r partialplot2}

    p1 <- plotPartial(n_export_base.ann_agr_pland.pdp, main = "Nitrogen base")
    p2 <- plotPartial(n_export_margin.ann_agr_pland.pdp, main = "Nitrogen margin")
    p3 <- plotPartial(n_export_base.ntr_pland.pdp, main = "Nitrogen base")
    p4 <- plotPartial(n_export_margin.ntr_pland.pdp, main = "Nitrogen margin")
    grid.arrange(p1, p3, p2, p4, ncol = 2)

```

```{r partialplot3}

    p1 <- plotPartial(sed_export_base.ann_agr_pland.pdp, main = "Sediment base")
    p2 <- plotPartial(sed_export_margin.ann_agr_pland.pdp, main = "Sediment margin")
    p3 <- plotPartial(sed_export_base.ntr_pland.pdp, main = "Sediment base")
    p4 <- plotPartial(sed_export_margin.ntr_pland.pdp, main = "Sediment margin")
    grid.arrange(p1, p3, p2, p4, ncol = 2)

```

```{r partialplot4}

    p1 <- plotPartial(qb_base.ann_agr_pland.pdp, main = "Baseflow base")
    p2 <- plotPartial(qb_margin.ann_agr_pland.pdp, main = "Baseflow margin")
    p3 <- plotPartial(qb_base.ntr_pland.pdp, main = "Baseflow base")
    p4 <- plotPartial(qb_margin.ntr_pland.pdp, main = "Baseflow margin")
    grid.arrange(p1, p3, p2, p4, ncol = 2)

```

### 2025-03-11 docs/old/landscapemetrics_quantile.Rmd
new results

--- 2025-02-19 landscapemetrics_quantile.Rmd
@@
 
-ggbiplot::ggbiplot(wfs_pca_all, groups = wfs_grp$habcat)
+ggbiplot::ggbiplot(wfs_pca_all, groups = wfs_grp$habcat) +
+  egg::theme_article()
 
@@
 
-ggbiplot::ggbiplot(wfs_pca_all, groups = wfs_grp$habcat, choices = c(2,3))
+ggbiplot::ggbiplot(wfs_pca_all, groups = wfs_grp$habcat, choices = c(2,3)) +
+  egg::theme_article()
 
@@
 lsm_pc <- wfs_grp %>% cbind(predict(wfs_pca_all))
+saveRDS(lsm_pc, "data/wfs_lsm_pc.rds")
 

### 2025-03-11 examine_sim_results.Rmd
new results

---
title: "examine simulation results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##

```{r load, include = FALSE}
library(tidyverse)
# wfs_lsm <- readRDS("data/wfs_lsm.rds")
# hu_meta <- readRDS("data/hu_meta.rds")
# 
# hulandstats <- readRDS("data/hulandstats.rds")
# 
# 
# wfs_meta <- wfs_lsm %>% 
#   left_join(hu_meta %>% sf::st_drop_geometry(), 
#             by = c("hu" = "huc12", "huc2", "name")) %>%
#   left_join(hulandstats)

wfs_meta <- readRDS("data/allresults_20250308.rds")

```

```{r visualize_base}

base_df <- wfs_meta %>% 
  select(c(hu, ntr_pland, pollinators_base:sed_export_base)) %>%
  pivot_longer(cols = pollinators_base:sed_export_base,
               names_to = "Ecosystem_service", values_to = "Ecosystem_value")

ggplot(base_df, aes(ntr_pland, Ecosystem_value)) +
  geom_point() +
  facet_wrap(~Ecosystem_service, scales = "free_y") + egg::theme_article()

```

```{r visualize_delta}

delta_df <- wfs_meta %>% 
  select(c(hu, ntr_pland, delta_pollinators:delta_sed)) %>%
  pivot_longer(cols = delta_pollinators:delta_sed,
               names_to = "Ecosystem_service", values_to = "ES_value_change")

ggplot(delta_df, aes(ntr_pland, ES_value_change)) +
  geom_point() +
  facet_wrap(~Ecosystem_service, scales = "free_y") + egg::theme_article()

```

```{r visualize_delta_margin}

delta_df <- wfs_meta %>% 
  select(c(hu, ntr_pland, delta_pollinators_max.margin:delta_sed_max.margin)) %>%
  pivot_longer(cols = delta_pollinators_max.margin:delta_sed_max.margin,
               names_to = "Ecosystem_service", values_to = "ES_value_change_ha_margin")

ggplot(delta_df, aes(ntr_pland, ES_value_change_ha_margin)) +
  geom_point() +
  facet_wrap(~Ecosystem_service, scales = "free_y") + egg::theme_article()

```

```{r visualize_delta_margin_onefield}

delta_df <- wfs_meta %>% 
  select(c(hu, ntr_pland,
           delta_pollinators_of.margin.median, 
           delta_n_export_of.margin.median,
           delta_qb_of.margin.median,
           delta_sed_of.margin.median)) %>%
  pivot_longer(cols = delta_pollinators_of.margin.median:delta_sed_of.margin.median,
               names_to = "Ecosystem_service", 
               values_to = "One_field_change_ha_margin")

ggplot(delta_df, aes(ntr_pland, One_field_change_ha_margin)) +
  geom_point() +
  facet_wrap(~Ecosystem_service, scales = "free_y") + egg::theme_article()

```

## Plots

```{r wetland}
ggplot(wfs_meta, aes(Wetland, delta_sed_max.margin)) + 
  geom_point()
```

```{r erosivity}
ggplot(wfs_meta, aes(erosivity_mean, delta_sed_max.margin)) + 
  geom_point()
```

```{r erodibility}
ggplot(wfs_meta, aes(erodibility_mean, delta_sed_max.margin)) + 
  geom_point()
```

```{r erodibility_2}
library(plotly)
fig <- plot_ly(wfs_meta, x = ~erodibility_mean, y = ~erosivity_mean, z = ~delta_sed_max.margin, color = ~delta_sed_max.margin, text = ~hu)

fig

```

```{r sediment_change}
ggplot(wfs_meta, aes(sed_export_base, sed_export_max, color = delta_sed_max.margin)) + 
  geom_point() + geom_abline(intercept = 0, slope = 1) + scale_color_continuous(type="viridis")

```

```{r nutrient_change}
ggplot(wfs_meta, aes(n_export_base, n_export_max, color = delta_n_export_max.margin)) + 
  geom_point() + geom_abline(intercept = 0, slope = 1) + scale_color_continuous(type="viridis")

```

```{r pollinator_change}
ggplot(wfs_meta, aes(pollinators_base, pollinators_max, color = delta_pollinators_max.margin)) + 
  geom_point() + geom_abline(intercept = 0, slope = 1) + scale_color_continuous(type="viridis")

```

```{r qb_change}
ggplot(wfs_meta, aes(qb_base, qb_max, color = delta_qb_max.margin)) + 
  geom_point() + geom_abline(intercept = 0, slope = 1) + scale_color_continuous(type="viridis")

```

### 2025-03-11 old/landscapemodels.Rmd
new results

---
title: "Landscape models"
---

```{r setup, echo = FALSE}
library(tidyverse)
library(performance)
library(psych)
source("R/utils-plot.r")

wfs <- readRDS("data/maxwfsresults_20241217_fix.rds")
lsm <- read.csv("data/lsm_reduced.csv") %>% 
  mutate(hu = gsub("hu", "", hu))

wfs_lsm <- wfs %>% left_join(lsm, by = "hu") %>%
  select(c(hu, huc2, name, margin.ha, lu_ta,
           pollinators_base:sed_export_base, 
           pollinators_max:sed_export_max, 
           delta_pollinators:delta_sed, 
           fields_max, field.ha, margin.ha, 
           lu_shdi:ntr_frac_mn)) %>%
  mutate(prn_ann_ratio = log(prn_agr_pland/ann_agr_pland),
         prn_pagr = 100*(prn_agr_pland/(ann_agr_pland+prn_agr_pland)),
         field_pland = field.ha/lu_ta,
         margin_dens = margin.ha/lu_ta)

scaled.data <- wfs_lsm %>% mutate(across(lu_shdi:field_pland, ~scale(.x)[,1]))

```

## Correlation between independent variables.

I selected a subset of variables that aren't strongly correlated and that represent potential drivers of ecosystem service patterns.

```{r correlation, echo = FALSE}

pairs.panels(wfs_lsm %>% 
               select(ann_agr_pland, prn_agr_pland, aglu_iji,
                      lu_shdi, aglu_frac_mn, margin_dens), 
             cex.cor = 1, cex.labels=1, cex=0.5)

```

## Pollinator models

### Pollinator delta

```{r pollinator_delta}

pol.dmod0 <- lm(delta_pollinators ~ margin_dens, data = scaled.data)
pol.dmod1 <- lm(delta_pollinators ~ ann_agr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
pol.dmod2 <- lm(delta_pollinators ~ ann_agr_pland + I(ann_agr_pland^2) + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)

performance::compare_performance(pol.dmod0, pol.dmod1, pol.dmod2)

summary(pol.dmod2)

```

```{r pollinator_plots, echo = FALSE}

pol.vars <- c("margin_dens", "ann_agr_pland", "aglu_frac_mn", "aglu_iji","prn_agr_pland")

pol.preds <- plot_fun(pol.dmod2, pol.vars, scaled.data)

ggplot(pol.preds[[1]], aes(x, delta_pollinators)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = pol.preds[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  xlab("z-score") +
  egg::theme_article()

```

## Seasonal water yield

### Seasonal water yield delta

```{r swy_delta}

swy.dmod0 <- lm(delta_qb ~ margin_dens, data = scaled.data)
swy.dmod1 <- lm(delta_qb ~ ann_agr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
swy.dmod2 <- lm(delta_qb ~ ann_agr_pland + I(ann_agr_pland^2) + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)

performance::compare_performance(swy.dmod0, swy.dmod1, swy.dmod2)

summary(swy.dmod2)

```

```{r swy_plots, echo = FALSE}

swy.vars <- c("margin_dens", "ann_agr_pland", "prn_agr_pland", "aglu_iji", "aglu_frac_mn")

swy.preds <- plot_fun(swy.dmod2, swy.vars, scaled.data)

ggplot(swy.preds[[1]], aes(x, delta_qb)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = swy.preds[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  xlab("z-score") +
  egg::theme_article()

```

## Nutrient delivery ratio

### Nutrient delivery ratio delta

```{r ndr_delta}

ndr.dmod0 <- lm(delta_n_export ~ margin_dens, data = scaled.data)
ndr.dmod1 <- lm(delta_n_export ~ ann_agr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
ndr.dmod2 <- lm(delta_n_export ~ ann_agr_pland + I(ann_agr_pland^2) + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)

performance::compare_performance(ndr.dmod0, ndr.dmod1, ndr.dmod2)

summary(ndr.dmod2)

```

```{r ndr_plots, echo = FALSE}

ndr.vars <- c("margin_dens", "ann_agr_pland", "prn_agr_pland", 
              "aglu_iji", "aglu_frac_mn", "lu_shdi")

ndr.preds <- plot_fun(ndr.dmod2, ndr.vars, scaled.data)

ggplot(ndr.preds[[1]], aes(x, delta_n_export)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = ndr.preds[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  xlab("z-score") +
  egg::theme_article()

```

## Sediment delivery ratio

### Sediment delivery ratio delta

```{r sdr_delta}

sdr.dmod0 <- lm(delta_sed ~ margin_dens, data = scaled.data)
sdr.dmod1 <- lm(delta_sed ~ ann_agr_pland + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)
sdr.dmod2 <- lm(delta_sed ~ ann_agr_pland + I(ann_agr_pland^2) + aglu_iji + prn_agr_pland + aglu_frac_mn + margin_dens + lu_shdi, data = scaled.data)

performance::compare_performance(sdr.dmod0, sdr.dmod1, sdr.dmod2)

summary(sdr.dmod2)

```

```{r sdr_plots, echo = FALSE}

sdr.vars <- c("margin_dens", "ann_agr_pland", "prn_agr_pland", "lu_shdi")

sdr.preds <- plot_fun(sdr.dmod2, sdr.vars, scaled.data)

ggplot(sdr.preds[[1]], aes(x, delta_sed)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = sdr.preds[[2]], 
             mapping = aes(color = margin_dens), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  xlab("z-score") +
  egg::theme_article()

```

## Residuals-based models

Since margin density is a major effect, I regress the other predictors against the residuals of univariate models with only margin density

### Pollinator model residuals

```{r pollinators_resid, echo = FALSE}

pol.resid <- data.frame(pol_resid = pol.dmod0$residuals, 
                        scaled.data %>% 
                          select(ann_agr_pland, prn_agr_pland, aglu_iji, 
                                 aglu_frac_mn, lu_shdi))

pol.resid.mod <- lm(pol_resid ~ ann_agr_pland + I(ann_agr_pland^2) + prn_agr_pland + aglu_iji + aglu_frac_mn + lu_shdi, data = pol.resid)

summary(pol.resid.mod)

pol.resid.vars <- c("ann_agr_pland", "prn_agr_pland", "aglu_frac_mn", "aglu_iji", "lu_shdi")

pol.resid.preds <- plot_fun(pol.resid.mod, pol.resid.vars, pol.resid, "ann_agr_pland")

ggplot(pol.resid.preds[[1]], aes(x, pol_resid)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = pol.resid.preds[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  xlab("z-score") +
  egg::theme_article()

```

### Seasonal water yield residuals model

```{r swy_resid, echo = FALSE}

swy.resid <- data.frame(swy_resid = swy.dmod0$residuals, 
                        scaled.data %>% 
                          select(ann_agr_pland, prn_agr_pland, aglu_iji, 
                                 aglu_frac_mn, lu_shdi))

swy.resid.mod <- lm(swy_resid ~ ann_agr_pland + I(ann_agr_pland^2) + prn_agr_pland + aglu_iji + aglu_frac_mn + lu_shdi, data = swy.resid)

summary(swy.resid.mod)

swy.resid.vars <- c("ann_agr_pland", "prn_agr_pland", "aglu_frac_mn", "aglu_iji", "lu_shdi")

swy.resid.preds <- plot_fun(swy.resid.mod, swy.resid.vars, swy.resid, "ann_agr_pland")

ggplot(swy.resid.preds[[1]], aes(x, swy_resid)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = swy.resid.preds[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  xlab("z-score") +
  egg::theme_article()

```

### Nutrient delivery residuals model

```{r ndr_resid, echo = FALSE}

ndr.resid <- data.frame(ndr_resid = ndr.dmod0$residuals, 
                        scaled.data %>% 
                          select(ann_agr_pland, prn_agr_pland, aglu_iji, 
                                 aglu_frac_mn, lu_shdi))

ndr.resid.mod <- lm(ndr_resid ~ ann_agr_pland + I(ann_agr_pland^2) + prn_agr_pland + aglu_iji + aglu_frac_mn + lu_shdi, data = ndr.resid)

summary(ndr.resid.mod)

ndr.resid.vars <- c("ann_agr_pland", "prn_agr_pland", "aglu_frac_mn", "aglu_iji", "lu_shdi")

ndr.resid.preds <- plot_fun(ndr.resid.mod, ndr.resid.vars, ndr.resid, "ann_agr_pland")

ggplot(ndr.resid.preds[[1]], aes(x, ndr_resid)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = ndr.resid.preds[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  xlab("z-score") +
  egg::theme_article()

```

### Sediment delivery residuals model

```{r sdr_resid, echo = FALSE}

sdr.resid <- data.frame(sdr_resid = sdr.dmod0$residuals, 
                        scaled.data %>% 
                          select(ann_agr_pland, prn_agr_pland, aglu_iji, 
                                 aglu_frac_mn, lu_shdi))

sdr.resid.mod <- lm(sdr_resid ~ ann_agr_pland + I(ann_agr_pland^2) + prn_agr_pland + aglu_iji + aglu_frac_mn + lu_shdi, data = sdr.resid)

summary(sdr.resid.mod)

sdr.resid.vars <- c("ann_agr_pland", "prn_agr_pland", "aglu_frac_mn", "aglu_iji", "lu_shdi")

sdr.resid.preds <- plot_fun(sdr.resid.mod, sdr.resid.vars, sdr.resid, "ann_agr_pland")

ggplot(sdr.resid.preds[[1]], aes(x, sdr_resid)) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
  geom_point(data = sdr.resid.preds[[2]], 
             mapping = aes(color = ann_agr_pland), 
             size = 0.6) + scale_color_continuous(type = "viridis") +
  facet_wrap(~pred.var, scales = "free_x") +
  xlab("z-score") +
  egg::theme_article()

```

### 2025-03-11 old/landscapemodels_margin.Rmd
new results

--- 2025-01-08 landscapemodels_margin.Rmd
@@
   egg::theme_article()
+
+```
+
+```{r plot_pol2, include=FALSE}
+
+# Above makes predictions based on population averages for all other variables. Look at predictions with assumed low and high levels of annual agriculture for other variables
+
+# polplot.df_agr5 <- plot_fun(pol.gam1, 
+#                        c("aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi", "lu_ta"),
+#                        wfs_lsm, "ann_agr_pland",
+#                        return_data = FALSE,
+#                        typical = "mean",
+#                        condition = c(ann_agr_pland = 5))
+
+# polplot.df_agr75 <- plot_fun(pol.gam1, 
+#                        c("aglu_iji", "prn_agr_pland", "aglu_frac_mn", "lu_shdi", "lu_ta"),
+#                        wfs_lsm, "ann_agr_pland",
+#                        typical = "mean",
+#                        condition = c(ann_agr_pland = 75))
+
+# ggplot(polplot.df[[1]], aes(x, delta_pollinators.margin)) +  
+#   geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) + geom_line() +
+#   geom_point(data = polplot.df[[2]], 
+#              mapping = aes(color = ann_agr_pland), 
+#              size = 0.6) + scale_color_continuous(type = "viridis") +
+#   facet_wrap(~pred.var, scales = "free_x") +
+#   egg::theme_article()
 

### 2025-03-11 old/newresults_20250308.Rmd
new results

---
title: "New results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load_data}
alldata <- readRDS("data/allresults_20250308.rds")
lsm <- read.csv("data/lsm_reduced.csv") %>%
  mutate(hu = gsub("hu", "", hu))
hu_meta <- readRDS("data/hu_meta.rds")

delta <- alldata$baseline %>% select(hu:sed_export,rawsdr_avoid_exp) %>%
  left_join(alldata$maxwfs %>% select(hu:margin_cells,rawsdr_avoid_exp),
            by = "hu", suffix = c("_base","_max")) %>%
  mutate(
    delta_pollinators = pollinators_max - pollinators_base,
    delta_n_ret = n_ret_max - n_ret_base,
    delta_qb = qb_max - qb_base,
    delta_avoidsedexp = rawsdr_avoid_exp_max - rawsdr_avoid_exp_base,
    delta_n_export = n_export_base - n_export_max,  # change per ha
    delta_sed = sed_export_base - sed_export_max   # change per ha
  ) %>%
  select(hu, delta_pollinators, delta_n_ret, delta_qb, delta_avoidsedexp, delta_n_export, delta_sed) %>%
  left_join(lsm)

ggplot(delta, aes(ntr_pland, delta_sed)) + geom_point()
ggplot(delta, aes(ntr_pland, delta_avoidsedexp)) + geom_point()
ggplot(delta, aes(ntr_pland, delta_n_export)) + geom_point()
ggplot(delta, aes(ntr_pland, delta_n_ret)) + geom_point()

```

### 2025-03-11 randomforest_kl.Qmd
new results

--- 2025-03-11 docs/RandomForest/randomforest.Qmd
@@
 
-    wfs <- readRDS("landscape-complexity-main/data/wfs_lsm.rds")
+    # wfs0 <- readRDS("data/old/wfs_lsm.rds")
+      wfs <- readRDS("data/allresults_20250308.rds")
 
-    dat.predictors <-  wfs[, 20:50]
+    dat.predictors <-  wfs[, c(81:107,120:123)]
 
@@
 
-    if(file.exists("ES-RF.RDA")) load("ES-RF.RDA")
+    # if(file.exists("RandomForest/ES-RF.RDA")) load("ES-RF.RDA")
 
@@
 
-        dat <- data.frame(y = wfs$delta_pollinators.margin, dat.predictors)
+        dat <- data.frame(y = wfs$delta_pollinators_max.margin, dat.predictors)
         pollinators_margin.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
@@
 
-        dat <- data.frame(y = wfs$delta_n_export.margin, dat.predictors)
+        dat <- data.frame(y = wfs$delta_n_export_max.margin, dat.predictors)
         n_export_margin.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
@@
 
-        dat <- data.frame(y = wfs$delta_sed.margin, dat.predictors)
+        dat <- data.frame(y = wfs$delta_sed_max.margin, dat.predictors)
         sed_export_margin.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
@@
 
-        dat <- data.frame(y = wfs$delta_qb.margin, dat.predictors)
+        dat <- data.frame(y = wfs$delta_qb_max.margin, dat.predictors)
         qb_margin.rf <- ranger(y ~ ., data = dat, importance = "permutation", num.trees = 1000)
@@
 
-    save(list = ls(pattern = "^[npqs]"), file = "ES-RF.RDA")
+    save(list = ls(pattern = "^[npqs]"), file = "data/ES-RF.RDA")
 

### 2025-03-11 data-raw/hulandstats.R

@@ -3,7 +3,7 @@ library(tidyverse)
 library(terra)
 library(sf)
 
-datadir <- "/Users/kevinli/Box Sync/InVEST/Optimization/Data/maxWFS/hu/"
+datadir <- "/Users/kevinli/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/maxWFS/hu/"
 hu.files <- list.files(datadir)
 
 hustats <- list()

### 2025-03-11 docs/data-raw/out_20250308.R

@@ -1,0 +1,108 @@
+library(tidyverse)
+
+datadir <- "data-raw/out_20250308/"
+
+outfiles <- list.files(datadir)
+
+base_out <- unlist(
+  lapply(paste0(datadir,outfiles[grepl("base*", outfiles)]), readRDS),
+  recursive = FALSE) %>%
+  bind_rows(.id = "hu") %>%
+  mutate(scenario="baseline")
+
+max_out <- unlist(
+  lapply(paste0(datadir,outfiles[grepl("max*", outfiles)]), readRDS),
+  recursive = FALSE) %>%
+  bind_rows(.id = "hu") %>%
+  mutate(scenario="maxwfs")
+
+one_out <- unlist(
+  lapply(paste0(datadir,outfiles[grepl("onefield*", outfiles)]), readRDS),
+  recursive = FALSE) %>%
+  bind_rows(.id = "hu") %>%
+  mutate(scenario="onefield")
+
+# add other data
+lsm <- read.csv("data/lsm_reduced.csv") %>%
+  mutate(hu = gsub("hu", "", hu))
+hu_meta <- readRDS("data/hu_meta.rds") %>% sf::st_drop_geometry()
+hulandstats <- readRDS("data/hulandstats.rds")
+
+## maximum vs baseline
+
+maxdelta <- base_out %>% select(hu:sed_export,rawsdr_avoid_exp, field_cells) %>%
+  left_join(max_out %>% select(hu:margin_cells,rawsdr_avoid_exp, margin_cells),
+            by = "hu", suffix = c("_base","_max")) %>%
+  mutate(
+    delta_pollinators = pollinators_max - pollinators_base,
+    # delta_n_ret = n_ret_max - n_ret_base,  # N retention - not used because N loading changes between scenarios so N retention is working from different assumptions
+    delta_qb = qb_max - qb_base,   # change per ha
+    # delta_avoidsedexp = rawsdr_avoid_exp_max - rawsdr_avoid_exp_base,  # avoided sediment export - not used because, similar to N retention, different scenarios work with different assumptions (pretty sure upslope sources of sedoiment changes)
+    delta_n_export = n_export_base - n_export_max,  # change per ha
+    delta_sed = sed_export_base - sed_export_max   # change per ha
+  ) %>%
+  select(hu, 
+         pollinators_base, qb_base, n_export_base, sed_export_base, 
+         pollinators_max, qb_max, n_export_max, sed_export_max,
[…]

### 2025-03-11 docs/scripts/atlas_max_one.R

@@ -1,0 +1,126 @@
+arryindx <- as.numeric(commandArgs(trailingOnly = TRUE))
+
+suppressMessages(library(devtools))
+suppressMessages(library(tidyverse))
+suppressMessages(library(sf))
+suppressMessages(library(terra))
+suppressMessages(library(optimEcoServices))
+
+# projdir <- "/Users/kevinli/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/maxWFS/"
+projdir <- "/90daydata/geoecoservices/optimization/maxWFS/"
+
+datadir <- paste0(projdir,"data/")
+
+# biophys directory
+biophysdir <- paste0(datadir, "biophys")
+
+# onefields
+fields.select <- readRDS(paste0(datadir,"selectfields.rds"))
+
+hu <- st_read(paste0(datadir, "hu12watersheds.gpkg"))
+
+hu.names <- hu$huc12
+
+batchnum <- (arryindx-1)*23+seq(1,23)
+
+baseline <- list()
+maxwfs <- list()
+onefield <- list()
+
+for(i in batchnum){
+  i.start <- Sys.time()
+  
+  hu.name <- hu.names[i]
+  
+  cat(paste(hu.name, "started on", lubridate::as_datetime(i.start), "\n"))
+
+  hudir <- paste0(projdir, "hu/hu",hu.name,"/")
+  hu.fields <- rast(paste0(hudir, "data/lu/fields", ".tif"))
+  hu.margins <- rast(paste0(hudir, "data/lu/margins", ".tif"))
+
+  hu.marginids <- unique(terra::values(hu.margins$id, na.rm=TRUE)) # added unique() because ids are section-based, not cell based
+  basemargin <- terra::ifel(test = !is.na(hu.margins$id), 0, 0)
+  maxmargin <- terra::ifel(test = !is.na(hu.margins$id), 1, 0)
+
+  # baseline
+  base.start <- Sys.time()
+  baseline[[hu.name]] <- wfs_fitness2(
+    margin.rast = basemargin,
+    fields.rast = hu.fields,
+    margin.lc = 171,   # land cover for 'pollinator habitat'
+    biophys.dir = biophysdir,
+    data.dir =  paste0(hudir, "data"),
+    workspace.dir = paste0(hudir,"baseline"),
+    suffix="base"
+  )
+  cat(
+    paste(
+      "baseline took",
[…]

### 2025-03-11 docs/scripts/atlas_test.R

@@ -1,0 +1,122 @@
+suppressMessages(library(devtools))
+suppressMessages(library(tidyverse))
+suppressMessages(library(sf))
+suppressMessages(library(terra))
+suppressMessages(library(optimEcoServices))
+
+# projdir <- "/Users/kevinli/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/maxWFS/"
+projdir <- "/90daydata/geoecoservices/optimization/maxWFS/"
+
+datadir <- paste0(projdir,"data/")
+
+# biophys directory
+biophysdir <- paste0(datadir, "biophys")
+
+# onefields
+fields.select <- readRDS(paste0(datadir,"selectfields.rds"))
+
+hu <- st_read(paste0(datadir, "hu12watersheds.gpkg"))
+
+hu.names <- hu$huc12
+
+baseline <- list()
+maxwfs <- list()
+onefield <- list()
+
+for(i in 1){
+  i.start <- Sys.time()
+  
+  hu.name <- hu.names[i]
+  
+  cat(paste(hu.name, "started on", lubridate::as_datetime(i.start), "\n"))
+
+  hudir <- paste0(projdir, "hu/hu",hu.name,"/")
+  hu.fields <- rast(paste0(hudir, "data/lu/fields", ".tif"))
+  hu.margins <- rast(paste0(hudir, "data/lu/margins", ".tif"))
+
+  hu.marginids <- unique(terra::values(hu.margins$id, na.rm=TRUE)) # added unique() because ids are section-based, not cell based
+  basemargin <- terra::ifel(test = !is.na(hu.margins$id), 0, 0)
+  maxmargin <- terra::ifel(test = !is.na(hu.margins$id), 1, 0)
+
+  # baseline
+  base.start <- Sys.time()
+  baseline[[hu.name]] <- wfs_fitness2(
+    margin.rast = basemargin,
+    fields.rast = hu.fields,
+    margin.lc = 171,   # land cover for 'pollinator habitat'
+    biophys.dir = biophysdir,
+    data.dir =  paste0(hudir, "data"),
+    workspace.dir = paste0(hudir,"baseline"),
+    suffix="base"
+  )
+  cat(
+    paste(
+      "baseline took",
+      round(as.numeric(difftime(Sys.time(), base.start, units="secs")), 0),
+      "seconds\n"
+    )
+    )
+
+  # max wfs
+  maxwfs.start <- Sys.time()
[…]

### 2025-03-11 docs/scripts/getCropBoundaries.R

@@ -1,0 +1,45 @@
+library(doParallel)
+library(tidyverse)
+library(sf)
+library(terra)
+
+datadir <- "~/Box Sync/InVEST/GIS_data/"
+
+cropbdry <- paste0(datadir,  "Crop boundaries/2022_National_CSB_gdb/CSB1522.gdb")
+hu <- st_read(paste0(datadir,"NEMWregion/selected_hu12_v2_candidates.gpkg"))
+
+# project to match crop boundry data
+hu.aea <- st_transform(hu, cropbdry.crs)
+
+# setup parallel
+
+parallel::detectCores()
+
+cl <- parallel::makeCluster(
+  8, type = "FORK"
+)
+
+#register it to be used by %dopar%
+doParallel::registerDoParallel(cl = cl)
+
+#check if it is registered (optional)
+foreach::getDoParRegistered()
+
+# loop through
+focalhu.cropbdry <- foreach(i = 1:nrow(hu.aea)) %dopar% {
+  hu.i <- hu.aea[i,]
+
+  query_cropbdry(hu.i, cropbdry.path = cropbdry)
+}
+
+names(focalhu.cropbdry) <- hu.aea %>% pull(huc12)
+
+hucrp <- bind_rows(focalhu.cropbdry, .id = "huc12")
+
+st_write(hucrp, paste0(datadir,"NEMWregion/hucrpbdry_437.gpkg"))
+
+hucrp_ag22 <- hucrp %>%
+  left_join(cdl.codes %>% select(lucode, CLASS_NAME, landuse_group, landcover_ag), by = c("R22"="lucode")) %>%
+  filter(landuse_group == "agriculture")
+
+st_write(hucrp_ag22, paste0(datadir,"NEMWregion/hucrpbdry_437_2022ag.gpkg"), append=FALSE)

### 2025-03-11 docs/scripts/getDataSarahServer.R

@@ -1,0 +1,196 @@
+# must be run from home directory /home/kevin on Sarah's server!
+
+library(EnvToolbox, lib.loc="/home/kevin/Rlib")
+library(sf)
+# library(terra)
+
+# hu <- sf::st_read("/home/kevin/selected_hu12_v2.gpkg")
+hu <- st_read("/home/kevin/selected_hu12_v2_candidates.gpkg")
+hu.buf <- st_buffer(hu, 1000)
+
+# parent directory
+parentdir <- "/home/kevin/gisdata/hu_437/"
+
+dir.create(parentdir, recursive = TRUE)
+
+# extract erodibility
+dir.create(paste0(parentdir, "erodibility2"))
+
+for(i in 1:nrow(hu)){
+  hu.i <- hu.buf[i,]
+  hu.name <- hu.i$huc12
+
+  hu.erod <- querySoils(dat=hu.i,
+                        return.grids = TRUE,
+                        index = "gssurgo",
+                        ssurgo.table = "chorizon",
+                        soil.layers = "kffact", # k-factor (erodibility)
+                        horizon.depth = 30,     # depth-weighted mean down to 10cm
+                        verbose=FALSE)
+
+  terra::writeRaster(hu.erod, paste0(parentdir, "erodibility2/erod_hu",
+                              hu.name,".tif"))
+}
+
+# extract hydrogroup
+dir.create(paste0(parentdir, "hydrogroup"))
+
+for(i in 1:nrow(hu)){
+  hu.i <- hu.buf[i,]
+  hu.name <- hu.i$huc12
+
+  hu.hdgrp <- querySoils(dat=hu.i,
+                        return.grids = TRUE,
+                        index = "gssurgo",
+                        ssurgo.table = "muaggatt",
+                        soil.layers = "hydgrpdcd",  # dominant hydrogroup
+                        verbose=FALSE)
+
+  terra::writeRaster(hu.hdgrp, paste0(parentdir, "hydrogroup/hdgrp_hu",
+                                     hu.name,".tif"))
+}
+
+# extract  cdl landcover
+dir.create(paste0(parentdir, "cdl"))
+
+for(i in 1:nrow(hu)){
+  hu.i <- hu.buf[i,]
+  hu.name <- hu.i$huc12
+
+  hu.cdl <- queryLandcover(dat=hu.i,
+                            yr = 2022, index = "CDL",
+                            return.grids = TRUE,
+                            verbose=FALSE)
+
[…]

### 2025-03-11 docs/scripts/makeScenarioData.R

@@ -1,0 +1,165 @@
+# create crop boundary data
+library(optimEcoServices)
+library(terra)
+library(sf)
+library(tidyverse)
+
+# crop boundaries
+crpbdry <- sf::st_read("/Users/kevinli/Box Sync/InVEST/GIS_data/NEMWregion/hucrpbdry_437_2022ag.gpkg")
+
+cdldir <- "/Users/kevinli/Box Sync/InVEST/GIS_data/Pareto_proj/hu_437_out/cdl/"
+
+hu <- sf::st_read("/Users/kevinli/Box Sync/InVEST/GIS_data/NEMWregion/selected_hu12_v2_candidates.gpkg") %>%
+  sf::st_cast("POLYGON")
+
+
+# make hu data directories and populate
+# hudir <- "/Users/kevinli/Box Sync/InVEST/Optimization/Data/maxWFS/hu/"
+hudir <- "/Users/kevinli/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/maxWFS/hu/"
+if(!dir.exists(hudir)){dir.create(hudir, recursive = TRUE)}
+
+field.reclass <- cdl.codes %>% select(lucode, landuse_group) %>%
+  mutate(landuse_group = ifelse(landuse_group=="agriculture",1,NA))
+
+hu.names <- hu$huc12
+
+# "out" data filepath - where the cleaned data is stored (from the "makeCleanInputData.R" script)
+out.filepath <- "/Users/kevinli/Box Sync/InVEST/GIS_data/Pareto_proj/hu_437_out/"
+
+# Parallel processing
+n.cores <- parallel::detectCores() - 1
+
+#create a cluster
+my.cluster <- parallel::makeCluster(
+  n.cores,
+  type = "PSOCK"
+)
+#register it to be used by %dopar%
+library(doParallel)
+doParallel::registerDoParallel(cl = my.cluster)
+foreach::getDoParRegistered()
+
+# library(doFuture)
+# registerDoFuture()  ## %dopar% parallelizes via future
+# plan(multicore)     ## forked parallel processing (via 'parallel')
+
+# for(i in 1:nrow(hu))
+foreach(i = 1:nrow(hu), .packages = c("terra","sf","dplyr")) %dopar%
+  {
+    devtools::load_all()
+  hu.name <- hu.names[i]
+
+  # make data folder for this hu
+  hudir.i <- paste0(hudir, "hu", hu.name,"/data/")
[…]

### 2025-03-11 docs/scripts/onefield.R

@@ -1,0 +1,60 @@
+library(tidyverse)
+library(sf)
+library(terra)
+# library(future)
+# library(doFuture)
+# devtools::install()
+library(optimEcoServices)
+
+projdir <- "/Users/kevinli/Box Sync/InVEST/Optimization/Data/maxWFS/"
+datadir <- paste0(projdir,"data/")
+
+hu <- st_read(paste0(datadir, "hu12watersheds.gpkg"))
+fields.select <- readRDS(paste0(datadir,"selectfields.rds"))
+
+hu.names <- hu$huc12
+
+hu.names.109 <- hu.names[1:109]
+hu.names.218 <- hu.names[110:218]
+hu.names.327 <- hu.names[219:327]
+hu.names.437 <- hu.names[328:437]
+
+hu.onefields <- list()
+
+for(i in seq_along(hu.names.109)){
+  hu.name <- hu.names.109[i]
+
+  hudir <- paste0(projdir, "hu/hu",hu.name,"/")
+  hu.fields <- rast(paste0(hudir, "data/lu/fields", ".tif"))
+  hu.fields.select.ids <- fields.select %>%
+    filter(hu == hu.name) %>% pull(id)
+
+  # randomly draw 10 without replacement
+  hu.fields.select.10 <- hu.fields.select.ids[
+    sample(seq_along(hu.fields.select.ids), size = 10, replace = FALSE)
+  ]
+
+  onefields <- list()
+
+  for(j in seq_along(hu.fields.select.10)){
+    field.ij <- terra::ifel(
+      test = hu.fields$id == hu.fields.select.10[j],
+      yes = 1, no = NA
+    )
+    margin.ij <- terra::boundaries(field.ij)
+
+    # add wfs to field j
+    onefields[[j]] <- wfs_fitness2(
+      margin.rast = margin.ij,
+      fields.rast = hu.fields,
+      margin.lc = 171,   # land cover for 'pollinator habitat'
+      biophys.dir = "biophys",
+      data.dir =  paste0(hudir, "data"),
+      workspace.dir = paste0(hudir,"onefield"),
+      suffix="onefield"
+    )
+  }
+  hu.onefields[[i]] <- bind_rows(onefields) %>% mutate(hu = hu.name)
+}
+
+hu.109 <- bind_rows(hu.onefields)

### 2025-03-11 docs/scripts/sbatch-test.sh

@@ -1,0 +1,24 @@
+#!/bin/bash -l
+
+#SBATCH --partition=atlas
+#SBATCH --qos=normal
+#SBATCH --job-name=landopt
+#SBATCH --output=/90daydata/geoecoservices/optimization/landcomplex-%j.out
+#SBATCH --error=/90daydata/geoecoservices/optimization/landcomplex-%j.err
+#SBATCH --account=geoecoservices
+#SBATCH --mail-user=likevin@psu.edu
+#SBATCH --mail-type=NONE
+#SBATCH --time=24:00:00
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=2
+#SBATCH --get-user-env              ### Import your user environment setup
+
+
+module purge
+
+module load miniconda3
+source activate
+conda activate r-invest2
+
+srun Rscript --vanilla /project/geoecoservices/optimization/landscape-complexity/scripts/atlas_test.R

### 2025-03-11 docs/scripts/sbatchlandcomplex.sh

@@ -1,0 +1,24 @@
+#!/bin/bash -l
+
+#SBATCH --partition=atlas
+#SBATCH --qos=normal
+#SBATCH --job-name=landopt
+#SBATCH --output=/90daydata/geoecoservices/optimization/landcomplex-%j.out
+#SBATCH --error=/90daydata/geoecoservices/optimization/landcomplex-%j.err
+#SBATCH --account=geoecoservices
+#SBATCH --mail-user=likevin@psu.edu
+#SBATCH --mail-type=NONE
+#SBATCH --time=24:00:00
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=4
+#SBATCH --get-user-env              ### Import your user environment setup
+#SBATCH --array=1-19                ### Array index
+
+module purge
+
+module load miniconda3
+source activate
+conda activate r-invest2
+
+srun Rscript --vanilla /project/geoecoservices/optimization/landscape-complexity/scripts/atlas_max_one.R $SLURM_ARRAY_TASK_ID

### 2025-03-11 randomforest_kl.Qmd
add random forest

@@
 
-    # if(file.exists("RandomForest/ES-RF.RDA")) load("ES-RF.RDA")
+    if(file.exists("data/ES-RF.RDA")) load("data/ES-RF.RDA")
 
@@
 
-    with(wfs, plot(pollinators_base, delta_pollinators.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Pollination"))
+    with(wfs, plot(pollinators_base, delta_pollinators_max.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Pollination"))
 
-    with(wfs, plot(ann_agr_pland, delta_pollinators.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Pollination"))
+    with(wfs, plot(ann_agr_pland, delta_pollinators_max.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Pollination"))
 
@@
 
-    with(wfs, plot(n_export_base, delta_n_export.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Nitrogen"))
+    with(wfs, plot(n_export_base, delta_n_export_max.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Nitrogen"))
 
-    with(wfs, plot(ann_agr_pland, delta_n_export.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Nitrogen"))
+    with(wfs, plot(ann_agr_pland, delta_n_export_max.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Nitrogen"))
 
@@
 
-    with(wfs, plot(sed_export_base, delta_sed.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Sediment"))
+    with(wfs, plot(sed_export_base, delta_sed_max.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Sediment"))
 
-    with(wfs, plot(ann_agr_pland, delta_sed.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Sediment"))
+    with(wfs, plot(ann_agr_pland, delta_sed_max.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Sediment"))
 
@@
 
-    with(wfs, plot(qb_base, delta_qb.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Baseflow"))
+    with(wfs, plot(qb_base, delta_qb_max.margin, col = "darkgreen", xlab = "Base value", ylab = "Delta", main = "Baseflow"))
 
-    with(wfs, plot(ann_agr_pland, delta_qb.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Baseflow"))
+    with(wfs, plot(ann_agr_pland, delta_qb_max.margin, col = "purple", xlab = "Annual agriculture (%)", ylab = "Delta", main = "Baseflow"))
 
@@
     if(!exists("pollinators_base.ann_agr_pland.pdp")) {
-        dat <- data.frame(y = wfs$delta_pollinators.margin, orig = wfs$pollinators_base, dat.predictors)
+        dat <- data.frame(y = wfs$delta_pollinators_max.margin, orig = wfs$pollinators_base, dat.predictors)
         pollinators_base.ann_agr_pland.pdp <- partial(pollinators_base.rf, train = dat, pred.var = "ann_agr_pland")
@@
     if(!exists("n_export_base.ann_agr_pland.pdp")) {
-        dat <- data.frame(y = wfs$delta_n_export.margin, orig = wfs$n_export_base, dat.predictors)
+        dat <- data.frame(y = wfs$delta_n_export_max.margin, orig = wfs$n_export_base, dat.predictors)
         n_export_base.ann_agr_pland.pdp <- partial(n_export_base.rf, train = dat, pred.var = "ann_agr_pland")
@@
     if(!exists("sed_export_base.ann_agr_pland.pdp")) {
-        dat <- data.frame(y = wfs$delta_sed.margin, orig = wfs$sed_export_base, dat.predictors)
+        dat <- data.frame(y = wfs$delta_sed_max.margin, orig = wfs$sed_export_base, dat.predictors)
         sed_export_base.ann_agr_pland.pdp <- partial(sed_export_base.rf, train = dat, pred.var = "ann_agr_pland")
@@
     if(!exists("qb_base.ann_agr_pland.pdp")) {
-        dat <- data.frame(y = wfs$delta_qb.margin, orig = wfs$qb_base, dat.predictors)
+        dat <- data.frame(y = wfs$delta_qb_max.margin, orig = wfs$qb_base, dat.predictors)
         qb_base.ann_agr_pland.pdp <- partial(qb_base.rf, train = dat, pred.var = "ann_agr_pland")

### 2025-03-17 randomforest_kl.Qmd
random forest kl

@@
 
-    p1 <- plotPartial(sed_export_base.ann_agr_pland.pdp, main = "Sediment base")
+    p1 <- plotPartial(sed_export_base.ann_agr_pland.pdp, main = "Sediment base", rug = TRUE)
     p2 <- plotPartial(sed_export_margin.ann_agr_pland.pdp, main = "Sediment margin")

### 2025-03-22 examine_random_forest.Rmd
Create examine_random_forest.Rmd

---
title: "Look at random forest"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

```{r load_rf}
library(tidyverse)
library(ranger)
library(pdp)
load("data/ES-RF.RDA")

wfs <- readRDS("data/allresults_20250308.rds")

dat.predictors <-  wfs[, c(81:107,120:123)]

# drop log-transform
dat.predictors <- dat.predictors[, !grepl("log", colnames(dat.predictors))]

```

```{r pca}
library(ggbiplot)
pca.predictors <- prcomp(dat.predictors,
                         scale. = TRUE, center = TRUE)

ggbiplot(pca.predictors, c(2,3))
```

## Including Plots

You can also embed plots, for example:

```{r }

    if(!exists("pollinators_base.ann_agr_pland.pdp2")) {
        dat <- data.frame(y = wfs$delta_pollinators_max.margin, orig = wfs$pollinators_base, dat.predictors)
        pollinators_base.ann_agr_pland.pdp2 <- pdp::partial(pollinators_base.rf, train = dat, pred.var = "ann_agr_pland", ice = TRUE, rug = TRUE)
        pollinators_margin.ann_agr_pland.pdp2 <- pdp::partial(pollinators_margin.rf, train = dat, pred.var = "ann_agr_pland", ice = TRUE, rug = TRUE)
        pollinators_base.ntr_pland.pdp2 <- pdp::partial(pollinators_base.rf, train = dat, pred.var = "ntr_pland", ice = TRUE, rug = TRUE)
        pollinators_margin.ntr_pland.pdp2 <- pdp::partial(pollinators_margin.rf, train = dat, pred.var = "ntr_pland", ice = TRUE, rug = TRUE)
        rm(dat)
    }

    if(!exists("n_export_base.ann_agr_pland.pdp2")) {
        dat <- data.frame(y = wfs$delta_n_export_max.margin, orig = wfs$n_export_base, dat.predictors)
        n_export_base.ann_agr_pland.pdp <- partial(n_export_base.rf, train = dat, pred.var = "ann_agr_pland")
        n_export_margin.ann_agr_pland.pdp <- partial(n_export_margin.rf, train = dat, pred.var = "ann_agr_pland")
        n_export_base.ntr_pland.pdp <- partial(n_export_base.rf, train = dat, pred.var = "ntr_pland")
        n_export_margin.ntr_pland.pdp <- partial(n_export_margin.rf, train = dat, pred.var = "ntr_pland")
        rm(dat)
    }

    if(!exists("sed_export_base.ann_agr_pland.pdp2")) {
        dat <- data.frame(y = wfs$delta_sed_max.margin, orig = wfs$sed_export_base, dat.predictors)
        sed_export_base.ann_agr_pland.pdp <- partial(sed_export_base.rf, train = dat, pred.var = "ann_agr_pland")
        sed_export_margin.ann_agr_pland.pdp <- partial(sed_export_margin.rf, train = dat, pred.var = "ann_agr_pland")
        sed_export_base.ntr_pland.pdp <- partial(sed_export_base.rf, train = dat, pred.var = "ntr_pland")
        sed_export_margin.ntr_pland.pdp <- partial(sed_export_margin.rf, train = dat, pred.var = "ntr_pland")
        rm(dat)
    }

    if(!exists("qb_base.ann_agr_pland.pdp2")) {
        dat <- data.frame(y = wfs$delta_qb_max.margin, orig = wfs$qb_base, dat.predictors)
        qb_base.ann_agr_pland.pdp <- partial(qb_base.rf, train = dat, pred.var = "ann_agr_pland")
        qb_margin.ann_agr_pland.pdp <- partial(qb_margin.rf, train = dat, pred.var = "ann_agr_pland")
        qb_base.ntr_pland.pdp <- partial(qb_base.rf, train = dat, pred.var = "ntr_pland")
        qb_margin.ntr_pland.pdp <- partial(qb_margin.rf, train = dat, pred.var = "ntr_pland")
        rm(dat)
    }

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2025-03-26 Paper_analysis.Rmd
update site

---
title: "Paper analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r load}
library(tidyverse)
load("data/ES-RF2.RDA")

```

## Patterns with composition

Making 4 x 3 plots, each showing relationships with

```{r compplots}

dat.long <- dat %>% 
  pivot_longer(cols = 1:12, names_to = "ES_response", values_to = "ES_value") %>%
  pivot_longer(cols = c("ntr_pland", "ann_agr_pland"), names_to = "Landuse",
               values_to = "Landuse_value") %>%
  select(ES_response, ES_value, Landuse, Landuse_value) %>%
  mutate(ES = gsub("delta_|_base|_max.margin","", ES_response)) %>%
  mutate(ES = gsub("sed$","sed_export", ES)) %>%
  mutate(response_type =
           gsub("sed|export|n_|pollinators|qb|_|max","",ES_response))

# fit some gams
dat.gam <- dat.long %>% filter(Landuse == "ntr_pland") %>% 
  group_by(response_type, ES) %>%
  nest() %>%
  # Fit GAM
  mutate(Mod = map(data, ~mgcv::gam(ES_value ~ s(Landuse_value), data = .x))) %>%
  # Get R2
  mutate(R2 = map_dbl(Mod, ~round(summary(.x)$r.sq, 2)))

ggplot(dat.long %>% filter(Landuse == "ntr_pland"),
       aes(Landuse_value, ES_value)) + 
  geom_point() + geom_smooth(method="gam") +
  ggh4x::facet_grid2(response_type ~ ES, 
                                    scales = "free", independent = "y") + 
  egg::theme_article() + 
  # Add label
  geom_label(data = dat.gam, 
             aes(x = Inf, y = Inf, 
                 label = paste("R2 = ", R2, sep = " ")),
             hjust = 1, vjust = 1)
```

# Model performance
*Pollinators:*

   - Base

     r2 = `r round(pollinators_base.rf$r.squared, 3)`
     OOB MSE = `r round(pollinators_base.rf$prediction.error, 3)`

  - Margin

     r2 = `r round(pollinators_margin.rf$r.squared, 3)`
     OOB MSE = `r round(pollinators_margin.rf$prediction.error, 3)`

*Nitrogen export:*

   - Base

     r2 = `r round(n_export_base.rf$r.squared, 3)`
     OOB MSE = `r round(n_export_base.rf$prediction.error, 3)`

  - Margin

     r2 = `r round(n_export_margin.rf$r.squared, 3)`
     OOB MSE = `r round(n_export_margin.rf$prediction.error, 3)`

*Sediment export:*

   - Base

     r2 = `r round(sed_export_base.rf$r.squared, 3)`
     OOB MSE = `r round(sed_export_base.rf$prediction.error, 3)`

  - Margin

     r2 = `r round(sed_export_margin.rf$r.squared, 3)`
     OOB MSE = `r round(sed_export_margin.rf$prediction.error, 3)`

*Baseflow:*

   - Base

     r2 = `r round(qb_base.rf$r.squared, 3)`
     OOB MSE = `r round(qb_base.rf$prediction.error, 3)`

  - Margin

     r2 = `r round(qb_margin.rf$r.squared, 3)`
     OOB MSE = `r round(qb_margin.rf$prediction.error, 3)`

# Variable importance

```{r importance, fig.height=16, fig.width=9}
library(ggpattern)

metric.descr <- read.csv("data/lsm_description.csv")

imp.df <- bind_rows(
  pollinators_base.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "Pollinators", model = "base"),
  pollinators_margin.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "Pollinators", model = "margin"),
  n_export_base.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "N export", model = "base"),
  n_export_margin.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "N export", model = "margin"),
  sed_export_base.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "Sed export", model = "base"),
  sed_export_margin.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "Sed export", model = "margin"),
  qb_base.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "Baseflow", model = "base"),
  qb_margin.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "Baseflow", model = "margin")
) %>%
  mutate(sig = ifelse(pvalue <0.05, TRUE, FALSE)) %>%
  mutate(metric2 = tidytext::reorder_within(metric, importance,
                                           within = list(ES, model))) %>%
  left_join(metric.descr %>% select(lsm, type), by = c("metric"="lsm")) %>%
  mutate(type = case_match(
    metric,
    "field_pland" ~ "area and edge metric",
    "field_frac_mn" ~ "shape metric",
    "field_frac_cv" ~ "shape metric",
    "prn_ann_ratio" ~ "area and edge metric",
    .default = type
  ))

ggplot(imp.df, aes(metric2, importance, fill = type, pattern = sig)) + 
  geom_bar_pattern(stat="identity",
                   color = "black", 
                   pattern_fill = "black",
                   pattern_angle = 45,
                   pattern_density = 0.1,
                   pattern_spacing = 0.025,
                   pattern_key_scale_factor = 0.6) + 
  tidytext::scale_x_reordered() +
  coord_flip() +
  ggh4x::facet_grid2(ES~model, scales = "free", independent = "all") +
  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
  egg::theme_article() +
  theme(legend.position = "top") + 
  guides(pattern = guide_legend(override.aes = list(fill = "white")),
         fill = guide_legend(override.aes = list(pattern = "none")))

```

*Significant metrics count*

```{r}

sigct <- imp.df %>% filter(sig==TRUE)

ggplot(sigct, aes(fct_infreq(metric), fill = type)) +
  scale_x_discrete(limits=rev) +
  geom_bar() + coord_flip() + labs(x="metric") + egg::theme_article()

```

# Interactions
```{r int_load}
library(iml)
library(gridExtra)
load("data/ES-RF2-preds.RDA")
```

## Pollinator base interactions

```{r polbase_int}
grid.arrange(
  int.polbase %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
  int.polbase.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
  ncol = 2
)
```

## Pollinators margins interactions

```{r polmarg_int}
grid.arrange(
  int.polmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
  int.polmarg.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
  ncol = 2
)
```

### 2025-03-26 randomforest_kl.Qmd
update site

@@
 
-     r2 = `{r} round(pollinators_base.rf$r.squared, 3)`
-     OOB MSE = `{r} round(pollinators_base.rf$prediction.error, 3)`
+     r2 = `r round(pollinators_base.rf$r.squared, 3)`
+     OOB MSE = `r round(pollinators_base.rf$prediction.error, 3)`
 
@@
 
-     r2 = `{r} round(pollinators_margin.rf$r.squared, 3)`
-     OOB MSE = `{r} round(pollinators_margin.rf$prediction.error, 3)`
+     r2 = `r round(pollinators_margin.rf$r.squared, 3)`
+     OOB MSE = `r round(pollinators_margin.rf$prediction.error, 3)`
 
@@
 
-     r2 = `{r} round(n_export_base.rf$r.squared, 3)`
-     OOB MSE = `{r} round(n_export_base.rf$prediction.error, 3)`
+     r2 = `r round(n_export_base.rf$r.squared, 3)`
+     OOB MSE = `r round(n_export_base.rf$prediction.error, 3)`
 
@@
 
-     r2 = `{r} round(n_export_margin.rf$r.squared, 3)`
-     OOB MSE = `{r} round(n_export_margin.rf$prediction.error, 3)`
+     r2 = `r round(n_export_margin.rf$r.squared, 3)`
+     OOB MSE = `r round(n_export_margin.rf$prediction.error, 3)`
 
@@
 
-     r2 = `{r} round(sed_export_base.rf$r.squared, 3)`
-     OOB MSE = `{r} round(sed_export_base.rf$prediction.error, 3)`
+     r2 = `r round(sed_export_base.rf$r.squared, 3)`
+     OOB MSE = `r round(sed_export_base.rf$prediction.error, 3)`
 
@@
 
-     r2 = `{r} round(sed_export_margin.rf$r.squared, 3)`
-     OOB MSE = `{r} round(sed_export_margin.rf$prediction.error, 3)`
+     r2 = `r round(sed_export_margin.rf$r.squared, 3)`
+     OOB MSE = `r round(sed_export_margin.rf$prediction.error, 3)`
 
@@
 
-     r2 = `{r} round(qb_base.rf$r.squared, 3)`
-     OOB MSE = `{r} round(qb_base.rf$prediction.error, 3)`
+     r2 = `r round(qb_base.rf$r.squared, 3)`
+     OOB MSE = `r round(qb_base.rf$prediction.error, 3)`
 
@@
 
-     r2 = `{r} round(qb_margin.rf$r.squared, 3)`
-     OOB MSE = `{r} round(qb_margin.rf$prediction.error, 3)`
+     r2 = `r round(qb_margin.rf$r.squared, 3)`
+     OOB MSE = `r round(qb_margin.rf$prediction.error, 3)`
 

### 2025-03-26 RandomForest/make_rf.R

@@ -1,0 +1,58 @@
+# make random forest from unique datasets
+
+library(ranger)
+
+wfs <- readRDS("data/allresults_20250308.rds")
+
+names(wfs)
+
+dat.predictors <-  wfs[, c(81:107,120,123)]
+
+dat.response <- wfs[, c(2:5, 10:13, 20:23)]
+
+dat <- cbind(dat.response, dat.predictors)
+
+# make rf formulas
+pol.base.form <- reformulate(names(dat.predictors), response = "pollinators_base")
+pol.margin.form <- reformulate(names(dat.predictors), response = "delta_pollinators_max.margin")
+
+n_exp.base.form <- reformulate(names(dat.predictors), response = "n_export_base")
+n_exp.margin.form <- reformulate(names(dat.predictors), response = "delta_n_export_max.margin")
+
+qb.base.form <- reformulate(names(dat.predictors), response = "qb_base")
+qb.margin.form <- reformulate(names(dat.predictors), response = "delta_qb_max.margin")
+
+sed.base.form <- reformulate(names(dat.predictors), response = "sed_export_base")
+sed.margin.form <- reformulate(names(dat.predictors), response = "delta_sed_max.margin")
+
+# run random forest
+pollinators_base.rf <- ranger(pol.base.form, data = dat, importance = "permutation", num.trees = 1000)
+pollinators_base.p <- data.frame(importance_pvalues(pollinators_base.rf, method = "altmann", 
+                                                    formula = pol.base.form, data = dat, num.permutations = 1000))
+pollinators_margin.rf <- ranger(pol.margin.form, data = dat, importance = "permutation", num.trees = 1000)
+pollinators_margin.p <- data.frame(importance_pvalues(pollinators_margin.rf, method = "altmann", 
+                                                      formula = pol.margin.form, data = dat, num.permutations = 1000))
+
+n_export_base.rf <- ranger(n_exp.base.form, data = dat, importance = "permutation", num.trees = 1000)
+n_export_base.p <- data.frame(importance_pvalues(n_export_base.rf, method = "altmann", 
+                                                 formula = n_exp.base.form, data = dat, num.permutations = 1000))
+n_export_margin.rf <- ranger(n_exp.margin.form, data = dat, importance = "permutation", num.trees = 1000)
+n_export_margin.p <- data.frame(importance_pvalues(n_export_margin.rf, method = "altmann", 
+                                                   formula = n_exp.margin.form, data = dat, num.permutations = 1000))
+
[…]

### 2025-03-26 docs/RandomForest/make_rf_interactions.R

@@ -1,0 +1,105 @@
+library(ranger)
+library(iml)
+library(future)
+load("data/ES-RF2.RDA")
+pfun <- function(object, newdata) predict(object, data = newdata)$predictions
+
+future::plan(multisession, workers = 8)
+
+# pollinators base
+
+pred.pollinators_base <- Predictor$new(
+  model = pollinators_base.rf,
+  data = dat,
+  y = "pollinators_base",
+  predict.function = pfun
+)
+
+int.polbase  <- Interaction$new(pred.pollinators_base)
+int.polbase.ntr_pland  <- Interaction$new(pred.pollinators_base, feature = "ntr_pland")
+
+# pollinators margin
+
+pred.pollinators_margin <- Predictor$new(
+  model = pollinators_margin.rf,
+  data = dat,
+  y = "delta_pollinators_max.margin",
+  predict.function = pfun
+)
+
+int.polmarg  <- Interaction$new(pred.pollinators_margin)
+int.polmarg.ntr_pland  <- Interaction$new(pred.pollinators_margin, feature = "ntr_pland")
+
+# N export base
+
+pred.n_export_base <- Predictor$new(
+  model = n_export_base.rf,
+  data = dat,
+  y = "n_export_base",
+  predict.function = pfun
+)
+
+int.nexpbase  <- Interaction$new(pred.n_export_base)
+int.nexpbase.ntr_pland  <- Interaction$new(pred.n_export_base, feature = "ntr_pland")
+
+# N export margin
+
+pred.n_export_margin <- Predictor$new(
+  model = n_export_margin.rf,
+  data = dat,
+  y = "delta_n_export_max.margin",
+  predict.function = pfun
+)
+
+int.nexpmarg  <- Interaction$new(pred.n_export_margin)
+int.nexpmarg.ntr_pland  <- Interaction$new(pred.n_export_margin, feature = "ntr_pland")
+
+# Sediment export base
+
+pred.sed_export_base <- Predictor$new(
+  model = sed_export_base.rf,
+  data = dat,
+  y = "sed_export_base",
+  predict.function = pfun
+)
+
+int.sedexpbase  <- Interaction$new(pred.sed_export_base)
+int.sedexpbase.ann_agr_ed  <- Interaction$new(pred.sed_export_base, feature = "ann_agr_ed")
+
+# Sediment export margin
+
[…]

### 2025-03-26 Paper_analysis.Rmd
update site

@@
 ```
+
+## N export base interactions
+
+```{r nexpbase_int}
+grid.arrange(
+  int.nexpbase %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.nexpbase.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
+)
+```
+
+## N export margin interactions
+
+```{r nexpmarg_int}
+grid.arrange(
+  int.nexpmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.nexpmarg.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
+)
+```
+
+## Sediment export base interactions
+
+```{r sedbase_int}
+grid.arrange(
+  int.sedexpbase %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.sedexpbase.ann_agr_ed %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
+)
+```
+
+## Sediment export margin interactions
+
+```{r sedmarg_int}
+grid.arrange(
+  int.sedexpmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.sedexpmarg.ann_agr_ed %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
+)
+```
+
+## Baseflow base interactions
+
+```{r qbbase_int}
+grid.arrange(
+  int.qbbase %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.qbbase.field_frac_mn %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
+)
+```
+
+## Baseflow margin interactions
+
+```{r qbmarg_int}
+grid.arrange(
+  int.qbmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.qbmarg.prn_agr_para_mn %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
+)
+```

### 2025-03-26 Paper_analysis.Rmd
update site

@@
   int.nexpmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  int.nexpmarg.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.nexpmarg.ann_agr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
   ncol = 2
@@
   int.sedexpmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  int.sedexpmarg.ann_agr_ed %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.sedexpmarg.lu_shdi %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
   ncol = 2
@@
   int.qbmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  int.qbmarg.prn_agr_para_mn %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.qbmarg.field_frac_cv %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
   ncol = 2

### 2025-03-26 docs/RandomForest/make_rf_interactions.R

@@ -52,7 +52,7 @@ pred.n_export_margin <- Predictor$new(
 )
 
 int.nexpmarg  <- Interaction$new(pred.n_export_margin)
-int.nexpmarg.ntr_pland  <- Interaction$new(pred.n_export_margin, feature = "ntr_pland")
+int.nexpmarg.ann_agr_pland  <- Interaction$new(pred.n_export_margin, feature = "ann_agr_pland")
 
 # Sediment export base
 
@@ -76,7 +76,7 @@ pred.sed_export_margin <- Predictor$new(
 )
 
 int.sedexpmarg  <- Interaction$new(pred.sed_export_margin)
-int.sedexpmarg.ann_agr_ed  <- Interaction$new(pred.sed_export_margin, feature = "ann_agr_ed")
+int.sedexpmarg.lu_shdi  <- Interaction$new(pred.sed_export_margin, feature = "lu_shdi")
 
 # Water recharge base
 
@@ -100,6 +100,6 @@ pred.qb_margin <- Predictor$new(
 )
 
 int.qbmarg  <- Interaction$new(pred.qb_margin)
-int.qbmarg.prn_agr_para_mn <- Interaction$new(pred.qb_margin, feature = "prn_agr_para_mn")
+int.qbmarg.field_frac_cv <- Interaction$new(pred.qb_margin, feature = "field_frac_cv")
 
 save(list = ls(pattern = "^int|^pred"), file = "data/ES-RF2-preds.RDA")

### 2025-03-28 Paper_analysis.Rmd
update site

--- 2025-03-26 Paper_analysis.Rmd
@@
 
-# Interactions
-```{r int_load}
-library(iml)
-library(gridExtra)
-load("data/ES-RF2-preds.RDA")
+```{r impurity_importance, fig.height=16, fig.width=9}
+
+load("data/ES-RF2-impurity.RDA")
+
+library(ggpattern)
+
+metric.descr <- read.csv("data/lsm_description.csv")
+
+imp.imp.df <- bind_rows(
+  pollinators_base.imp.p %>% rownames_to_column("metric") %>% 
+    mutate(ES = "Pollinators", model = "base"),
+  pollinators_margin.imp.p %>% rownames_to_column("metric") %>% 
+    mutate(ES = "Pollinators", model = "margin"),
+  n_export_base.imp.p %>% rownames_to_column("metric") %>% 
+    mutate(ES = "N export", model = "base"),
+  n_export_margin.imp.p %>% rownames_to_column("metric") %>% 
+    mutate(ES = "N export", model = "margin"),
+  sed_export_base.imp.p %>% rownames_to_column("metric") %>% 
+    mutate(ES = "Sed export", model = "base"),
+  sed_export_margin.imp.p %>% rownames_to_column("metric") %>% 
+    mutate(ES = "Sed export", model = "margin"),
+  qb_base.imp.p %>% rownames_to_column("metric") %>% 
+    mutate(ES = "Baseflow", model = "base"),
+  qb_margin.imp.p %>% rownames_to_column("metric") %>% 
+    mutate(ES = "Baseflow", model = "margin")
+) %>%
+  mutate(sig = ifelse(pvalue <0.05, TRUE, FALSE)) %>%
+  mutate(metric2 = tidytext::reorder_within(metric, importance,
+                                           within = list(ES, model))) %>%
+  left_join(metric.descr %>% select(lsm, type), by = c("metric"="lsm")) %>%
+  mutate(type = case_match(
+    metric,
+    "field_pland" ~ "area and edge metric",
+    "field_frac_mn" ~ "shape metric",
+    "field_frac_cv" ~ "shape metric",
+    "prn_ann_ratio" ~ "area and edge metric",
+    .default = type
+  ))
+
+ggplot(imp.imp.df, aes(metric2, importance, fill = type, pattern = sig)) + 
+  geom_bar_pattern(stat="identity",
+                   color = "black", 
+                   pattern_fill = "black",
+                   pattern_angle = 45,
+                   pattern_density = 0.1,
+                   pattern_spacing = 0.025,
+                   pattern_key_scale_factor = 0.6) + 
+  tidytext::scale_x_reordered() +
+  coord_flip() +
+  ggh4x::facet_grid2(ES~model, scales = "free", independent = "all") +
+  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
+  egg::theme_article() +
+  theme(legend.position = "top") + 
+  guides(pattern = guide_legend(override.aes = list(fill = "white")),
+         fill = guide_legend(override.aes = list(pattern = "none")))
+
 ```
 
-## Pollinator base interactions
+## Shapley importance
+```{r importance_shap}
+load("data/ES-RF2-shapimportance.RDA")
+library(shapviz)
+library(gridExtra)
 
-```{r polbase_int}
+```
+
+### Pollinators
+
+*Pollinators*
+
+```{r shap_pol}
 grid.arrange(
-  int.polbase %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  int.polbase.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  sv_importance(ex.polbase.glb), sv_importance(ex.polmarg.glb),
   ncol = 2
 )
+
 ```
 
-## Pollinators margins interactions
+*N Export*
 
-```{r polmarg_int}
+```{r shap_Nexp}
 grid.arrange(
-  int.polmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  int.polmarg.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  sv_importance(ex.nexpbase.glb), sv_importance(ex.nexpmarg.glb),
   ncol = 2
 )
+
 ```
+
+*Sediment Export*
+
+```{r shap_sed}
+grid.arrange(
+  sv_importance(ex.sedbase.glb), sv_importance(ex.sedmarg.glb),
+  ncol = 2
+)
+
+```
+
+*Water recharge*
+
+```{r shap_wb}
+grid.arrange(
+  sv_importance(ex.qbbase.glb), sv_importance(ex.qbmarg.glb),
+  ncol = 2
+)
+
+```

### 2025-03-28 RandomForest/make_rf.R

@@ -1,6 +1,7 @@
 # make random forest from unique datasets
 
 library(ranger)
+options(ranger.num.threads = 8)
 
 wfs <- readRDS("data/allresults_20250308.rds")
 
@@ -54,5 +55,38 @@ qb_margin.rf <- ranger(qb.margin.form, data = dat, importance = "permutation", n
 qb_margin.p <- data.frame(importance_pvalues(qb_margin.rf, method = "altmann",
                                              formula = qb.margin.form, data = dat, num.permutations = 1000))
 
-# save the outputs
+# run random forest using different importance
+pollinators_base.imp.rf <- ranger(pol.base.form, data = dat, importance = "impurity", num.trees = 1000)
+pollinators_base.imp.p <- data.frame(importance_pvalues(pollinators_base.imp.rf, method = "altmann", 
+                                                    formula = pol.base.form, data = dat, num.permutations = 1000))
+pollinators_margin.imp.rf <- ranger(pol.margin.form, data = dat, importance = "impurity", num.trees = 1000)
+pollinators_margin.imp.p <- data.frame(importance_pvalues(pollinators_margin.imp.rf, method = "altmann", 
+                                                      formula = pol.margin.form, data = dat, num.permutations = 1000))
+
+n_export_base.imp.rf <- ranger(n_exp.base.form, data = dat, importance = "impurity", num.trees = 1000)
+n_export_base.imp.p <- data.frame(importance_pvalues(n_export_base.imp.rf, method = "altmann", 
+                                                 formula = n_exp.base.form, data = dat, num.permutations = 1000))
+n_export_margin.imp.rf <- ranger(n_exp.margin.form, data = dat, importance = "impurity", num.trees = 1000)
+n_export_margin.imp.p <- data.frame(importance_pvalues(n_export_margin.imp.rf, method = "altmann", 
+                                                   formula = n_exp.margin.form, data = dat, num.permutations = 1000))
+
+sed_export_base.imp.rf <- ranger(sed.base.form, data = dat, importance = "impurity", num.trees = 1000)
+sed_export_base.imp.p <- data.frame(importance_pvalues(sed_export_base.imp.rf, method = "altmann", 
+                                                   formula = sed.base.form, data = dat, num.permutations = 1000))
+sed_export_margin.imp.rf <- ranger(sed.margin.form, data = dat, importance = "impurity", num.trees = 1000)
[…]

### 2025-03-28 RandomForest/make_rf_importance.R

@@ -1,0 +1,20 @@
+# make importance values
+library(iml)
+load("data/ES-RF2.RDA")
+load("data/ES-RF2-preds.RDA")
+
+# pollinators
+imp.polbase <- FeatureImp$new(pred.pollinators_base, loss = "mae")
+imp.polmarg <- FeatureImp$new(pred.pollinators_margin, loss = "mae")
+
+# N export
+imp.nexpbase <- FeatureImp$new(pred.n_export_base, loss = "mae")
+imp.nexpmarg <- FeatureImp$new(pred.n_export_margin, loss = "mae")
+
+# sed export
+imp.sedbase <- FeatureImp$new(pred.sed_export_base, loss = "mae")
+imp.sedmarg <- FeatureImp$new(pred.sed_export_margin, loss = "mae")
+
+# baseflow
+imp.qbbase <- FeatureImp$new(pred.qb_base, loss = "mae")
+imp.qbmarg <- FeatureImp$new(pred.qb_margin, loss = "mae")

### 2025-03-28 RandomForest/make_rf_interactions.R

@@ -103,3 +103,91 @@ int.qbmarg  <- Interaction$new(pred.qb_margin)
 int.qbmarg.field_frac_cv <- Interaction$new(pred.qb_margin, feature = "field_frac_cv")
 
 save(list = ls(pattern = "^int|^pred"), file = "data/ES-RF2-preds.RDA")
+
+## Interactions
+```{r int_load}
+library(iml)
+library(gridExtra)
+load("data/ES-RF2-preds.RDA")
+```
+
+## Pollinator base interactions
+
+```{r polbase_int}
+grid.arrange(
+  int.polbase %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.polbase.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
+)
+```
+
+## Pollinators margins interactions
+
+```{r polmarg_int}
+grid.arrange(
+  int.polmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.polmarg.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
+)
+```
+
+## N export base interactions
+
+```{r nexpbase_int}
+grid.arrange(
+  int.nexpbase %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.nexpbase.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
+)
+```
+
+## N export margin interactions
+
+```{r nexpmarg_int}
+grid.arrange(
+  int.nexpmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  int.nexpmarg.ann_agr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
+  ncol = 2
[…]

### 2025-03-28 docs/RandomForest/make_shap_importance.R

@@ -1,0 +1,47 @@
+# shapley importance
+library(fastshap)
+library(shapviz)
+
+load("data/ES-RF2.RDA")
+
+pfun <- function(object, newdata) predict(object, data = newdata)$predictions
+
+dat.x <- dat[,13:41]
+
+# pollinators
+ex.polbase <- explain(pollinators_base.rf, X=dat.x, pred_wrapper = pfun,
+                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+ex.polbase.glb <- shapviz(ex.polbase)
+
+ex.polmarg <- explain(pollinators_margin.rf, X=dat.x, pred_wrapper = pfun,
+                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+ex.polmarg.glb <- shapviz(ex.polmarg)
+
+# N export
+ex.nexpbase <- explain(n_export_base.rf, X=dat.x, pred_wrapper = pfun,
+                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+ex.nexpbase.glb <- shapviz(ex.nexpbase)
+
+ex.nexpmarg <- explain(n_export_margin.rf, X=dat.x, pred_wrapper = pfun,
+                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+ex.nexpmarg.glb <- shapviz(ex.nexpmarg)
+
+# sediment export
+ex.sedbase <- explain(sed_export_base.rf, X=dat.x, pred_wrapper = pfun,
+                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+ex.sedbase.glb <- shapviz(ex.sedbase)
+
+ex.sedmarg <- explain(sed_export_margin.rf, X=dat.x, pred_wrapper = pfun,
+                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+ex.sedmarg.glb <- shapviz(ex.sedmarg)
+
+# baseflow
+ex.qbbase <- explain(qb_base.rf, X=dat.x, pred_wrapper = pfun,
+                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+ex.qbbase.glb <- shapviz(ex.qbbase)
+
+ex.qbmarg <- explain(qb_margin.rf, X=dat.x, pred_wrapper = pfun,
+                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+ex.qbmarg.glb <- shapviz(ex.qbmarg)
+
[…]

### 2025-03-28 Paper_analysis.Rmd
update site

@@
 
-```{r shap_pol}
+```{r shap_pol, fig.width=9}
 grid.arrange(
@@
 
+Base pollinator model Shapley dependence plot. Colors are the strongest interactor with that variable.
+
+```{r shap_pol_baseshap, fig.width=9}
+grid.arrange(
+  sv_dependence(ex.polbase.glb, v="ntr_pland"),
+  sv_dependence(ex.polbase.glb, v="ann_agr_pland"), 
+  ncol = 2
+)
+
+```
+
+Margin pollinator model Shapley dependence plot. Colors are the strongest interactor with that variable.
+
+```{r shap_pol_margshap}
+grid.arrange(
+  sv_dependence(ex.polmarg.glb, v="ntr_pland")
+)
+
+```
+
 *N Export*
 
-```{r shap_Nexp}
+```{r shap_nexp, fig.width=9}
 grid.arrange(
@@
 
+```{r shap_nexp_baseshap, fig.width=9}
+grid.arrange(
+  sv_dependence(ex.nexpbase.glb, v="ntr_pland"),
+  sv_dependence(ex.nexpbase.glb, v="ann_agr_pland"), 
+  sv_dependence(ex.nexpbase.glb, v="ntr_ed"), 
+  ncol = 2
+)
+
+```
+
+```{r shap_nexp_margshap, fig.width=9}
+grid.arrange(
+  sv_dependence(ex.nexpmarg.glb, v="ntr_pland"),
+  sv_dependence(ex.nexpmarg.glb, v="ann_agr_pland"), 
+  sv_dependence(ex.nexpmarg.glb, v="ntr_ed"), 
+  ncol = 2
+)
+
+```
+
 *Sediment Export*
 
-```{r shap_sed}
+```{r shap_sed, fig.width=9}
 grid.arrange(
@@
 
+```{r shap_sed_baseshap}
+grid.arrange(
+  sv_dependence(ex.sedbase.glb, v="ann_agr_ed")
+)
+
+```
+
+```{r shap_sed_margshap}
+grid.arrange(
+  sv_dependence(ex.sedmarg.glb, v="ann_agr_ed"),
+  sv_dependence(ex.sedmarg.glb, v="field_frac_mn"),
+  ncol=2
+)
+
+```
+
 *Water recharge*
 
-```{r shap_wb}
+```{r shap_wb, fig.width=9}
 grid.arrange(
@@
 ```
+
+```{r shap_qb_baseshap}
+grid.arrange(
+  sv_dependence(ex.qbbase.glb, v="field_frac_mn")
+)
+
+```
+
+```{r shap_qb_margshap}
+grid.arrange(
+  sv_dependence(ex.qbmarg.glb, v="prn_ann_ratio"),
+  sv_dependence(ex.qbmarg.glb, v="prn_agr_para_mn"),
+  ncol=2
+)
+
+```

### 2025-03-28 Paper_analysis.Rmd
update site again

@@
 
-```{r shap_pol_margshap}
+```{r shap_pol_margshap, fig.width=9}
 grid.arrange(
@@
 
-```{r shap_sed_baseshap}
+```{r shap_sed_baseshap, fig.width=9}
 grid.arrange(
@@
 
-```{r shap_sed_margshap}
+```{r shap_sed_margshap, fig.width=9}
 grid.arrange(
@@
 
-```{r shap_qb_baseshap}
+```{r shap_qb_baseshap, fig.width=9}
 grid.arrange(
@@
 
-```{r shap_qb_margshap}
+```{r shap_qb_margshap, fig.width=9}
 grid.arrange(

### 2025-03-28 Paper_analysis.Rmd
re-update site

--- 2025-03-28 Paper_analysis.Rmd
@@
 
-*Significant metrics count*
-
-```{r}
-
-sigct <- imp.df %>% filter(sig==TRUE)
-
-ggplot(sigct, aes(fct_infreq(metric), fill = type)) +
-  scale_x_discrete(limits=rev) +
-  geom_bar() + coord_flip() + labs(x="metric") + egg::theme_article()
-
-```
-
-```{r impurity_importance, fig.height=16, fig.width=9}
-
-load("data/ES-RF2-impurity.RDA")
-
-library(ggpattern)
-
-metric.descr <- read.csv("data/lsm_description.csv")
-
-imp.imp.df <- bind_rows(
-  pollinators_base.imp.p %>% rownames_to_column("metric") %>% 
-    mutate(ES = "Pollinators", model = "base"),
-  pollinators_margin.imp.p %>% rownames_to_column("metric") %>% 
-    mutate(ES = "Pollinators", model = "margin"),
-  n_export_base.imp.p %>% rownames_to_column("metric") %>% 
-    mutate(ES = "N export", model = "base"),
-  n_export_margin.imp.p %>% rownames_to_column("metric") %>% 
-    mutate(ES = "N export", model = "margin"),
-  sed_export_base.imp.p %>% rownames_to_column("metric") %>% 
-    mutate(ES = "Sed export", model = "base"),
-  sed_export_margin.imp.p %>% rownames_to_column("metric") %>% 
-    mutate(ES = "Sed export", model = "margin"),
-  qb_base.imp.p %>% rownames_to_column("metric") %>% 
-    mutate(ES = "Baseflow", model = "base"),
-  qb_margin.imp.p %>% rownames_to_column("metric") %>% 
-    mutate(ES = "Baseflow", model = "margin")
-) %>%
-  mutate(sig = ifelse(pvalue <0.05, TRUE, FALSE)) %>%
-  mutate(metric2 = tidytext::reorder_within(metric, importance,
-                                           within = list(ES, model))) %>%
-  left_join(metric.descr %>% select(lsm, type), by = c("metric"="lsm")) %>%
-  mutate(type = case_match(
-    metric,
-    "field_pland" ~ "area and edge metric",
-    "field_frac_mn" ~ "shape metric",
-    "field_frac_cv" ~ "shape metric",
-    "prn_ann_ratio" ~ "area and edge metric",
-    .default = type
-  ))
-
-ggplot(imp.imp.df, aes(metric2, importance, fill = type, pattern = sig)) + 
-  geom_bar_pattern(stat="identity",
-                   color = "black", 
-                   pattern_fill = "black",
-                   pattern_angle = 45,
-                   pattern_density = 0.1,
-                   pattern_spacing = 0.025,
-                   pattern_key_scale_factor = 0.6) + 
-  tidytext::scale_x_reordered() +
-  coord_flip() +
-  ggh4x::facet_grid2(ES~model, scales = "free", independent = "all") +
-  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
-  egg::theme_article() +
-  theme(legend.position = "top") + 
-  guides(pattern = guide_legend(override.aes = list(fill = "white")),
-         fill = guide_legend(override.aes = list(pattern = "none")))
-
-```
-
 ## Shapley importance
@@
 
-*Pollinators*
-
-```{r shap_pol}
+```{r shap_pol, fig.width=9}
 grid.arrange(
-  sv_importance(ex.polbase.glb), sv_importance(ex.polmarg.glb),
+  sv_importance(ex.polbase.glb) + egg::theme_article(),
+  sv_importance(ex.polmarg.glb) + egg::theme_article(),
   ncol = 2
@@
 
-*N Export*
+**_Base pollinator model Shapley dependence plots_** The two major contributors. Colors are the strongest interactor with that variable.
 
-```{r shap_Nexp}
+```{r shap_pol_baseshap, fig.width=9}
 grid.arrange(
-  sv_importance(ex.nexpbase.glb), sv_importance(ex.nexpmarg.glb),
+  sv_dependence(ex.polbase.glb, v="ntr_pland") + egg::theme_article(),
+  sv_dependence(ex.polbase.glb, v="ann_agr_pland") + egg::theme_article(), 
   ncol = 2
@@
 
-*Sediment Export*
+**_Margin pollinator model Shapley dependence plot_** Colors are the strongest interactor with that variable.
 
-```{r shap_sed}
+```{r shap_pol_margshap, fig.width=9}
 grid.arrange(
-  sv_importance(ex.sedbase.glb), sv_importance(ex.sedmarg.glb),
+  sv_dependence(ex.polmarg.glb, v="ntr_pland") + egg::theme_article()
+)
+
+```
+
+### N Export
+
+```{r shap_nexp, fig.width=9}
+
+grid.arrange(
+  sv_importance(ex.nexpbase.glb) + egg::theme_article(), 
+  sv_importance(ex.nexpmarg.glb) + egg::theme_article(),
   ncol = 2
@@
 
-*Water recharge*
+**_Base N export shapley dependence plots_** Three major -ish contributors.
 
-```{r shap_wb}
+```{r shap_nexp_baseshap, fig.width=9}
 grid.arrange(
-  sv_importance(ex.qbbase.glb), sv_importance(ex.qbmarg.glb),
+  sv_dependence(ex.nexpbase.glb, v="ntr_pland") + egg::theme_article(),
+  sv_dependence(ex.nexpbase.glb, v="ann_agr_pland") + egg::theme_article(), 
+  sv_dependence(ex.nexpbase.glb, v="ntr_ed") + egg::theme_article(), 
   ncol = 2
@@
 ```
+
+**_Margin N export Shapley dependence plots_** Three major contributors.
+
+```{r shap_nexp_margshap, fig.width=9}
+grid.arrange(
+  sv_dependence(ex.nexpmarg.glb, v="ntr_pland") + egg::theme_article(),
+  sv_dependence(ex.nexpmarg.glb, v="ann_agr_pland") + egg::theme_article(), 
+  sv_dependence(ex.nexpmarg.glb, v="ntr_ed") + egg::theme_article(), 
+  ncol = 2
+)
+
+```
+
+### Sediment Export
+
+```{r shap_sed, fig.width=9}
+grid.arrange(
+  sv_importance(ex.sedbase.glb) + egg::theme_article(),
+  sv_importance(ex.sedmarg.glb) + egg::theme_article(),
+  ncol = 2
+)
+
+```
+
+**_Base sediment export Shapley dependence plots_** One major contributor
+
+```{r shap_sed_baseshap, fig.width=9}
+grid.arrange(
+  sv_dependence(ex.sedbase.glb, v="ann_agr_ed") + egg::theme_article()
+)
+
+```
+
+**_Margin sediment runoff Shapley dependence plots_** One or two major contributors.
+`field_frac_mn` is the strongest interactor with `ann_agr_ed`, and vice versa, while `field_frac_mn` does not stand out as strongly in the permutation results.
+
+```{r shap_sed_margshap, fig.width=9}
+grid.arrange(
+  sv_dependence(ex.sedmarg.glb, v="ann_agr_ed") + egg::theme_article(),
+  sv_dependence(ex.sedmarg.glb, v="field_frac_mn") + egg::theme_article(),
+  ncol=2
+)
+
+```
+
+### Water recharge
+
+```{r shap_wb, fig.width=9}
+grid.arrange(
+  sv_importance(ex.qbbase.glb) + egg::theme_article(),
+  sv_importance(ex.qbmarg.glb) + egg::theme_article(),
+  ncol = 2
+)
+
+```
+
+**_Base water runoff Shapley depednence plot_** One major contributor.
+
+```{r shap_qb_baseshap, fig.width=9}
+grid.arrange(
+  sv_dependence(ex.qbbase.glb, v="field_frac_mn") + egg::theme_article()
+)
+
+```
+
+**_Margin water runoff Shapley dependence plots_** One or two major contributors.
+Differs a little from the permutation results from above.
+In the permutation results, `prn_agr_para_mn` does not have as high importance so that it isn't separated from other variables.
+
+```{r shap_qb_margshap, fig.width=9}
+grid.arrange(
+  sv_dependence(ex.qbmarg.glb, v="prn_ann_ratio") + egg::theme_article(),
+  sv_dependence(ex.qbmarg.glb, v="prn_agr_para_mn") + egg::theme_article(),
+  ncol=2
+)
+
+```

### 2025-04-01 Paper_analysis.Rmd
new results data

@@
 library(tidyverse)
-load("data/ES-RF2.RDA")
+library(iml)
+library(ranger)
+load("data/ES-RF3.RDA")
 
@@
     "field_frac_cv" ~ "shape metric",
-    "prn_ann_ratio" ~ "area and edge metric",
+    "prn_pagr" ~ "area and edge metric",
     .default = type
@@
 
-ggplot(imp.df, aes(metric2, importance, fill = type, pattern = sig)) + 
+# ggplot(imp.df, aes(metric2, importance, fill = type, pattern = sig)) + 
+#   geom_bar_pattern(stat="identity",
+#                    color = "black", 
+#                    pattern_fill = "black",
+#                    pattern_angle = 45,
+#                    pattern_density = 0.1,
+#                    pattern_spacing = 0.025,
+#                    pattern_key_scale_factor = 0.6) + 
+#   tidytext::scale_x_reordered() +
+#   coord_flip() +
+#   ggh4x::facet_grid2(ES~model, scales = "free", independent = "all") +
+#   scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
+#   egg::theme_article() +
+#   theme(legend.position = "top") + 
+#   guides(pattern = guide_legend(override.aes = list(fill = "white")),
+#          fill = guide_legend(override.aes = list(pattern = "none")))
+
+```
+
+```{r importance_shap}
+load("data/ES-RF3-preds.RDA")   # prediction objects from iml package
+load("data/ES-RF3-shapimportance.RDA")
+library(shapviz)
+library(gridExtra)
+
+```
+
+### Pollinators
+
+#### Permutation based importance
+
+```{r pol_permimport, fig.width=9}
+
+ggplot(imp.df %>% filter(ES=="Pollinators"), aes(metric2, importance, fill = type, pattern = sig)) + 
   geom_bar_pattern(stat="identity",
@@
   coord_flip() +
-  ggh4x::facet_grid2(ES~model, scales = "free", independent = "all") +
+  facet_wrap(~model, scales = "free") +
   scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
@@
 
-## Shapley importance
-```{r importance_shap}
-load("data/ES-RF2-shapimportance.RDA")
-library(shapviz)
-library(gridExtra)
-
-```
-
-### Pollinators
+#### Shapley importance
 
@@
 
-**_Base pollinator model Shapley dependence plots_** The two major contributors. Colors are the strongest interactor with that variable.
+#### Base pollinator model Shapley dependence plots
+
+The two major contributors. Colors are the strongest interactor with that variable.
 
@@
 
-**_Margin pollinator model Shapley dependence plot_** Colors are the strongest interactor with that variable.
+**_Compare to PDP and ALE_**
+
+```{r pol_base_ale, fig.width=9}
+
+polbase.ntr_pland.pdp = FeatureEffect$new(pred.pollinators_base, 
+                                          feature = "ntr_pland",
+                                          method = "pdp")
+polbase.ann_agr_pland.pdp = FeatureEffect$new(pred.pollinators_base, 
+                                          feature = "ann_agr_pland",
+                                          method = "pdp")
+
+polbase.ntr_pland.ale = FeatureEffect$new(pred.pollinators_base, 
+                                          feature = "ntr_pland",
+                                          method = "ale")
+polbase.ann_agr_pland.ale = FeatureEffect$new(pred.pollinators_base, 
+                                          feature = "ann_agr_pland",
+                                          method = "ale")
+
+grid.arrange(
+  polbase.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP"),
+  polbase.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE"),
+
+  polbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP"),
+  polbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland ALE"),
+  ncol = 2
+)
+
+```
+
+#### Margin pollinator model Shapley dependence plot
+
+Colors are the strongest interactor with that variable.
 
@@
 grid.arrange(
-  sv_dependence(ex.polmarg.glb, v="ntr_pland") + egg::theme_article()
+  sv_dependence(ex.polmarg.glb, v="ntr_pland") + egg::theme_article(),
+  sv_dependence(ex.polmarg.glb, v="ann_agr_pland") + egg::theme_article()
+)
+
+```
+
+**_Compare to PDP and ALE_**
+
+```{r pol_marg_ale, fig.width=9}
+
+polmarg.ntr_pland.pdp = FeatureEffect$new(pred.pollinators_margin, 
+                                          feature = "ntr_pland",
+                                          method = "pdp")
+polmarg.ann_agr_pland.pdp = FeatureEffect$new(pred.pollinators_margin, 
+                                          feature = "ann_agr_pland",
+                                          method = "pdp")
+
+polmarg.ntr_pland.ale = FeatureEffect$new(pred.pollinators_margin, 
+                                          feature = "ntr_pland",
+                                          method = "ale")
+polmarg.ann_agr_pland.ale = FeatureEffect$new(pred.pollinators_margin, 
+                                          feature = "ann_agr_pland",
+                                          method = "ale")
+
+grid.arrange(
+  polmarg.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP"),
+  polmarg.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE"),
+
+  polmarg.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP"),
+  polmarg.ann_agr_pland.ale$plot() + ggtitle("ann_agr_pland ALE"),
+  ncol = 2
 )
@@
 ### N Export
+
+### Permutation based importance
+
+```{r nexp_permimport, fig.width=9}
+
+ggplot(imp.df %>% filter(ES=="N export"), aes(metric2, importance, fill = type, pattern = sig)) + 
+  geom_bar_pattern(stat="identity",
+                   color = "black", 
+                   pattern_fill = "black",
+                   pattern_angle = 45,
+                   pattern_density = 0.1,
+                   pattern_spacing = 0.025,
+                   pattern_key_scale_factor = 0.6) + 
+  tidytext::scale_x_reordered() +
+  coord_flip() +
+  facet_wrap(~model, scales = "free") +
+  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
+  egg::theme_article() +
+  theme(legend.position = "top") + 
+  guides(pattern = guide_legend(override.aes = list(fill = "white")),
+         fill = guide_legend(override.aes = list(pattern = "none")))
+
+```
+
+#### Shapley importance
 
@@
 
-**_Base N export shapley dependence plots_** Three major -ish contributors.
+#### Base N export shapley dependence plots
+
+Three major -ish contributors.
 
 ```{r shap_nexp_baseshap, fig.width=9}
+
 grid.arrange(
@@
 
-**_Margin N export Shapley dependence plots_** Three major contributors.
+**_Compare to PDP and ALE_**
+
+```{r nexp_base_ale, fig.width=9}
+
+nexpbase.ntr_pland.pdp = FeatureEffect$new(pred.n_export_base, 
+                                          feature = "ntr_pland",
+                                          method = "pdp")
+nexpbase.ann_agr_pland.pdp = FeatureEffect$new(pred.n_export_base, 
+                                          feature = "ann_agr_pland",
+                                          method = "pdp")
+nexpbase.ntr_ed.pdp = FeatureEffect$new(pred.n_export_base, 
+                                          feature = "ntr_ed",
+                                          method = "pdp")
+
+nexpbase.ntr_pland.ale = FeatureEffect$new(pred.n_export_base, 
+                                          feature = "ntr_pland",
+                                          method = "ale")
+nexpbase.ann_agr_pland.ale = FeatureEffect$new(pred.n_export_base, 
+                                          feature = "ann_agr_pland",
+                                          method = "ale")
+nexpbase.ntr_ed.ale = FeatureEffect$new(pred.n_export_base, 
+                                          feature = "ntr_ed",
+                                          method = "ale")
+
+grid.arrange(
+  nexpbase.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP"),
+  nexpbase.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE"),
+
+  nexpbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP"),
+  nexpbase.ann_agr_pland.ale$plot() + ggtitle("ann_agr_pland ALE"),
+
+  nexpbase.ntr_ed.pdp$plot() + ggtitle("ntr_ed PDP"),
+  nexpbase.ntr_ed.ale$plot() + ggtitle("ntr_ed ALE"),
+  ncol = 2
+)
+
+```
+
+#### Margin N export Shapley dependence plots
+
+Three major contributors.
 
@@
 
+**_Compare to PDP and ALE_**
+
+```{r nexp_marg_ale, fig.width=9}
+
+nexpmarg.ntr_pland.pdp = FeatureEffect$new(pred.n_export_margin, 
+                                          feature = "ntr_pland",
+                                          method = "pdp")
+nexpmarg.ann_agr_pland.pdp = FeatureEffect$new(pred.n_export_margin, 
+                                          feature = "ann_agr_pland",
+                                          method = "pdp")
+nexpmarg.ntr_ed.pdp = FeatureEffect$new(pred.n_export_margin, 
+                                          feature = "ntr_ed",
+                                          method = "pdp")
+
+nexpmarg.ntr_pland.ale = FeatureEffect$new(pred.n_export_margin, 
+                                          feature = "ntr_pland",
+                                          method = "ale")
+nexpmarg.ann_agr_pland.ale = FeatureEffect$new(pred.n_export_margin, 
+                                          feature = "ann_agr_pland",
+                                          method = "ale")
+nexpmarg.ntr_ed.ale = FeatureEffect$new(pred.n_export_margin, 
+                                          feature = "ntr_ed",
+                                          method = "ale")
+
+grid.arrange(
+  nexpmarg.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP"),
+  nexpmarg.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE"),
+
+  nexpmarg.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP"),
+  nexpmarg.ann_agr_pland.ale$plot() + ggtitle("ann_agr_pland ALE"),
+
+  nexpmarg.ntr_ed.pdp$plot() + ggtitle("ntr_ed PDP"),
+  nexpmarg.ntr_ed.ale$plot() + ggtitle("ntr_ed ALE"),
+  ncol = 2
+)
+
+```
+
 ### Sediment Export
+
+### Permutation based importance
+
+```{r sed_permimport, fig.width=9}
+
+ggplot(imp.df %>% filter(ES=="Sed export"), aes(metric2, importance, fill = type, pattern = sig)) + 
+  geom_bar_pattern(stat="identity",
+                   color = "black", 
+                   pattern_fill = "black",
+                   pattern_angle = 45,
+                   pattern_density = 0.1,
+                   pattern_spacing = 0.025,
+                   pattern_key_scale_factor = 0.6) + 
+  tidytext::scale_x_reordered() +
+  coord_flip() +
+  facet_wrap(~model, scales = "free") +
+  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
+  egg::theme_article() +
+  theme(legend.position = "top") + 
+  guides(pattern = guide_legend(override.aes = list(fill = "white")),
+         fill = guide_legend(override.aes = list(pattern = "none")))
+
+```
+
+#### Shapley importance
 
@@
 
-**_Base sediment export Shapley dependence plots_** One major contributor
+#### Base sediment export Shapley dependence plots
+
+One major contributor
 
@@
 
-**_Margin sediment runoff Shapley dependence plots_** One or two major contributors.
-`field_frac_mn` is the strongest interactor with `ann_agr_ed`, and vice versa, while `field_frac_mn` does not stand out as strongly in the permutation results.
+**_Compare to PDP and ALE_**
+
+```{r sedexp_base_ale, fig.width=9}
+
+sedbase.ann_agr_ed.pdp = FeatureEffect$new(pred.sed_export_base, 
+                                          feature = "ann_agr_ed",
+                                          method = "pdp")
+
+sedbase.ann_agr_ed.ale = FeatureEffect$new(pred.sed_export_base, 
+                                          feature = "ann_agr_ed",
+                                          method = "ale")
+
+grid.arrange(
+  sedbase.ann_agr_ed.pdp$plot() + ggtitle("ann_agr_ed PDP") +
+    egg::theme_article(),
+  sedbase.ann_agr_ed.ale$plot() + ggtitle("ann_agr_ed ALE") +
+    egg::theme_article(),
+
+  ncol = 2
+)
+
+```
+
+#### Margin sediment runoff Shapley dependence plots
+
+One or two major contributors. `field_frac_mn` is the strongest interactor with `ann_agr_ed`, and vice versa, while `field_frac_mn` does not stand out as strongly in the permutation results.
 
@@
 
+**_Compare to PDP and ALE_**
+
+```{r sedexp_marg_ale, fig.width=9}
+
+sedmarg.ann_agr_ed.pdp = FeatureEffect$new(pred.sed_export_margin, 
+                                          feature = "ann_agr_ed",
+                                          method = "pdp")
+sedmarg.field_frac_mn.pdp = FeatureEffect$new(pred.sed_export_margin, 
+                                          feature = "field_frac_mn",
+                                          method = "pdp")
+
+sedmarg.ann_agr_ed.ale = FeatureEffect$new(pred.sed_export_margin, 
+                                          feature = "ann_agr_ed",
+                                          method = "ale")
+sedmarg.field_frac_mn.ale = FeatureEffect$new(pred.sed_export_margin, 
+                                          feature = "field_frac_mn",
+                                          method = "ale")
+
+grid.arrange(
+  sedmarg.ann_agr_ed.pdp$plot() + ggtitle("ann_agr_ed PDP") +
+    egg::theme_article(),
+  sedmarg.ann_agr_ed.ale$plot() + ggtitle("ann_agr_ed ALE") +
+    egg::theme_article(),
+
+  sedmarg.field_frac_mn.pdp$plot() + ggtitle("field_frac_mn PDP") +
+    egg::theme_article(),
+  sedmarg.field_frac_mn.ale$plot() + ggtitle("field_frac_mn ALE") +
+    egg::theme_article(),
+
+  ncol = 2
+)
+
+```
+
 ### Water recharge
 
-```{r shap_wb, fig.width=9}
+### Permutation based importance
+
+```{r qb_permimport, fig.width=9}
+
+ggplot(imp.df %>% filter(ES=="Baseflow"), aes(metric2, importance, fill = type, pattern = sig)) + 
+  geom_bar_pattern(stat="identity",
+                   color = "black", 
+                   pattern_fill = "black",
+                   pattern_angle = 45,
+                   pattern_density = 0.1,
+                   pattern_spacing = 0.025,
+                   pattern_key_scale_factor = 0.6) + 
+  tidytext::scale_x_reordered() +
+  coord_flip() +
+  facet_wrap(~model, scales = "free") +
+  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
+  egg::theme_article() +
+  theme(legend.position = "top") + 
+  guides(pattern = guide_legend(override.aes = list(fill = "white")),
+         fill = guide_legend(override.aes = list(pattern = "none")))
+
+```
+
+#### Shapley importance
+
+```{r shap_qb, fig.width=9}
 grid.arrange(
@@
 
-**_Base water runoff Shapley depednence plot_** One major contributor.
+#### Base water runoff Shapley depednence plot
+
+One major contributor.
 
@@
 
-**_Margin water runoff Shapley dependence plots_** One or two major contributors.
+**_Compare to PDP and ALE_**
+
+```{r qb_base_ale, fig.width=9}
+
+qbbase.field_frac_mn.pdp = FeatureEffect$new(pred.qb_base, 
+                                             feature = "field_frac_mn",
+                                             method = "pdp")
+
+sedbase.field_frac_mn.ale = FeatureEffect$new(pred.qb_base, 
+                                              feature = "field_frac_mn",
+                                              method = "ale")
+
+grid.arrange(
+  qbbase.field_frac_mn.pdp$plot() + ggtitle("field_frac_mn PDP") +
+    egg::theme_article(),
+  sedbase.field_frac_mn.ale$plot() + ggtitle("field_frac_mn ALE") +
+    egg::theme_article(),
+
+  ncol = 2
+)
+
+```
+
+#### Margin water runoff Shapley dependence plots
+
+One or two major contributors.
 Differs a little from the permutation results from above.
@@
 grid.arrange(
-  sv_dependence(ex.qbmarg.glb, v="prn_ann_ratio") + egg::theme_article(),
+  sv_dependence(ex.qbmarg.glb, v="ann_agr_cohesion") + egg::theme_article(),
+  sv_dependence(ex.qbmarg.glb, v="prn_pagr") + egg::theme_article(),
   sv_dependence(ex.qbmarg.glb, v="prn_agr_para_mn") + egg::theme_article(),
@@
 ```
+
+**_Compare to PDP and ALE_**
+
+```{r qb_marg_ale, fig.width=9}
+
+qbmarg.ann_agr_cohesion.pdp = FeatureEffect$new(pred.qb_margin, 
+                                             feature = "ann_agr_cohesion",
+                                             method = "pdp")
+qbmarg.prn_pagr.pdp = FeatureEffect$new(pred.qb_margin, 
+                                             feature = "prn_pagr",
+                                             method = "pdp")
+qbmarg.prn_agr_para_mn.pdp = FeatureEffect$new(pred.qb_margin, 
+                                             feature = "prn_agr_para_mn",
+                                             method = "pdp")
+
+qbmarg.ann_agr_cohesion.ale = FeatureEffect$new(pred.qb_margin, 
+                                              feature = "ann_agr_cohesion",
+                                              method = "ale")
+qbmarg.prn_pagr.ale = FeatureEffect$new(pred.qb_margin, 
+                                              feature = "prn_pagr",
+                                              method = "ale")
+qbmarg.prn_agr_para_mn.ale = FeatureEffect$new(pred.qb_margin, 
+                                              feature = "prn_agr_para_mn",
+                                              method = "ale")
+
+grid.arrange(
+  qbmarg.ann_agr_cohesion.pdp$plot() + ggtitle("ann_agr_cohesion PDP") +
+    egg::theme_article(),
+  qbmarg.ann_agr_cohesion.ale$plot() + ggtitle("ann_agr_cohesion ALE") +
+    egg::theme_article(),
+
+  qbmarg.prn_pagr.pdp$plot() + ggtitle("prn_pagr PDP") +
+    egg::theme_article(),
+  qbmarg.prn_pagr.ale$plot() + ggtitle("prn_pagr ALE") +
+    egg::theme_article(),
+
+  qbmarg.prn_agr_para_mn.pdp$plot() + ggtitle("prn_agr_para_mn PDP") +
+    egg::theme_article(),
+  qbmarg.prn_agr_para_mn.ale$plot() + ggtitle("prn_agr_para_mn ALE") +
+    egg::theme_article(),
+
+  ncol = 2
+)
+
+```
+
+### SHAP force plots
+
+I am plotting the combined SHAP force values for each watershed together.
+The important features identified above are colored while the rest are left grey.
+To plot watersheds on a common axis and visualize tradeoffs, I used PCA to identify the two major axes.
+The first PC axis for both baseline and margin results show a tradeoff between N export (and improvement in N export) and pollinator abundance (pollinator improvement).
+The second PC shows sediment export and baseflow changing together while with margin improvement are in the same direction.
+
+```{r paircor}
+selectedvar <- c("ntr_pland", "ann_agr_pland", "ntr_ed", 
+                 "field_frac_mn", "prn_pagr", "prn_agr_para_mn",
+                 "ann_agr_ed", "ann_agr_cohesion"
+                 )
+
+# selectedvar.base <- c("ntr_pland", "ann_agr_pland",  "ntr_ed", 
+#                       "ann_agr_ed", "field_frac_mn")
+# 
+# selectedvar.margin <- c("ntr_pland", "ann_agr_pland",  "ntr_ed",
+#                         "ann_agr_ed", "field_frac_mn",
+#                         "prn_ann_ratio", "prn_agr_para_mn")
+
+# psych::pairs.panels(dat.selectedvar)
+
+```
+
+```{r force_fcns}
+source("R/utils-shap.r")
+focus.vars <- function(x.features){
+  dplyr::case_match(
+    x.features,
+    "ntr_pland" ~ "Natural land cover",
+    "ann_agr_pland" ~ "Annual ag. land cover",
+    "field_frac_mn" ~ "Field fractal dimension",
+    "prn_pagr" ~ "Agriculture % perennial",
+    "prn_agr_para_mn" ~ "Perennial ag. perimeter:area",
+    "ann_agr_ed" ~ "Annual ag. edge density",
+    "ann_agr_cohesion" ~ "Annual ag. cohesion",
+    .default = "Other features"
+  )
+}
+
+feature.colors <- c(
+  "Natural land cover" = "#33a02c",
+  "Annual ag. land cover" = "#fdbf6f",
+  "Field fractal dimension" = "#56B4E9",
+  "Agriculture % perennial" = "#CC79A7",
+  "Perennial ag. perimeter:area" = "#D55E00",
+  "Annual ag. edge density" = "#F0E442",
+  "Annual ag. cohesion" = "#0072B2",
+  "Other features" = "gray80"
+)
+
+```
+
+```{r make_pca_baseoutcomes_forces, include = FALSE}
+
+base_esoutcome <- dat %>% select(pollinators_base:sed_export_base) %>%
+  mutate(sed_export_base = sed_export_base,
+         n_export_base = n_export_base)
+
+base_pca <- prcomp(base_esoutcome,
+                   scale=TRUE, center=TRUE)
+
+ntr_class <- cut(dat$ntr_pland, breaks = c(0, 10, 20, 30 , 50, 100),
+                 labels = c("5-10%", "10-20%", "20-30%", "30-50%", "50-90%"))
+
+biplot_base_pca <- ggbiplot::ggbiplot(base_pca, groups = ntr_class) + 
+  expand_limits(x=c(-2.5,2.5)) +
+  scale_color_manual(name="Natural land cover",
+                     values = c("#d7191c", "#fdae61", "#ffffbf", "#abdda4", "#2b83ba")) +
+  egg::theme_article() + 
+  theme(legend.position = "inside", legend.position.inside = c(0.2,0.12))
+
+biplot_base_arrows <- biplot_base_pca$layers[[3]]$data
+biplot_base_arrows$varname <- c("Pollinators", "Baseflow", "N export", "Sediment export")
+
+# biplot_base_pca$layers[[1]]$aes_params$colour = "gray"
+biplot_base_pca$layers[[1]]$aes_params$size = 1
+biplot_base_pca$layers[[3]] <- geom_label(aes(x = xvar, y = yvar, label = varname,
+                                  angle = angle, hjust = hjust), 
+                              label.size = NA, size = 3,
+                              data = biplot_base_arrows, 
+                              fill = '#dddddd80')
+biplot_base_pca$layers <- cbind(biplot_base_pca$layers[[3]], biplot_base_pca$layers)
+
+psych::pairs.panels(dat%>% select(pollinators_base:sed_export_base))
+
+polbase_pc1 <- makeforcedata.orderby(ex.polbase$shapley_values,
+                                     predict(base_pca)[,"PC1"],
+                               rename.fn = focus.vars, other.var = "Other features")
+
+nexpbase_pc1 <- makeforcedata.orderby(ex.nexpbase$shapley_values,
+                                  predict(base_pca)[,"PC1"],
+                               rename.fn = focus.vars, other.var = "Other features")
+
+sedbase_pc2 <- makeforcedata.orderby(ex.sedbase$shapley_values,
+                                     predict(base_pca)[,"PC2"],
+                               rename.fn = focus.vars, other.var = "Other features")
+
+qbbase_pc2 <- makeforcedata.orderby(ex.qbbase$shapley_values,
+                                    predict(base_pca)[,"PC2"],
+                               rename.fn = focus.vars, other.var = "Other features")
+
+polbase_pc1_force <- ggplot(polbase_pc1, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, guide = guide_legend(reverse=TRUE)) +
+  xlab("PC1") + ylab("SHAP values") + ggtitle("Baseline pollinator abundance") +
+  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),
+        legend.position = "none")
+
+nexpbase_pc1_force <- ggplot(nexpbase_pc1, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, guide = guide_legend(reverse=TRUE)) +
+  xlab("PC1") + ylab("SHAP values") + ggtitle("Baseline N export") +
+  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),
+        legend.position = "none")
+
+sedbase_pc2_force <- ggplot(sedbase_pc2, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, guide = guide_legend(reverse=TRUE)) +
+  xlab("PC2") + ylab("SHAP values") + ggtitle("Baseline sediment export") +
+  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),
+        legend.position = "none")
+
+qbbase_pc2_force <- ggplot(qbbase_pc2, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, guide = guide_legend(reverse=TRUE)) +
+  xlab("PC2") + ylab("SHAP values") + ggtitle("Baseline baseflow") +
+  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),
+        legend.position = "none")
+
+base_forcestackPC1 <- cowplot::plot_grid(polbase_pc1_force, nexpbase_pc1_force, align='v',
+                                         labels = c("B","C"), ncol=1) %>% 
+  ggplotGrob()
+base_forcestackPC2 <- cowplot::plot_grid(sedbase_pc2_force, qbbase_pc2_force, align='v',
+                                         labels = c("D","E"), ncol=1) %>%
+  ggplotGrob()
+
+# make a legend
+
+baselegend_dummy <- ggplotGrob(ggplot(polbase_pc1, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, 
+                    guide = guide_legend(reverse=TRUE)) +
+  theme(legend.position = "bottom", legend.title = element_blank()))$grobs
+
+base_legend <- baselegend_dummy[[which(sapply(baselegend_dummy, function(x)
+      x$name) == "guide-box")]]
+
+base_plots <- cowplot::plot_grid(base_forcestackPC1, base_forcestackPC2, ncol = 2)
+
+base_plots_legend <- grid.arrange(base_plots, base_legend, 
+                                  layout_matrix = 
+                                    rbind(
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(2,2,2,2,2,2)
+                                    ))
+
+base_plots_all <- cowplot::plot_grid(biplot_base_pca, base_plots_legend, 
+                                     labels = c("A",""), rel_widths = c(1,2.2))
+
+# ggsave("Figs/base_pcashap.png", base_plots_all, width = 13, height = 5.5, bg = "white", dpi = 300)
+```
+
+```{r plot_pca_baseoutcomes, fig.width = 12, fig.height = 5}
+base_plots_all
+```
+
+```{r make_pca_marginoutcomes, include=FALSE}
+
+marg_esoutcome <- dat %>% select(delta_pollinators_max.margin:delta_sed_max.margin)
+
+marg_pca <- prcomp(marg_esoutcome,
+                   scale=TRUE, center=TRUE)
+
+ntr_class <- cut(dat$ntr_pland, breaks = c(0, 10, 20, 30 , 50, 100),
+                 labels = c("5-10%", "10-20%", "20-30%", "30-50%", "50-90%"))
+
+biplot_marg_pca <- ggbiplot::ggbiplot(marg_pca, groups = ntr_class) + 
+  expand_limits(x=c(-3,3)) +
+  scale_color_manual(name="Natural land cover",
+                     values = c("#d7191c", "#fdae61", "#ffffbf", "#abdda4", "#2b83ba")) +
+  guides(color = guide_legend(label.position = "left")) +
+  egg::theme_article() + 
+  theme(legend.position = "inside", legend.position.inside = c(0.8,0.85), legend.title.align=1)
+
+biplot_marg_arrows <- biplot_marg_pca$layers[[3]]$data
+biplot_marg_arrows$varname <- c("\u0394 Pollinators", "\u0394 Baseflow", "N reduction", "Sediment reduction")
+
+biplot_marg_pca$layers[[1]]$aes_params$size = 1
+biplot_marg_pca$layers[[3]] <- geom_label(aes(x = xvar, y = yvar, label = varname,
+                                  angle = angle, hjust = hjust), 
+                              label.size = NA, size = 3,
+                              data = biplot_marg_arrows, 
+                              fill = '#dddddd80')
+biplot_marg_pca$layers <- cbind(biplot_marg_pca$layers[[3]], biplot_marg_pca$layers)
+
+polmarg_pc1 <- makeforcedata.orderby(ex.polmarg$shapley_values,
+                                     predict(marg_pca)[,"PC1"],
+                               rename.fn = focus.vars, other.var = "Other features")
+
+nexpmarg_pc1 <- makeforcedata.orderby(ex.nexpmarg$shapley_values,
+                                  predict(marg_pca)[,"PC1"],
+                               rename.fn = focus.vars, other.var = "Other features")
+
+sedmarg_pc2 <- makeforcedata.orderby(ex.sedmarg$shapley_values,
+                                     predict(marg_pca)[,"PC2"],
+                               rename.fn = focus.vars, other.var = "Other features")
+
+qbmarg_pc2 <- makeforcedata.orderby(ex.qbmarg$shapley_values,
+                                    predict(marg_pca)[,"PC2"],
+                               rename.fn = focus.vars, other.var = "Other features")
+
+polmarg_pc1_force <- ggplot(polmarg_pc1, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, guide = guide_legend(reverse=TRUE)) +
+  xlab("PC1") + ylab("SHAP values") + ggtitle("Change in pollinator abundance") +
+  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),
+        legend.position = "none")
+
+nexpmarg_pc1_force <- ggplot(nexpmarg_pc1, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, guide = guide_legend(reverse=TRUE)) +
+  xlab("PC1") + ylab("SHAP values") + ggtitle("N export reduction") +
+  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),
+        legend.position = "none")
+
+sedmarg_pc2_force <- ggplot(sedmarg_pc2, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, guide = guide_legend(reverse=TRUE)) +
+  xlab("PC2") + ylab("SHAP values") + ggtitle("Sediment export reduction") +
+  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),
+        legend.position = "none")
+
+qbmarg_pc2_force <- ggplot(qbmarg_pc2, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, guide = guide_legend(reverse=TRUE)) +
+  xlab("PC2") + ylab("SHAP values") + ggtitle("Baseflow change") +
+  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),
+        legend.position = "none")
+
+marg_forcestackPC1 <- cowplot::plot_grid(polmarg_pc1_force, nexpmarg_pc1_force, align='v',
+                                         labels = c("B","C"), ncol=1) %>% 
+  ggplotGrob()
+marg_forcestackPC2 <- cowplot::plot_grid(sedmarg_pc2_force, qbmarg_pc2_force, align='v',
+                                         labels = c("D","E"), ncol=1) %>%
+  ggplotGrob()
+
+# make a legend
+
+marglegend_dummy <- ggplotGrob(ggplot(polmarg_pc1, aes(rowid, value, fill = feature2)) +
+  geom_col(width = 1) + egg::theme_article() +
+  scale_fill_manual(values = feature.colors, 
+                    guide = guide_legend(reverse=TRUE)) +
+  theme(legend.position = "bottom", legend.title = element_blank()))$grobs
+
+marg_legend <- marglegend_dummy[[which(sapply(marglegend_dummy, function(x)
+      x$name) == "guide-box")]]
+
+marg_plots <- cowplot::plot_grid(marg_forcestackPC1, marg_forcestackPC2, ncol = 2)
+
+marg_plots_legend <- grid.arrange(marg_plots, marg_legend, 
+                                  layout_matrix = 
+                                    rbind(
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(1,1,1,1,1,1),
+                                      c(2,2,2,2,2,2)
+                                    ))
+
+marg_plots_all <- cowplot::plot_grid(biplot_marg_pca, marg_plots_legend, 
+                                     labels = c("A",""), rel_widths = c(1,2.2))
+
+# ggsave("Figs/marg_pcashap.png", marg_plots_all, width = 13, height = 5.5, bg = "white", dpi = 300)
+
+```
+
+```{r plot_pca_margoutcomes, fig.width = 12, fig.height = 5}
+marg_plots_all
+```

### 2025-04-01 docs/randomforest_kl.Qmd
new results data

--- 2025-03-26 randomforest_kl.Qmd
@@
     # wfs0 <- readRDS("data/old/wfs_lsm.rds")
-      wfs <- readRDS("data/allresults_20250308.rds")
+      wfs <- readRDS("data/old/allresults_20250308.rds")
 
@@
 
-    if(file.exists("data/ES-RF.RDA")) load("data/ES-RF.RDA")
+    if(file.exists("data/old/ES-RF.RDA")) load("data/old/ES-RF.RDA")
 

### 2025-04-01 FigureMap/Fig1_lcplot.R

@@ -1,0 +1,48 @@
+library(tidyverse)
+
+hu_lc <- read.csv("data-raw/hu_rawlc.csv") %>%
+  mutate(huc12 = as.character(paste0("0",huc12)))
+lc.agg <- readRDS("data/aglu_codes.rds")
+
+hu_lclong <- hu_lc %>% select(huc12, areasqkm, X1:X52) %>% 
+  pivot_longer(cols=X1:X52, names_to = "Value", values_to = "coverage") %>%
+  mutate(Value = as.numeric(gsub("X","",Value))) %>%
+  mutate(coverage = replace_na(coverage,0)) %>% 
+  left_join(lc.agg)
+
+hu_aglu <- hu_lclong %>% 
+  group_by(huc12, landuse_group, areasqkm) %>%
+  summarize(pixel.coverage = sum(coverage), .groups = "drop") %>%
+  mutate(percent.coverage = 100*(pixel.coverage*900/(1000^2))/areasqkm) %>%
+  group_by(huc12) %>% 
+  group_modify(~ .x %>% cbind(natural = .x[.x$landuse_group=="ntr",]$percent.coverage)) %>%
+  ungroup() %>%
+  mutate(huc12 = fct_reorder(as.factor(huc12), natural)) %>% 
+  mutate(landuse_group = fct_relevel(landuse_group, 
+                                     c("ann_agr", "prn_agr","wtr","dvl","non","ntr")))
+
+# check they add up to 100. Pretty much, maybe some rounding error from areasqkm
+hu_aglu_check <- hu_aglu %>% group_by(huc12) %>%
+  summarize(total.percent = sum(percent.coverage), .groups = "drop")
+
+# land cover colors
+lccols <- c("ann_agr" = "#fdbf6f", "dvl" = "#bfa4a4", "ntr" = "#33a02c", 
+            "wtr" = "#a6cee3", "prn_agr" = "#b2df8a", "non" = "black")
+
+ggplot(hu_aglu, aes(huc12, percent.coverage, fill = landuse_group)) +
+  geom_col(width = 1) + scale_fill_manual(values = lccols) +
+  egg::theme_presentation() + ylab("Cumulative \n% coverage") + 
+  xlab("Watersheds ordered by increasing % natural land cover") +
+  geom_vline(xintercept = 
+               which(levels(hu_aglu$huc12) %in% c("071000010602", "071401061005", "020802070704"))) +
+  scale_y_continuous(expand = c(0, 0)) +
+  theme(
[…]

### 2025-04-01 FigureMap/aglu_colormap.R

@@ -1,0 +1,28 @@
+library(tidyverse)
+
+aglu_colortable <- read.table("FigureMap/cdlhabitat_colortable.txt", sep = ",")
+lc.agg <- readRDS("data/aglu_codes.rds")
+
+new_colors <- data.frame(
+  landuse_group = c("ann_agr", "dvl", "non","ntr", "prn_agr", "wtr"),
+  r = c(253,191,0,51,178,166),
+  g = c(191,164,0,160,223,206),
+  b = c(111,164,0,44,138,227),
+  a = c(255,255,255,255,255,255)
+)
+
+aglu_colortable_new <- aglu_colortable %>% 
+  left_join(lc.agg, by = c("V1" = "Value")) %>% 
+  left_join(new_colors, by = "landuse_group") %>%
+  mutate(across(r:a, ~replace_na(.x,0))) %>%
+  select(V1, r,g,b,a, landuse_group)
+
+write.table(aglu_colortable_new, file = "FigureMap/aglu_colortable.txt",
+            row.names = FALSE, col.names = FALSE, sep = ",")
+
+# add the following manually to the head of the colormap for QGIS:
+`
+# QGIS Generated Color Map Export File
+INTERPOLATION:INTERPOLATED
+`
+

### 2025-04-01 R/utils-shap.R

@@ -1,0 +1,67 @@
+makeforcedata <- function(shaptab, order.col){
+  shaptab <- as.data.frame(shaptab)
+  features <- names(shaptab)
+  shaptab <- shaptab |> 
+    tibble::rownames_to_column("rowid") |>
+    tidyr::pivot_longer(cols=features, names_to = "feature", values_to = "value") |>
+    dplyr::group_by(rowid) |>
+    group_modify(~ .x %>% cbind(ordercol = .x[.x$feature==order.col,]$value)) |>
+    ungroup() |>
+    mutate(rowid = fct_reorder(as.factor(rowid), ordercol))
+}
+
+makeforcedata.hclust <- function(shaptab, ...){
+  shaptab <- as.data.frame(shaptab)
+  features <- names(shaptab)
+  
+  shap.hclust <- hclust(dist(shaptab), ...)
+  
+  shaptab <- shaptab |> 
+    tibble::rownames_to_column("rowid") |>
+    dplyr::bind_cols(ordercol = shap.hclust$order) |>
+    tidyr::pivot_longer(cols=features, names_to = "feature", values_to = "value") |>
+    mutate(rowid = fct_reorder(as.factor(rowid), ordercol))
+}
+
+makeforcedata.predict <- function(shaptab, predictions, ...){
+  shaptab <- as.data.frame(shaptab)
+  features <- names(shaptab)
+  
+  shaptab <- shaptab |> 
+    tibble::rownames_to_column("rowid") |>
+    dplyr::bind_cols(ordercol = predictions) |>
+    tidyr::pivot_longer(cols=features, names_to = "feature", values_to = "value") |>
+    mutate(rowid = fct_reorder(as.factor(rowid), ordercol))
+}
+
+makeforcedata.orderby <- function(shaptab, orderby, rename.fn, other.var){
+  shaptab <- as.data.frame(shaptab)
+  features <- names(shaptab)
+  
+  shaptab <- shaptab |> 
+    tibble::rownames_to_column("rowid") |>
+    dplyr::bind_cols(ordercol = orderby) |>
+    tidyr::pivot_longer(cols=features, names_to = "feature", values_to = "value") |>
+    mutate(rowid = fct_reorder(as.factor(rowid), ordercol))
+  
+  if(!missing(rename.fn)){
+    if(is.function(rename.fn)){
+      shaptab <- shaptab |>
+        mutate(feature2 = rename.fn(feature)) |>
+        # order feature2 by absolute magnitude of total shap values
[…]

### 2025-04-01 RandomForest/make_rf.R

@@ -3,11 +3,11 @@
 library(ranger)
 options(ranger.num.threads = 8)
 
-wfs <- readRDS("data/allresults_20250308.rds")
+wfs <- readRDS("data/allresults_20250401.rds")
 
 names(wfs)
 
-dat.predictors <-  wfs[, c(81:107,120,123)]
+dat.predictors <-  wfs[, c(81:111,125,126)]
 
 dat.response <- wfs[, c(2:5, 10:13, 20:23)]
 
@@ -86,7 +86,7 @@ qb_margin.imp.p <- data.frame(importance_pvalues(qb_margin.imp.rf, method = "alt
 
 
 # save the permutation outputs
-save(list = c("dat",ls(pattern = "^[npqs]")), file = "data/ES-RF2.RDA")
+save(list = c("dat",ls(pattern = "^[npqs]")), file = "data/ES-RF3.RDA")
 
 # save just the impurity-based outputs
-save(list = ls(pattern = "imp"), file = "data/ES-RF2-impurity.RDA")
+# save(list = ls(pattern = "imp"), file = "data/ES-RF2-impurity.RDA")

### 2025-04-01 RandomForest/make_rf_importance.R

@@ -1,7 +1,7 @@
 # make importance values
 library(iml)
-load("data/ES-RF2.RDA")
-load("data/ES-RF2-preds.RDA")
+load("data/ES-RF3.RDA")
+load("data/ES-RF3-preds.RDA")
 
 # pollinators
 imp.polbase <- FeatureImp$new(pred.pollinators_base, loss = "mae")

### 2025-04-01 RandomForest/make_rf_interactions.R

@@ -1,7 +1,7 @@
 library(ranger)
 library(iml)
 library(future)
-load("data/ES-RF2.RDA")
+load("data/ES-RF3.RDA")
 pfun <- function(object, newdata) predict(object, data = newdata)$predictions
 
 future::plan(multisession, workers = 8)
@@ -102,92 +102,92 @@ pred.qb_margin <- Predictor$new(
 int.qbmarg  <- Interaction$new(pred.qb_margin)
 int.qbmarg.field_frac_cv <- Interaction$new(pred.qb_margin, feature = "field_frac_cv")
 
-save(list = ls(pattern = "^int|^pred"), file = "data/ES-RF2-preds.RDA")
+save(list = ls(pattern = "^int|^pred"), file = "data/ES-RF3-preds.RDA")
 
-## Interactions
-```{r int_load}
-library(iml)
-library(gridExtra)
-load("data/ES-RF2-preds.RDA")
-```
-
-## Pollinator base interactions
-
-```{r polbase_int}
-grid.arrange(
-  int.polbase %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  int.polbase.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  ncol = 2
-)
-```
-
-## Pollinators margins interactions
-
-```{r polmarg_int}
-grid.arrange(
-  int.polmarg %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  int.polmarg.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  ncol = 2
-)
-```
-
-## N export base interactions
-
-```{r nexpbase_int}
-grid.arrange(
-  int.nexpbase %>% plot() + ggtitle("Interaction strength") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  int.nexpbase.ntr_pland %>% plot() + ggtitle("Pairwise interactions with strongest") + egg::theme_article() + theme(plot.title.position = "plot", axis.title.y = element_blank()),
-  ncol = 2
-)
-```
-
-## N export margin interactions
-
-```{r nexpmarg_int}
[…]

### 2025-04-01 data-raw/allresults_20250401.R

@@ -1,0 +1,110 @@
+library(tidyverse)
+
+datadir <- "data-raw/out_20250308/"
+
+outfiles <- list.files(datadir)
+
+base_out <- unlist(
+  lapply(paste0(datadir,outfiles[grepl("base*", outfiles)]), readRDS),
+  recursive = FALSE) %>%
+  bind_rows(.id = "hu") %>%
+  mutate(scenario="baseline")
+
+max_out <- unlist(
+  lapply(paste0(datadir,outfiles[grepl("max*", outfiles)]), readRDS),
+  recursive = FALSE) %>%
+  bind_rows(.id = "hu") %>%
+  mutate(scenario="maxwfs")
+
+one_out <- unlist(
+  lapply(paste0(datadir,outfiles[grepl("onefield*", outfiles)]), readRDS),
+  recursive = FALSE) %>%
+  bind_rows(.id = "hu") %>%
+  mutate(scenario="onefield")
+
+# add other data
+lsm <- read.csv("data/lsm_reduced.csv") %>%
+  mutate(hu = gsub("hu", "", hu))
+lsm2 <- read.csv("data/lsm.csv") %>%
+  mutate(hu = gsub("hu", "", hu))
+hu_meta <- readRDS("data/hu_meta.rds") %>% sf::st_drop_geometry()
+hulandstats <- readRDS("data/hulandstats.rds")
+
+## maximum vs baseline
+
+maxdelta <- base_out %>% select(hu:sed_export,rawsdr_avoid_exp, field_cells) %>%
+  left_join(max_out %>% select(hu:margin_cells,rawsdr_avoid_exp, margin_cells),
+            by = "hu", suffix = c("_base","_max")) %>%
+  mutate(
+    delta_pollinators = pollinators_max - pollinators_base,
+    # delta_n_ret = n_ret_max - n_ret_base,  # N retention - not used because N loading changes between scenarios so N retention is working from different assumptions
+    delta_qb = qb_max - qb_base,   # change per ha
+    # delta_avoidsedexp = rawsdr_avoid_exp_max - rawsdr_avoid_exp_base,  # avoided sediment export - not used because, similar to N retention, different scenarios work with different assumptions (pretty sure upslope sources of sedoiment changes)
+    delta_n_export = n_export_base - n_export_max,  # change per ha
+    delta_sed = sed_export_base - sed_export_max   # change per ha
+  ) %>%
+  select(hu, 
[…]

### 2025-04-01 data-raw/out_20250308.R

@@ -1,108 +1,0 @@
-library(tidyverse)
-
-datadir <- "data-raw/out_20250308/"
-
-outfiles <- list.files(datadir)
-
-base_out <- unlist(
-  lapply(paste0(datadir,outfiles[grepl("base*", outfiles)]), readRDS),
-  recursive = FALSE) %>%
-  bind_rows(.id = "hu") %>%
-  mutate(scenario="baseline")
-
-max_out <- unlist(
-  lapply(paste0(datadir,outfiles[grepl("max*", outfiles)]), readRDS),
-  recursive = FALSE) %>%
-  bind_rows(.id = "hu") %>%
-  mutate(scenario="maxwfs")
-
-one_out <- unlist(
-  lapply(paste0(datadir,outfiles[grepl("onefield*", outfiles)]), readRDS),
-  recursive = FALSE) %>%
-  bind_rows(.id = "hu") %>%
-  mutate(scenario="onefield")
-
-# add other data
-lsm <- read.csv("data/lsm_reduced.csv") %>%
-  mutate(hu = gsub("hu", "", hu))
-hu_meta <- readRDS("data/hu_meta.rds") %>% sf::st_drop_geometry()
-hulandstats <- readRDS("data/hulandstats.rds")
-
-## maximum vs baseline
-
-maxdelta <- base_out %>% select(hu:sed_export,rawsdr_avoid_exp, field_cells) %>%
-  left_join(max_out %>% select(hu:margin_cells,rawsdr_avoid_exp, margin_cells),
-            by = "hu", suffix = c("_base","_max")) %>%
-  mutate(
-    delta_pollinators = pollinators_max - pollinators_base,
-    # delta_n_ret = n_ret_max - n_ret_base,  # N retention - not used because N loading changes between scenarios so N retention is working from different assumptions
-    delta_qb = qb_max - qb_base,   # change per ha
-    # delta_avoidsedexp = rawsdr_avoid_exp_max - rawsdr_avoid_exp_base,  # avoided sediment export - not used because, similar to N retention, different scenarios work with different assumptions (pretty sure upslope sources of sedoiment changes)
-    delta_n_export = n_export_base - n_export_max,  # change per ha
-    delta_sed = sed_export_base - sed_export_max   # change per ha
-  ) %>%
-  select(hu, 
-         pollinators_base, qb_base, n_export_base, sed_export_base, 
-         pollinators_max, qb_max, n_export_max, sed_export_max,
[…]

### 2025-04-01 docs/RandomForest/make_shap_importance.R

@@ -2,11 +2,11 @@
 library(fastshap)
 library(shapviz)
 
-load("data/ES-RF2.RDA")
+load("data/ES-RF3.RDA")
 
 pfun <- function(object, newdata) predict(object, data = newdata)$predictions
 
-dat.x <- dat[,13:41]
+dat.x <- dat[,13:45]
 
 # pollinators
 ex.polbase <- explain(pollinators_base.rf, X=dat.x, pred_wrapper = pfun,
@@ -44,4 +44,4 @@ ex.qbmarg <- explain(qb_margin.rf, X=dat.x, pred_wrapper = pfun,
                       nsim = 100, adjust = TRUE, shap_only = FALSE)
 ex.qbmarg.glb <- shapviz(ex.qbmarg)
 
-save(list = ls(pattern = "^ex"), file = "data/ES-RF2-shapimportance.RDA")
+save(list = ls(pattern = "^ex"), file = "data/ES-RF3-shapimportance.RDA")

### 2025-04-01 docs/data-raw/landmetrics.R

@@ -1,0 +1,109 @@
+library(landscapemetrics)
+library(terra)
+library(tidyverse)
+
+datadir <- "/Users/kevinli/Box Sync/InVEST/Optimization/Data/maxWFS/"
+lcgroup <- read.csv("data-raw/lc_grouping.csv")
+
+# view landscape metrics
+viewlsm <- list_lsm()
+
+hulist <- list.files(paste0(datadir,"hu"))
+
+# aggregated classes reclass table
+lc.agg <- lcgroup %>% 
+  mutate(landcover_ag = abbreviate(landcover_ag,3),
+         landuse_group = abbreviate(landuse_group,3)) %>%
+  unite(col = "landuse_group", landcover_ag,landuse_group, na.rm = TRUE,sep = "_") %>%
+  select(Value, CDL_name, landuse_group) %>%
+  mutate(groupcode = as.numeric(as.factor(landuse_group)))
+
+# saveRDS(lc.agg, "data/aglu_codes.rds")
+# write.csv(lc.agg, "data/aglu_codes.csv", row.names = FALSE)
+
+reclass.agg <- lc.agg %>% select(Value, groupcode)
+
+# landscape metrics
+lsm.list <- c("lsm_l_ta", "lsm_l_prd", "lsm_l_pr", "lsm_l_shdi")
+lsm.agg.list <- c("lsm_l_iji", "lsm_l_frac_mn", "lsm_l_frac_cv", "lsm_l_para_mn", "lsm_l_para_cv", "lsm_l_contag", "lsm_l_ed", "lsm_l_split", "lsm_l_lsi")
+
+# class metrics
+list.agg.cm <- c("lsm_c_frac_mn", "lsm_c_frac_cv", "lsm_c_para_mn", "lsm_c_para_cv", "lsm_c_area_mn", "lsm_c_area_cv", "lsm_c_pland", "lsm_c_pd", "lsm_c_cohesion", "lsm_c_ed", "lsm_c_split", "lsm_c_lsi")
+
+ls.result <- list()
+ls.result.agg <- list()
+ls.result.class <- list()
+ls.check <- list()
+
+for(i in seq_along(hulist)){
+  
+  hu.i <- terra::rast(paste0(datadir,"hu/",hulist[i],"/data/lu/fields.tif"))
+  mask.i <- terra::vect(paste0(datadir,"hu/",hulist[i],"/data/watershed.gpkg"))
[…]

### 2025-04-01 docs/data-raw/old/out_20250308.R

@@ -1,0 +1,111 @@
+library(tidyverse)
+
+datadir <- "data-raw/out_20250308/"
+
+outfiles <- list.files(datadir)
+
+base_out <- unlist(
+  lapply(paste0(datadir,outfiles[grepl("base*", outfiles)]), readRDS),
+  recursive = FALSE) %>%
+  bind_rows(.id = "hu") %>%
+  mutate(scenario="baseline")
+
+max_out <- unlist(
+  lapply(paste0(datadir,outfiles[grepl("max*", outfiles)]), readRDS),
+  recursive = FALSE) %>%
+  bind_rows(.id = "hu") %>%
+  mutate(scenario="maxwfs")
+
+one_out <- unlist(
+  lapply(paste0(datadir,outfiles[grepl("onefield*", outfiles)]), readRDS),
+  recursive = FALSE) %>%
+  bind_rows(.id = "hu") %>%
+  mutate(scenario="onefield")
+
+# add other data
+lsm <- read.csv("data/lsm_reduced.csv") %>%
+  mutate(hu = gsub("hu", "", hu))
+lsm2 <- read.csv("data/lsm.csv") %>%
+  mutate(hu = gsub("hu", "", hu))
+hu_meta <- readRDS("data/hu_meta.rds") %>% sf::st_drop_geometry()
+hulandstats <- readRDS("data/hulandstats.rds")
+
+## maximum vs baseline
+
+maxdelta <- base_out %>% select(hu:sed_export,rawsdr_avoid_exp, field_cells) %>%
+  left_join(max_out %>% select(hu:margin_cells,rawsdr_avoid_exp, margin_cells),
+            by = "hu", suffix = c("_base","_max")) %>%
+  mutate(
+    delta_pollinators = pollinators_max - pollinators_base,
+    # delta_n_ret = n_ret_max - n_ret_base,  # N retention - not used because N loading changes between scenarios so N retention is working from different assumptions
+    delta_qb = qb_max - qb_base,   # change per ha
+    # delta_avoidsedexp = rawsdr_avoid_exp_max - rawsdr_avoid_exp_base,  # avoided sediment export - not used because, similar to N retention, different scenarios work with different assumptions (pretty sure upslope sources of sedoiment changes)
+    delta_n_export = n_export_base - n_export_max,  # change per ha
+    delta_sed = sed_export_base - sed_export_max   # change per ha
+  ) %>%
+  select(hu, 
[…]

### 2025-04-01 docs/data-raw/old/wfs_lsm.R

@@ -1,0 +1,26 @@
+## code to prepare `wfs_lsm` dataset goes here
+
+wfs <- readRDS("data/maxwfsresults_20241217_fix.rds")
+lsm <- read.csv("data/lsm_reduced.csv") %>% 
+  mutate(hu = gsub("hu", "", hu))
+
+wfs_lsm <- wfs %>% left_join(lsm, by = "hu") %>%
+  select(c(hu, huc2, name, margin.ha, lu_ta,
+           pollinators_base:sed_export_base, 
+           pollinators_max:sed_export_max, 
+           delta_pollinators:delta_sed, 
+           fields_max, field.ha, margin.ha, 
+           lu_shdi:ntr_frac_mn)) %>%
+  mutate(prn_ann_ratio = log(prn_agr_pland/ann_agr_pland),
+         prn_pagr = 100*(prn_agr_pland/(ann_agr_pland+prn_agr_pland)),
+         prn_ann_lograt = log(prn_agr_pland/ann_agr_pland),
+         field_pland = field.ha/lu_ta,
+         margin_dens = margin.ha/lu_ta,
+         delta_pollinators.margin = delta_pollinators/margin_dens,
+         delta_qb.margin = delta_qb/margin_dens,
+         delta_n_export.margin = delta_n_export/margin_dens,
+         delta_sed.margin = delta_sed/margin_dens) %>%
+  mutate(habcat = cut(ntr_pland, breaks = c(-Inf,20,50,Inf),
+                      labels = c("simple","moderate","natural")))
+
+saveRDS(wfs_lsm, "data/wfs_lsm.rds")

### 2025-04-01 docs/data-raw/wfs_lsm.R

@@ -1,26 +1,0 @@
-## code to prepare `wfs_lsm` dataset goes here
-
-wfs <- readRDS("data/maxwfsresults_20241217_fix.rds")
-lsm <- read.csv("data/lsm_reduced.csv") %>% 
-  mutate(hu = gsub("hu", "", hu))
-
-wfs_lsm <- wfs %>% left_join(lsm, by = "hu") %>%
-  select(c(hu, huc2, name, margin.ha, lu_ta,
-           pollinators_base:sed_export_base, 
-           pollinators_max:sed_export_max, 
-           delta_pollinators:delta_sed, 
-           fields_max, field.ha, margin.ha, 
-           lu_shdi:ntr_frac_mn)) %>%
-  mutate(prn_ann_ratio = log(prn_agr_pland/ann_agr_pland),
-         prn_pagr = 100*(prn_agr_pland/(ann_agr_pland+prn_agr_pland)),
-         prn_ann_lograt = log(prn_agr_pland/ann_agr_pland),
-         field_pland = field.ha/lu_ta,
-         margin_dens = margin.ha/lu_ta,
-         delta_pollinators.margin = delta_pollinators/margin_dens,
-         delta_qb.margin = delta_qb/margin_dens,
-         delta_n_export.margin = delta_n_export/margin_dens,
-         delta_sed.margin = delta_sed/margin_dens) %>%
-  mutate(habcat = cut(ntr_pland, breaks = c(-Inf,20,50,Inf),
-                      labels = c("simple","moderate","natural")))
-
-saveRDS(wfs_lsm, "data/wfs_lsm.rds")

### 2025-04-01 docs/scripts/make_aglu_map.R

@@ -1,0 +1,19 @@
+library(terra)
+
+lc.agg <- readRDS("data/aglu_codes.rds")
+lc.agg[which(lc.agg$Value==0),"groupcode"] <- NA
+
+nemw <- rast("/Users/kevinl/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/NEMWregion/focalarea_cdl.tif")
+
+# active category to first column (value)
+activeCat(nemw) <- 0
+
+nemw_aglu <- classify(nemw, lc.agg[,c(1,4)])
+
+# create a new aggregated land use table
+aglu <- lc.agg %>% select(groupcode, landuse_group) %>% distinct() %>% filter(!is.na(groupcode))
+
+levels(nemw_aglu) <- aglu
+
+writeRaster(nemw_aglu, 
+            "/Users/kevinl/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/GIS/NEMWregion/focalarea_rcl.tif")

### 2025-04-01 landmetrics.R

@@ -1,106 +1,0 @@
-library(landscapemetrics)
-library(terra)
-library(tidyverse)
-
-datadir <- "/Users/kevinli/Box Sync/InVEST/Optimization/Data/maxWFS/"
-lcgroup <- read.csv("lc_grouping.csv")
-
-# view landscape metrics
-viewlsm <- list_lsm()
-
-hulist <- list.files(paste0(datadir,"hu"))
-
-# aggregated classes reclass table
-lc.agg <- lcgroup %>% 
-  mutate(landcover_ag = abbreviate(landcover_ag,3),
-         landuse_group = abbreviate(landuse_group,3)) %>%
-  unite(col = "landuse_group", landcover_ag,landuse_group, na.rm = TRUE,sep = "_") %>%
-  select(Value, CDL_name, landuse_group) %>%
-  mutate(groupcode = as.numeric(as.factor(landuse_group)))
-
-reclass.agg <- lc.agg %>% select(Value, groupcode)
-
-# landscape metrics
-lsm.list <- c("lsm_l_ta", "lsm_l_prd", "lsm_l_pr", "lsm_l_shdi")
-lsm.agg.list <- c("lsm_l_iji", "lsm_l_frac_mn", "lsm_l_frac_cv", "lsm_l_para_mn", "lsm_l_para_cv", "lsm_l_contag", "lsm_l_ed", "lsm_l_split", "lsm_l_lsi")
-
-# class metrics
-list.agg.cm <- c("lsm_c_frac_mn", "lsm_c_frac_cv", "lsm_c_para_mn", "lsm_c_para_cv", "lsm_c_area_mn", "lsm_c_area_cv", "lsm_c_pland", "lsm_c_pd", "lsm_c_cohesion", "lsm_c_ed", "lsm_c_split", "lsm_c_lsi")
-
-ls.result <- list()
-ls.result.agg <- list()
-ls.result.class <- list()
-ls.check <- list()
-
-for(i in seq_along(hulist)){
-  
-  hu.i <- terra::rast(paste0(datadir,"hu/",hulist[i],"/data/lu/fields.tif"))
-  mask.i <- terra::vect(paste0(datadir,"hu/",hulist[i],"/data/watershed.gpkg"))
-  hucrop <- terra::mask(hu.i$cdl.clean, mask.i)
-  
-  hucrop.agg <- hucrop %>% classify(rcl = reclass.agg)
-  
[…]

### 2025-04-01 Paper_analysis.Rmd
update again

@@
 ```{r shap_pol, fig.width=9}
-grid.arrange(
+cowplot::plot_grid(
   sv_importance(ex.polbase.glb) + egg::theme_article(),
@@
 
-grid.arrange(
-  polbase.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP"),
-  polbase.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE"),
+cowplot::plot_grid(
+  polbase.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP") + egg::theme_article(),
+  polbase.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE") + egg::theme_article(),
 
-  polbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP"),
-  polbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland ALE"),
-  ncol = 2
+  polbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP") + egg::theme_article(),
+  polbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland ALE") + egg::theme_article(),
+  ncol = 2, align = "v"
 )
@@
 ```{r shap_pol_margshap, fig.width=9}
-grid.arrange(
+cowplot::plot_grid(
   sv_dependence(ex.polmarg.glb, v="ntr_pland") + egg::theme_article(),
-  sv_dependence(ex.polmarg.glb, v="ann_agr_pland") + egg::theme_article()
+  sv_dependence(ex.polmarg.glb, v="ann_agr_pland") + egg::theme_article(),
+  ncol = 2, align = "v"
 )
@@
 
-grid.arrange(
-  polmarg.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP"),
-  polmarg.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE"),
+cowplot::plot_grid(
+  polmarg.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP") + egg::theme_article(),
+  polmarg.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE") + egg::theme_article(),
 
-  polmarg.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP"),
-  polmarg.ann_agr_pland.ale$plot() + ggtitle("ann_agr_pland ALE"),
-  ncol = 2
+  polmarg.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP") + egg::theme_article(),
+  polmarg.ann_agr_pland.ale$plot() + ggtitle("ann_agr_pland ALE") + egg::theme_article(),
+  ncol = 2, align = "v"
 )
@@
 
-grid.arrange(
+cowplot::plot_grid(
   sv_importance(ex.nexpbase.glb) + egg::theme_article(), 
@@
 
-grid.arrange(
+cowplot::plot_grid(
   sv_dependence(ex.nexpbase.glb, v="ntr_pland") + egg::theme_article(),
@@
   sv_dependence(ex.nexpbase.glb, v="ntr_ed") + egg::theme_article(), 
-  ncol = 2
+  ncol = 2, align = "v"
 )
@@
 
-grid.arrange(
-  nexpbase.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP"),
-  nexpbase.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE"),
+cowplot::plot_grid(
+  nexpbase.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP") + egg::theme_article(),
+  nexpbase.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE") + egg::theme_article(),
 
-  nexpbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP"),
-  nexpbase.ann_agr_pland.ale$plot() + ggtitle("ann_agr_pland ALE"),
+  nexpbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP") + egg::theme_article(),
+  nexpbase.ann_agr_pland.ale$plot() + ggtitle("ann_agr_pland ALE") + egg::theme_article(),
 
-  nexpbase.ntr_ed.pdp$plot() + ggtitle("ntr_ed PDP"),
-  nexpbase.ntr_ed.ale$plot() + ggtitle("ntr_ed ALE"),
-  ncol = 2
+  nexpbase.ntr_ed.pdp$plot() + ggtitle("ntr_ed PDP") + egg::theme_article(),
+  nexpbase.ntr_ed.ale$plot() + ggtitle("ntr_ed ALE") + egg::theme_article(),
+  ncol = 2, align = "v"
 )
@@
 ```{r shap_nexp_margshap, fig.width=9}
-grid.arrange(
+cowplot::plot_grid(
   sv_dependence(ex.nexpmarg.glb, v="ntr_pland") + egg::theme_article(),
@@
   sv_dependence(ex.nexpmarg.glb, v="ntr_ed") + egg::theme_article(), 
-  ncol = 2
+  ncol = 2, align = "v"
 )
@@
 
-grid.arrange(
-  nexpmarg.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP"),
-  nexpmarg.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE"),
+cowplot::plot_grid(
+  nexpmarg.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP") + egg::theme_article(),
+  nexpmarg.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE") + egg::theme_article(),
 
-  nexpmarg.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP"),
-  nexpmarg.ann_agr_pland.ale$plot() + ggtitle("ann_agr_pland ALE"),
+  nexpmarg.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP") + egg::theme_article(),
+  nexpmarg.ann_agr_pland.ale$plot() + ggtitle("ann_agr_pland ALE") + egg::theme_article(),
 
-  nexpmarg.ntr_ed.pdp$plot() + ggtitle("ntr_ed PDP"),
-  nexpmarg.ntr_ed.ale$plot() + ggtitle("ntr_ed ALE"),
-  ncol = 2
+  nexpmarg.ntr_ed.pdp$plot() + ggtitle("ntr_ed PDP") + egg::theme_article(),
+  nexpmarg.ntr_ed.ale$plot() + ggtitle("ntr_ed ALE") + egg::theme_article(),
+  ncol = 2, align = "v"
 )
@@
 ```{r shap_sed, fig.width=9}
-grid.arrange(
+cowplot::plot_grid(
   sv_importance(ex.sedbase.glb) + egg::theme_article(),
@@
 ```{r shap_sed_baseshap, fig.width=9}
-grid.arrange(
 sv_dependence(ex.sedbase.glb, v="ann_agr_ed") + egg::theme_article()
-)
 
@@
 
-grid.arrange(
+cowplot::plot_grid(
   sedbase.ann_agr_ed.pdp$plot() + ggtitle("ann_agr_ed PDP") +
@@
 
-  ncol = 2
+  ncol = 2, align = "v"
 )
@@
 ```{r shap_sed_margshap, fig.width=9}
-grid.arrange(
+cowplot::plot_grid(
   sv_dependence(ex.sedmarg.glb, v="ann_agr_ed") + egg::theme_article(),
   sv_dependence(ex.sedmarg.glb, v="field_frac_mn") + egg::theme_article(),
-  ncol=2
+  ncol=2, align = "v"
 )
@@
 
-grid.arrange(
+cowplot::plot_grid(
   sedmarg.ann_agr_ed.pdp$plot() + ggtitle("ann_agr_ed PDP") +
@@
 ```{r shap_qb, fig.width=9}
-grid.arrange(
+cowplot::plot_grid(
   sv_importance(ex.qbbase.glb) + egg::theme_article(),
@@
 ```{r shap_qb_baseshap, fig.width=9}
-grid.arrange(
-  sv_dependence(ex.qbbase.glb, v="field_frac_mn") + egg::theme_article()
+cowplot::plot_grid(
+  sv_dependence(ex.qbbase.glb, v="ann_agr_cohesion") + egg::theme_article(),
+  sv_dependence(ex.qbbase.glb, v="field_frac_mn") + egg::theme_article(),
+  ncol = 2, align = "v"
 )
@@
 ```{r qb_base_ale, fig.width=9}
+
+qbbase.ann_agr_cohesion.pdp = FeatureEffect$new(pred.qb_base, 
+                                             feature = "ann_agr_cohesion",
+                                             method = "pdp")
+
+sedbase.ann_agr_cohesion.ale = FeatureEffect$new(pred.qb_base, 
+                                              feature = "ann_agr_cohesion",
+                                              method = "ale")
 
@@
 
-grid.arrange(
+cowplot::plot_grid(
+  qbbase.ann_agr_cohesion.pdp$plot() + ggtitle("ann_agr_cohesion PDP") +
+    egg::theme_article(),
+  sedbase.ann_agr_cohesion.ale$plot() + ggtitle("ann_agr_cohesion ALE") +
+    egg::theme_article(),
+
   qbbase.field_frac_mn.pdp$plot() + ggtitle("field_frac_mn PDP") +
@@
 ```{r shap_qb_margshap, fig.width=9}
-grid.arrange(
+cowplot::plot_grid(
   sv_dependence(ex.qbmarg.glb, v="ann_agr_cohesion") + egg::theme_article(),
@@
   sv_dependence(ex.qbmarg.glb, v="prn_agr_para_mn") + egg::theme_article(),
-  ncol=2
+  ncol=2, align = "v"
 )
@@
 
-grid.arrange(
-  qbmarg.ann_agr_cohesion.pdp$plot() + ggtitle("ann_agr_cohesion PDP") +
+cowplot::plot_grid(
+  qbmarg.ann_agr_cohesion.pdp$plot() + ggtitle("ann_agr_cohesion PDP") + ylab("PDP") +
     egg::theme_article(),
-  qbmarg.ann_agr_cohesion.ale$plot() + ggtitle("ann_agr_cohesion ALE") +
+  qbmarg.ann_agr_cohesion.ale$plot() + ggtitle("ann_agr_cohesion ALE") + ylab("ALE") +
     egg::theme_article(),
 
-  qbmarg.prn_pagr.pdp$plot() + ggtitle("prn_pagr PDP") +
+  qbmarg.prn_pagr.pdp$plot() + ggtitle("prn_pagr PDP") + ylab("PDP") +
     egg::theme_article(),
-  qbmarg.prn_pagr.ale$plot() + ggtitle("prn_pagr ALE") +
+  qbmarg.prn_pagr.ale$plot() + ggtitle("prn_pagr ALE") + ylab("ALE") +
     egg::theme_article(),
 
-  qbmarg.prn_agr_para_mn.pdp$plot() + ggtitle("prn_agr_para_mn PDP") +
+  qbmarg.prn_agr_para_mn.pdp$plot() + ggtitle("prn_agr_para_mn PDP") + ylab("PDP") +
     egg::theme_article(),
-  qbmarg.prn_agr_para_mn.ale$plot() + ggtitle("prn_agr_para_mn ALE") +
+  qbmarg.prn_agr_para_mn.ale$plot() + ggtitle("prn_agr_para_mn ALE") + ylab("ALE") +
     egg::theme_article(),
 
-  ncol = 2
+  ncol = 2, align = "v"
 )
@@
 
-### SHAP force plots
-
-I am plotting the combined SHAP force values for each watershed together.
-The important features identified above are colored while the rest are left grey.
-To plot watersheds on a common axis and visualize tradeoffs, I used PCA to identify the two major axes.
-The first PC axis for both baseline and margin results show a tradeoff between N export (and improvement in N export) and pollinator abundance (pollinator improvement).
-The second PC shows sediment export and baseflow changing together while with margin improvement are in the same direction.
+#### Correlation of important variables
 
@@
 
-# psych::pairs.panels(dat.selectedvar)
+dat.selectedvar <- dat %>% select(selectedvar)
+psych::pairs.panels(dat.selectedvar)
 
 ```
+
+### SHAP force plots
+
+I am plotting the combined SHAP force values for each watershed together.
+The important features identified above are colored while the rest are left grey.
+To plot watersheds on a common axis and visualize tradeoffs, I used PCA to identify the two major axes.
+The first PC axis for both baseline and margin results show a tradeoff between N export (and improvement in N export) and pollinator abundance (pollinator improvement).
+The second PC shows sediment export and baseflow changing together while with margin improvement are in the same direction.
 

### 2025-04-04 Paper_analysis.Rmd
another update
another update to site

@@
 
-```{r shap_nexp, fig.width=9}
+```{r shap_nexp2, fig.width=9}
 
@@
   sv_importance(ex.nexpmarg.glb) + egg::theme_article(),
-  ncol = 2
+  ncol = 2, align = "v"
 )
@@
   sv_dependence(ex.qbmarg.glb, v="prn_pagr") + egg::theme_article(),
-  sv_dependence(ex.qbmarg.glb, v="prn_agr_para_mn") + egg::theme_article(),
   ncol=2, align = "v"
@@
                                              method = "pdp")
-qbmarg.prn_agr_para_mn.pdp = FeatureEffect$new(pred.qb_margin, 
-                                             feature = "prn_agr_para_mn",
-                                             method = "pdp")
 
@@
                                               feature = "prn_pagr",
-                                              method = "ale")
-qbmarg.prn_agr_para_mn.ale = FeatureEffect$new(pred.qb_margin, 
-                                              feature = "prn_agr_para_mn",
                                               method = "ale")
@@
 
-  qbmarg.prn_agr_para_mn.pdp$plot() + ggtitle("prn_agr_para_mn PDP") + ylab("PDP") +
-    egg::theme_article(),
-  qbmarg.prn_agr_para_mn.ale$plot() + ggtitle("prn_agr_para_mn ALE") + ylab("ALE") +
     egg::theme_article(),
@@
 selectedvar <- c("ntr_pland", "ann_agr_pland", "ntr_ed", 
-                 "field_frac_mn", "prn_pagr", "prn_agr_para_mn",
+                 "field_frac_mn", "prn_pagr",
                  "ann_agr_ed", "ann_agr_cohesion"
@@
     "prn_pagr" ~ "Agriculture % perennial",
-    "prn_agr_para_mn" ~ "Perennial ag. perimeter:area",
     "ann_agr_ed" ~ "Annual ag. edge density",
@@
   "Agriculture % perennial" = "#CC79A7",
-  "Perennial ag. perimeter:area" = "#D55E00",
   "Annual ag. edge density" = "#F0E442",

### 2025-04-04 Paper_analysis_percent.Rmd
another update to site

---
title: "Percent change"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r load}
library(tidyverse)
library(iml)
library(ranger)
load("data/ES-RF3pc.RDA")

```

## Ecosystem service improvements depend on the baseline condition

```{r data_setup}

dat.long <- dat %>% 
  pivot_longer(cols = c(1:8,13:16), names_to = "ES_response", values_to = "ES_value") %>%
  pivot_longer(cols = c("ntr_pland", "ann_agr_pland"), names_to = "Landuse",
               values_to = "Landuse_value") %>%
  select(ES_response, ES_value, Landuse, Landuse_value) %>%
  mutate(ES = gsub("delta_|_base|_(incr|red).margin|pc_","", ES_response)) %>%
  mutate(ES = case_match(ES,
                         "sed" ~ "sed_export",
                         "pol" ~ "pollinators",
                         "nexp" ~ "n_export",
                         .default = ES)) %>%
  mutate(response_type =
           gsub("sed|export|n_|pollinators|qb|_|pol|nexp|red.|incr.","",ES_response)) %>%
  mutate(response_type = case_match(
    response_type,
    "base" ~ "Baseline ES [/ha]",
    "delta" ~ "\u0394 ES [/ha]",
    "pcmargin" ~ "% \u0394 ES per ha margin"
  )) %>%
  mutate(response_type = factor(response_type, 
                                levels = c("Baseline ES [/ha]",
                                           "\u0394 ES [/ha]",
                                           "% \u0394 ES per ha margin")))

```

Ecosystem service improvements depend on the baseline condition, except for pollination.

This pattern holds even if we factor out margin area.

The trend is different if we look at % change in ecosystem services from baseline, per margin area

```{r percentchangedemo}

dat_wider <- dat.long %>% select(-ES_response) %>% 
  pivot_wider(names_from = response_type, values_from = ES_value)

base.v.margin <- ggplot(dat_wider, aes(`Baseline ES [/ha]`, `Δ ES [/ha]`)) + geom_point() + 
  xlab("Baseline ES per ha") + ylab("Increase in ES per ha margin (ES units)") +
  geom_smooth(method = "gam") + facet_wrap(~ES, scales = "free") + 
  ggtitle("ES baseline vs \nES improvement per margin") + egg::theme_article()

base.v.margin.base <- ggplot(dat_wider, aes(`Baseline ES [/ha]`, `% Δ ES per ha margin`)) + 
  geom_point() + geom_smooth(method = "gam") + facet_wrap(~ES, scales = "free") +
  ggtitle("ES Baseline vs \n% ES improvement per margin") + 
  xlab("Baseline ES per ha") +
  ylab("% improvement from baseline per ha margin") +
  egg::theme_article()

cowplot::plot_grid(base.v.margin, base.v.margin.base)

```

## Patterns with annual agriculture

Looking at ES % change from baseline shows some stronger patterns than before for sediment and water recharge.

```{r compplots, fig.width=9, fig.height=6}

# fit some gams
dat.gam <- dat.long %>% filter(Landuse == "ntr_pland") %>% 
  group_by(response_type, ES) %>%
  nest() %>%
  # Fit GAM
  mutate(Mod = map(data, ~mgcv::gam(ES_value ~ s(Landuse_value), data = .x))) %>%
  # Get R2
  mutate(R2 = map_dbl(Mod, ~round(summary(.x)$r.sq, 2)))

ggplot(dat.long %>% filter(Landuse == "ann_agr_pland"),
       aes(Landuse_value, ES_value)) + 
  geom_point() + geom_smooth(method="gam") +
  ggh4x::facet_grid2(response_type ~ ES,  scales = "free", independent = "y", 
                     switch = "y") + 
  egg::theme_article() + 
  # Add label
  geom_label(data = dat.gam, 
             aes(x = Inf, y = Inf, 
                 label = paste("R2 = ", R2, sep = " ")),
             hjust = 1, vjust = 1) +
  xlab("% Annual agriculture") +
  ylab("ES value") + theme(strip.placement = "outside")
```

## Random forest models for % improvement

Units for % improvement per margin ha are in tenths of a percent, so MSE looks smaller.

### Model performance
*Pollinators:*

     r2 = `r round(pollinators_pc.rf$r.squared, 3)`
     OOB MSE = `r format(round(pollinators_pc.rf$prediction.error, 5),scientific=F)`

*Nitrogen export:*

     r2 = `r round(n_export_pc.rf$r.squared, 3)`
     OOB MSE = `r format(round(n_export_pc.rf$prediction.error, 5),scientific=F)`

*Sediment export:*

     r2 = `r round(sed_pc.rf$r.squared, 3)`
     OOB MSE = `r format(round(sed_pc.rf$prediction.error, 5),scientific=F)`

*Baseflow:*

     r2 = `r round(qb_pc.rf$r.squared, 3)`
     OOB MSE = `r format(round(qb_pc.rf$prediction.error, 5),scientific=F)`

### Variable importance

```{r importance, fig.height=16, fig.width=9}
library(ggpattern)

metric.descr <- read.csv("data/lsm_description.csv")

imp.df <- bind_rows(
  pollinators_pc.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "Pollinators", model = "base"),
  n_export_pc.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "N export", model = "base"),
  sed_pc.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "Sed export", model = "base"),
  qb_pc.p %>% rownames_to_column("metric") %>% 
    mutate(ES = "Baseflow", model = "base")
) %>%
  mutate(sig = ifelse(pvalue <0.05, TRUE, FALSE)) %>%
  mutate(metric2 = tidytext::reorder_within(metric, importance,
                                           within = list(ES, model))) %>%
  left_join(metric.descr %>% select(lsm, type), by = c("metric"="lsm")) %>%
  mutate(type = case_match(
    metric,
    "field_pland" ~ "area and edge metric",
    "field_frac_mn" ~ "shape metric",
    "field_frac_cv" ~ "shape metric",
    "prn_pagr" ~ "area and edge metric",
    .default = type
  ))

# ggplot(imp.df, aes(metric2, importance, fill = type, pattern = sig)) + 
#   geom_bar_pattern(stat="identity",
#                    color = "black", 
#                    pattern_fill = "black",
#                    pattern_angle = 45,
#                    pattern_density = 0.1,
#                    pattern_spacing = 0.025,
#                    pattern_key_scale_factor = 0.6) + 
#   tidytext::scale_x_reordered() +
#   coord_flip() +
#   ggh4x::facet_grid2(ES~model, scales = "free", independent = "all") +
#   scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
#   egg::theme_article() +
#   theme(legend.position = "top") + 
#   guides(pattern = guide_legend(override.aes = list(fill = "white")),
#          fill = guide_legend(override.aes = list(pattern = "none")))

```

```{r importance_shap}
load("data/ES-RF3pc-preds.RDA")   # prediction objects from iml package
load("data/ES-RF3pc-shapimportance.RDA")
library(shapviz)
library(gridExtra)

```

### Pollinators

#### Permutation based importance

```{r pol_permimport, fig.width=9}

ggplot(imp.df %>% filter(ES=="Pollinators"), aes(metric2, importance, fill = type, pattern = sig)) + 
  geom_bar_pattern(stat="identity",
                   color = "black", 
                   pattern_fill = "black",
                   pattern_angle = 45,
                   pattern_density = 0.1,
                   pattern_spacing = 0.025,
                   pattern_key_scale_factor = 0.6) + 
  tidytext::scale_x_reordered() +
  coord_flip() +
  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
  egg::theme_article() +
  theme(legend.position = "top") + 
  guides(pattern = guide_legend(override.aes = list(fill = "white")),
         fill = guide_legend(override.aes = list(pattern = "none")))

```

#### Shapley importance

```{r shap_pol, fig.width=9}
  sv_importance(ex.polpc.glb) + egg::theme_article()

```

#### Base pollinator model Shapley dependence plots

The two major contributors. Colors are the strongest interactor with that variable.

```{r shap_pol_baseshap, fig.width=9}
grid.arrange(
  sv_dependence(ex.polpc.glb, v="ntr_pland") + egg::theme_article(),
  sv_dependence(ex.polpc.glb, v="ann_agr_pland") + egg::theme_article(), 
  ncol = 2
)

```

**_Compare to PDP and ALE_**

```{r pol_base_ale, fig.width=9}

polbase.ntr_pland.pdp = FeatureEffect$new(pred.pollinators_pc, 
                                          feature = "ntr_pland",
                                          method = "pdp")
polbase.ann_agr_pland.pdp = FeatureEffect$new(pred.pollinators_pc, 
                                          feature = "ann_agr_pland",
                                          method = "pdp")

polbase.ntr_pland.ale = FeatureEffect$new(pred.pollinators_pc, 
                                          feature = "ntr_pland",
                                          method = "ale")
polbase.ann_agr_pland.ale = FeatureEffect$new(pred.pollinators_pc, 
                                          feature = "ann_agr_pland",
                                          method = "ale")

cowplot::plot_grid(
  polbase.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP") + egg::theme_article(),
  polbase.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE") + egg::theme_article(),

  polbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland PDP") + egg::theme_article(),
  polbase.ann_agr_pland.pdp$plot() + ggtitle("ann_agr_pland ALE") + egg::theme_article(),
  ncol = 2, align = "v"
)

```

### N Export

### Permutation based importance

```{r nexp_permimport, fig.width=9}

ggplot(imp.df %>% filter(ES=="N export"), aes(metric2, importance, fill = type, pattern = sig)) + 
  geom_bar_pattern(stat="identity",
                   color = "black", 
                   pattern_fill = "black",
                   pattern_angle = 45,
                   pattern_density = 0.1,
                   pattern_spacing = 0.025,
                   pattern_key_scale_factor = 0.6) + 
  tidytext::scale_x_reordered() +
  coord_flip() +
  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
  egg::theme_article() +
  theme(legend.position = "top") + 
  guides(pattern = guide_legend(override.aes = list(fill = "white")),
         fill = guide_legend(override.aes = list(pattern = "none")))

```

#### Shapley importance

```{r shap_nexp2, fig.width=9}

  sv_importance(ex.nexppc.glb) + egg::theme_article()

```

#### Base N export shapley dependence plots

Two major -ish contributors.

```{r shap_nexp_pcshap, fig.width=9}

cowplot::plot_grid(
  sv_dependence(ex.nexppc.glb, v="ntr_pland") + egg::theme_article(),
  sv_dependence(ex.nexppc.glb, v="ntr_cohesion") + egg::theme_article(), 
  ncol = 2, align = "v"
)

```

**_Compare to PDP and ALE_**

```{r nexp_pc_ale, fig.width=9}

nexppc.ntr_pland.pdp = FeatureEffect$new(pred.nexp_pc, 
                                          feature = "ntr_pland",
                                          method = "pdp")
nexppc.ntr_cohesion.pdp = FeatureEffect$new(pred.nexp_pc, 
                                          feature = "ntr_cohesion",
                                          method = "pdp")

nexppc.ntr_pland.ale = FeatureEffect$new(pred.nexp_pc, 
                                          feature = "ntr_pland",
                                          method = "ale")
nexppc.ntr_cohesion.ale = FeatureEffect$new(pred.nexp_pc, 
                                          feature = "ntr_cohesion",
                                          method = "ale")

cowplot::plot_grid(
  nexppc.ntr_pland.pdp$plot() + ggtitle("ntr_pland PDP") + egg::theme_article(),
  nexppc.ntr_pland.ale$plot() + ggtitle("ntr_pland ALE") + egg::theme_article(),

  nexppc.ntr_cohesion.pdp$plot() + ggtitle("ntr_cohesion PDP") + egg::theme_article(),
  nexppc.ntr_cohesion.ale$plot() + ggtitle("ntr_cohesion ALE") + egg::theme_article(),

  ncol = 2, align = "v"
)

```

### Sediment Export

### Permutation based importance

```{r sed_permimport, fig.width=9}

ggplot(imp.df %>% filter(ES=="Sed export"), aes(metric2, importance, fill = type, pattern = sig)) + 
  geom_bar_pattern(stat="identity",
                   color = "black", 
                   pattern_fill = "black",
                   pattern_angle = 45,
                   pattern_density = 0.1,
                   pattern_spacing = 0.025,
                   pattern_key_scale_factor = 0.6) + 
  tidytext::scale_x_reordered() +
  coord_flip() +
  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
  egg::theme_article() +
  theme(legend.position = "top") + 
  guides(pattern = guide_legend(override.aes = list(fill = "white")),
         fill = guide_legend(override.aes = list(pattern = "none")))

```

#### Shapley importance

```{r shap_sed, fig.width=9}

sv_importance(ex.sedpc.glb) + egg::theme_article()

```

#### Percent improvement in sediment export Shapley dependence plots

One major contributor

```{r shap_sed_baseshap, fig.width=9}
sv_dependence(ex.sedpc.glb, v="ntr_pland") + egg::theme_article()

```

**_Compare to PDP and ALE_**

```{r sedexp_pc_ale, fig.width=9}

sedpc.ntr_pland.pdp = FeatureEffect$new(pred.sed_pc, 
                                          feature = "ntr_pland",
                                          method = "pdp")

sedpc.ntr_pland.ale = FeatureEffect$new(pred.sed_pc, 
                                          feature = "ntr_pland",
                                          method = "ale")

cowplot::plot_grid(
  sedpc.ntr_pland.pdp$plot() + ggtitle("ann_agr_ed PDP") +
    egg::theme_article(),
  sedpc.ntr_pland.ale$plot() + ggtitle("ann_agr_ed ALE") +
    egg::theme_article(),

  ncol = 2, align = "v"
)

```

### Water recharge

### Permutation based importance

```{r qb_permimport, fig.width=9}

ggplot(imp.df %>% filter(ES=="Baseflow"), aes(metric2, importance, fill = type, pattern = sig)) + 
  geom_bar_pattern(stat="identity",
                   color = "black", 
                   pattern_fill = "black",
                   pattern_angle = 45,
                   pattern_density = 0.1,
                   pattern_spacing = 0.025,
                   pattern_key_scale_factor = 0.6) + 
  tidytext::scale_x_reordered() +
  coord_flip() +
  scale_pattern_manual(values = c("TRUE" = "none", "FALSE" = "stripe")) +
  egg::theme_article() +
  theme(legend.position = "top") + 
  guides(pattern = guide_legend(override.aes = list(fill = "white")),
         fill = guide_legend(override.aes = list(pattern = "none")))

```

#### Shapley importance

```{r shap_qb, fig.width=9}

sv_importance(ex.qbpc.glb) + egg::theme_article()

```

#### Base water runoff Shapley depednence plot

One major contributor.

```{r shap_qb_baseshap, fig.width=9}
cowplot::plot_grid(
  sv_dependence(ex.qbpc.glb, v="ntr_ed") + egg::theme_article(),
  sv_dependence(ex.qbpc.glb, v="prn_agr_pd") + egg::theme_article(),
  ncol = 2, align = "v"
)

```

**_Compare to PDP and ALE_**

```{r qb_pc_ale, fig.width=9}

qbpc.ntr_ed.pdp = FeatureEffect$new(pred.qb_pc, 
                                             feature = "ntr_ed",
                                             method = "pdp")

qbpc.ntr_ed.ale = FeatureEffect$new(pred.qb_pc, 
                                              feature = "ntr_ed",
                                              method = "ale")

qbpc.prn_agr_pd.pdp = FeatureEffect$new(pred.qb_pc, 
                                             feature = "prn_agr_pd",
                                             method = "pdp")

qbpc.prn_agr_pd.ale = FeatureEffect$new(pred.qb_pc, 
                                              feature = "prn_agr_pd",
                                              method = "ale")

cowplot::plot_grid(
  qbpc.ntr_ed.pdp$plot() + ggtitle("ntr_ed PDP") +
    egg::theme_article(),
  qbpc.ntr_ed.ale$plot() + ggtitle("ntr_ed ALE") +
    egg::theme_article(),

  qbpc.prn_agr_pd.pdp$plot() + ggtitle("prn_agr_pd PDP") +
    egg::theme_article(),
  qbpc.prn_agr_pd.ale$plot() + ggtitle("prn_agr_pd ALE") +
    egg::theme_article(),

  ncol = 2, align = "v"
)

```

### 2025-04-04 RandomForest/make_rf.R

@@ -3,13 +3,14 @@
 library(ranger)
 options(ranger.num.threads = 8)
 
-wfs <- readRDS("data/allresults_20250401.rds")
+# wfs <- readRDS("data/allresults_20250401.rds")
+wfs <- readRDS("data/allresults_20250403.rds")
 
 names(wfs)
 
-dat.predictors <-  wfs[, c(81:111,125,126)]
+dat.predictors <-  wfs[, c(85:112,126,127)]
 
-dat.response <- wfs[, c(2:5, 10:13, 20:23)]
+dat.response <- wfs[, c(2:5, 10:13, 20:27)]
 
 dat <- cbind(dat.response, dat.predictors)
 
@@ -55,34 +56,58 @@ qb_margin.rf <- ranger(qb.margin.form, data = dat, importance = "permutation", n
 qb_margin.p <- data.frame(importance_pvalues(qb_margin.rf, method = "altmann",
                                              formula = qb.margin.form, data = dat, num.permutations = 1000))
 
-# run random forest using different importance
-pollinators_base.imp.rf <- ranger(pol.base.form, data = dat, importance = "impurity", num.trees = 1000)
-pollinators_base.imp.p <- data.frame(importance_pvalues(pollinators_base.imp.rf, method = "altmann", 
-                                                    formula = pol.base.form, data = dat, num.permutations = 1000))
-pollinators_margin.imp.rf <- ranger(pol.margin.form, data = dat, importance = "impurity", num.trees = 1000)
-pollinators_margin.imp.p <- data.frame(importance_pvalues(pollinators_margin.imp.rf, method = "altmann", 
-                                                      formula = pol.margin.form, data = dat, num.permutations = 1000))
+# Percent change models
+# formulae
+pol.pc.form <- reformulate(names(dat.predictors), response = "pc_pol_incr.margin")
+n_exp.pc.form <- reformulate(names(dat.predictors), response = "pc_nexp_red.margin")
+qb.pc.form <- reformulate(names(dat.predictors), response = "pc_qb_incr.margin")
+sed.pc.form <- reformulate(names(dat.predictors), response = "pc_sed_red.margin")
 
-n_export_base.imp.rf <- ranger(n_exp.base.form, data = dat, importance = "impurity", num.trees = 1000)
-n_export_base.imp.p <- data.frame(importance_pvalues(n_export_base.imp.rf, method = "altmann", 
-                                                 formula = n_exp.base.form, data = dat, num.permutations = 1000))
[…]

### 2025-04-04 RandomForest/make_rf_predictions.R

@@ -1,0 +1,226 @@
+library(ranger)
+library(iml)
+library(future)
+load("data/ES-RF3.RDA")
+pfun <- function(object, newdata) predict(object, data = newdata)$predictions
+
+future::plan(multisession, workers = 8)
+
+# pollinators base
+
+pred.pollinators_base <- Predictor$new(
+  model = pollinators_base.rf,
+  data = dat,
+  y = "pollinators_base",
+  predict.function = pfun
+)
+
+# int.polbase  <- Interaction$new(pred.pollinators_base)
+# int.polbase.ntr_pland  <- Interaction$new(pred.pollinators_base, feature = "ntr_pland")
+
+# pollinators margin
+
+pred.pollinators_margin <- Predictor$new(
+  model = pollinators_margin.rf,
+  data = dat,
+  y = "delta_pollinators_max.margin",
+  predict.function = pfun
+)
+
+# int.polmarg  <- Interaction$new(pred.pollinators_margin)
+# int.polmarg.ntr_pland  <- Interaction$new(pred.pollinators_margin, feature = "ntr_pland")
+
+# N export base
+
+pred.n_export_base <- Predictor$new(
+  model = n_export_base.rf,
+  data = dat,
+  y = "n_export_base",
+  predict.function = pfun
+)
+
+# int.nexpbase  <- Interaction$new(pred.n_export_base)
+# int.nexpbase.ntr_pland  <- Interaction$new(pred.n_export_base, feature = "ntr_pland")
+
+# N export margin
+
+pred.n_export_margin <- Predictor$new(
+  model = n_export_margin.rf,
+  data = dat,
+  y = "delta_n_export_max.margin",
+  predict.function = pfun
+)
+
+# int.nexpmarg  <- Interaction$new(pred.n_export_margin)
+# int.nexpmarg.ann_agr_pland  <- Interaction$new(pred.n_export_margin, feature = "ann_agr_pland")
+
+# Sediment export base
+
+pred.sed_export_base <- Predictor$new(
+  model = sed_export_base.rf,
+  data = dat,
+  y = "sed_export_base",
+  predict.function = pfun
+)
+
+# int.sedexpbase  <- Interaction$new(pred.sed_export_base)
+# int.sedexpbase.ann_agr_ed  <- Interaction$new(pred.sed_export_base, feature = "ann_agr_ed")
+
+# Sediment export margin
+
+pred.sed_export_margin <- Predictor$new(
[…]

### 2025-04-04 RandomForest/make_surrogate_rf.R

@@ -1,0 +1,97 @@
+# make random forest from unique datasets
+
+library(ranger)
+options(ranger.num.threads = 8)
+
+wfs <- readRDS("data/allresults_20250401.rds")
+
+names(wfs)
+
+selectedvar <- c("ntr_pland", "ann_agr_pland", "ntr_ed", 
+                 "field_frac_mn", "prn_pagr", "prn_agr_para_mn",
+                 "ann_agr_ed", "ann_agr_cohesion"
+)
+
+dat.predictors <-  wfs[, selectedvar]
+
+dat.response <- wfs[, c(2:5, 10:13, 20:23)]
+
+sur.dat <- cbind(dat.response, dat.predictors)
+
+# make rf formulas
+sur.pol.base.form <- reformulate(names(dat.predictors), response = "pollinators_base")
+sur.pol.margin.form <- reformulate(names(dat.predictors), response = "delta_pollinators_max.margin")
+
+sur.n_exp.base.form <- reformulate(names(dat.predictors), response = "n_export_base")
+sur.n_exp.margin.form <- reformulate(names(dat.predictors), response = "delta_n_export_max.margin")
+
+sur.qb.base.form <- reformulate(names(dat.predictors), response = "qb_base")
+sur.qb.margin.form <- reformulate(names(dat.predictors), response = "delta_qb_max.margin")
+
+sur.sed.base.form <- reformulate(names(dat.predictors), response = "sed_export_base")
+sur.sed.margin.form <- reformulate(names(dat.predictors), response = "delta_sed_max.margin")
+
+# run random forest
+sur.pollinators_base.rf <- ranger(sur.pol.base.form, data = sur.dat, importance = "permutation", num.trees = 1000)
+sur.pollinators_base.p <- data.frame(importance_pvalues(sur.pollinators_base.rf, method = "altmann", 
+                                                    formula = sur.pol.base.form, data = sur.dat, num.permutations = 1000))
+sur.pollinators_margin.rf <- ranger(sur.pol.margin.form, data = sur.dat, importance = "permutation", num.trees = 1000)
+sur.pollinators_margin.p <- data.frame(importance_pvalues(sur.pollinators_margin.rf, method = "altmann", 
+                                                      formula = sur.pol.margin.form, data = sur.dat, num.permutations = 1000))
+
+sur.n_export_base.rf <- ranger(sur.n_exp.base.form, data = vdat, importance = "permutation", num.trees = 1000)
[…]

### 2025-04-04 docs/RandomForest/make_rf_interactions.R

@@ -15,8 +15,8 @@ pred.pollinators_base <- Predictor$new(
   predict.function = pfun
 )
 
-int.polbase  <- Interaction$new(pred.pollinators_base)
-int.polbase.ntr_pland  <- Interaction$new(pred.pollinators_base, feature = "ntr_pland")
+# int.polbase  <- Interaction$new(pred.pollinators_base)
+# int.polbase.ntr_pland  <- Interaction$new(pred.pollinators_base, feature = "ntr_pland")
 
 # pollinators margin
 
@@ -27,8 +27,8 @@ pred.pollinators_margin <- Predictor$new(
   predict.function = pfun
 )
 
-int.polmarg  <- Interaction$new(pred.pollinators_margin)
-int.polmarg.ntr_pland  <- Interaction$new(pred.pollinators_margin, feature = "ntr_pland")
+# int.polmarg  <- Interaction$new(pred.pollinators_margin)
+# int.polmarg.ntr_pland  <- Interaction$new(pred.pollinators_margin, feature = "ntr_pland")
 
 # N export base
 
@@ -39,8 +39,8 @@ pred.n_export_base <- Predictor$new(
   predict.function = pfun
 )
 
-int.nexpbase  <- Interaction$new(pred.n_export_base)
-int.nexpbase.ntr_pland  <- Interaction$new(pred.n_export_base, feature = "ntr_pland")
+# int.nexpbase  <- Interaction$new(pred.n_export_base)
+# int.nexpbase.ntr_pland  <- Interaction$new(pred.n_export_base, feature = "ntr_pland")
 
 # N export margin
 
@@ -51,8 +51,8 @@ pred.n_export_margin <- Predictor$new(
   predict.function = pfun
 )
 
-int.nexpmarg  <- Interaction$new(pred.n_export_margin)
-int.nexpmarg.ann_agr_pland  <- Interaction$new(pred.n_export_margin, feature = "ann_agr_pland")
+# int.nexpmarg  <- Interaction$new(pred.n_export_margin)
+# int.nexpmarg.ann_agr_pland  <- Interaction$new(pred.n_export_margin, feature = "ann_agr_pland")
 
 # Sediment export base
 
@@ -63,8 +63,8 @@ pred.sed_export_base <- Predictor$new(
   predict.function = pfun
 )
 
-int.sedexpbase  <- Interaction$new(pred.sed_export_base)
[…]

### 2025-04-04 docs/RandomForest/make_shap_importance.R

@@ -3,45 +3,65 @@ library(fastshap)
 library(shapviz)
 
 load("data/ES-RF3.RDA")
+load("data/ES-RF3pc.RDA")
 
 pfun <- function(object, newdata) predict(object, data = newdata)$predictions
 
-dat.x <- dat[,13:45]
+dat.x <- dat[,17:46]
 
 # pollinators
 ex.polbase <- explain(pollinators_base.rf, X=dat.x, pred_wrapper = pfun,
-                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+                      nsim = 1000, adjust = TRUE, shap_only = FALSE)
 ex.polbase.glb <- shapviz(ex.polbase)
 
 ex.polmarg <- explain(pollinators_margin.rf, X=dat.x, pred_wrapper = pfun,
-                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+                      nsim = 1000, adjust = TRUE, shap_only = FALSE)
 ex.polmarg.glb <- shapviz(ex.polmarg)
 
 # N export
 ex.nexpbase <- explain(n_export_base.rf, X=dat.x, pred_wrapper = pfun,
-                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+                      nsim = 1000, adjust = TRUE, shap_only = FALSE)
 ex.nexpbase.glb <- shapviz(ex.nexpbase)
 
 ex.nexpmarg <- explain(n_export_margin.rf, X=dat.x, pred_wrapper = pfun,
-                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+                      nsim = 1000, adjust = TRUE, shap_only = FALSE)
 ex.nexpmarg.glb <- shapviz(ex.nexpmarg)
 
 # sediment export
 ex.sedbase <- explain(sed_export_base.rf, X=dat.x, pred_wrapper = pfun,
-                      nsim = 100, adjust = TRUE, shap_only = FALSE)
+                      nsim = 1000, adjust = TRUE, shap_only = FALSE)
 ex.sedbase.glb <- shapviz(ex.sedbase)
 
 ex.sedmarg <- explain(sed_export_margin.rf, X=dat.x, pred_wrapper = pfun,
-                      nsim = 100, adjust = TRUE, shap_only = FALSE)
[…]

### 2025-04-04 docs/data-raw/allresults_20250401.R

@@ -54,7 +54,15 @@ maxdelta <- base_out %>% select(hu:sed_export,rawsdr_avoid_exp, field_cells) %>%
     field.ha = (field_cells*900)/10000,
     margin.ha = (margin_cells*900)/10000,
     margin_den = margin.ha/lu_ta) %>%
-  mutate(across(delta_pollinators:delta_sed, \(x) x/margin_den, .names = "{.col}_max.margin"))  # normalize by margin density results in change in es per margin area (ws area cancels)
+  # normalize by margin density results in change in es per margin area (ws area cancels)
+  mutate(across(delta_pollinators:delta_sed, \(x) x/margin_den, .names = "{.col}_max.margin")) %>% 
+  # calculate percent ES change from baseline per margin area
+  mutate(
+    pc_pol_incr.margin = 100*(delta_pollinators/pollinators_base)/margin.ha,
+    pc_nexp_red.margin = 100*(delta_n_export/n_export_base)/margin.ha,
+    pc_qb_incr.margin = 100*(delta_qb/qb_base)/margin.ha,
+    pc_sed_red.margin = 100*(delta_sed/sed_export_base)/margin.ha,
+  )
 
 
 # one field vs baseline
@@ -96,7 +104,12 @@ allresults <- maxdelta %>%
   left_join(ofdelta.md) %>%
   left_join(hu_meta %>% rename(hu=huc12)) %>%
   left_join(lsm) %>%
-  left_join(lsm2 %>% select(hu, aglu_contag, ann_agr_cohesion, ntr_cohesion, prn_agr_cohesion)) %>%
+  left_join(lsm2 %>% 
+              select(hu, aglu_contag, ann_agr_cohesion, ntr_cohesion, prn_agr_cohesion,
+                     aglu_ed, ann_agr_area_mn, ntr_area_mn, prn_agr_area_mn)) %>%
+  # remove some metrics that are redundant/confusing/irrelevant
+  select(-ann_agr_para_mn, -ntr_para_mn, -prn_agr_para_mn, -field_para_mn,
+         -field_para_mn, -field_para_cv, -field_frac_cv, -field_area_cv) %>%
   left_join(hulandstats) %>%
   mutate(prn_ann_ratio = log(prn_agr_pland/ann_agr_pland),
          prn_pagr = 100*(prn_agr_pland/(ann_agr_pland+prn_agr_pland)),
@@ -105,6 +118,6 @@ allresults <- maxdelta %>%
[…]

### 2025-04-04 scripts/look_at_influences.R

@@ -1,0 +1,52 @@
+wfs <- readRDS("data/allresults_20250401.rds") 
+
+lsm <- read.csv("data/lsm.csv") %>%
+  mutate(hu = gsub("hu", "", hu))
+
+wfs.long <- wfs %>% 
+  pivot_longer(cols = c(pollinators_base:sed_export_base, delta_pollinators:delta_sed, delta_pollinators_max.margin:delta_sed_max.margin), names_to = "ES_response", values_to = "ES_value") %>%
+  select(hu, ES_response, ES_value, precip_mean) %>%
+  mutate(ES = gsub("delta_|_base|_max.margin","", ES_response)) %>%
+  mutate(ES = gsub("sed$","sed_export", ES)) %>%
+  mutate(response_type =
+           gsub("sed|export|n_|pollinators|qb|_|max","",ES_response))
+
+# fit some gams
+wfs.gam <- wfs.long %>% 
+  group_by(response_type, ES) %>%
+  nest() %>%
+  # Fit GAM
+  mutate(Mod = map(data, ~ lm(ES_value ~ precip_mean, data = .x))) %>%
+  # Get R2
+  mutate(R2 = map_dbl(Mod, ~round(summary(.x)$r.sq, 2)))  
+
+ggplot(wfs.long,
+       aes(precip_mean, ES_value)) + 
+  geom_point() + geom_smooth(method="lm") +
+  ggh4x::facet_grid2(response_type ~ ES, 
+                     scales = "free", independent = "y") + 
+  egg::theme_article() + 
+  # Add label
+  geom_label(data = wfs.gam, 
+             aes(x = Inf, y = Inf, 
+                 label = paste("R2 = ", R2, sep = " ")),
+             hjust = 1, vjust = 1)
+
+wfs.wider <- wfs.long %>% select(-ES_response) %>% 
+  pivot_wider(names_from = response_type, values_from = ES_value) %>%
+  left_join(lsm %>% select(hu, lu_ta))
+
+base.v.delta <- ggplot(wfs.wider, aes(base, delta)) + geom_point() + facet_wrap(~ES, scales = "free")
+delta.v.margin <- ggplot(wfs.wider, aes(delta, delta.margin)) + geom_point() + facet_wrap(~ES, scales = "free")
+base.v.margin <- ggplot(wfs.wider, aes(base, delta.margin)) + geom_point() + 
+  xlab("Baseline ES per ha") + ylab("Increase in ES per ha margin (ES units)") +
[…]
